\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}
\hypersetup{hidelinks}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}{Definition}

\setlength{\parskip}{4pt plus 1pt minus 1pt}
\setlength{\parindent}{0pt}

\title{Graph Structure Is Necessary for\\Information Preservation Under Bounded Context}
\author{Patrick D.\ McCarthy}
\date{}

\begin{document}
\maketitle

\begin{abstract}
Any system that (i) accumulates propositions about a world it cannot
fully observe, (ii) operates under a bounded active context --- a
hard limit on the number of propositions simultaneously available for
inference --- and (iii) must preserve information as the proposition
count grows beyond that bound, necessarily maintains a directed graph
with typed edges over its propositions. We prove this claim from two
independent arguments. The \emph{information-preservation argument}
shows that the only space-creating operation that does not lose
information is factoring --- extracting shared structure into named
nodes with typed edges --- and that factoring \emph{is} graph
construction. The \emph{retention-dynamics argument} shows that any
representation supporting sub-linear dependency queries, contextual
retrieval, and selective removal with cascade necessarily encodes
directed adjacency, which is a semantic graph regardless of substrate.
We show that continuous-weight mechanisms (e.g., transformer
self-attention) implement graph structure rather than constitute an
alternative to it. The result is substrate-independent: it applies to
biological neural networks, artificial neural networks, institutional
knowledge systems, or any other medium that stores propositions under
bounded context.
\end{abstract}

% ------------------------------------------------------------------
\section{Introduction}
\label{sec:intro}
% ------------------------------------------------------------------

Systems that accumulate knowledge face a fundamental tension. The
knowledge base grows, but the capacity to reason over it at any single
moment does not grow proportionally. Biological working memory holds
roughly $7 \pm 2$ items \cite{Miller1956}. Transformer context windows,
though large, degrade in inference quality as they fill with content
irrelevant to the current problem \cite{Liu2024}. Institutional
decision-makers cannot hold an entire organization's knowledge in a
single meeting. In each case, there is a hard or effective bound on
\emph{active context}: the number of propositions simultaneously
available for inference.

This paper asks a narrow question: \textbf{what structure must a
knowledge representation have if information is not to be lost as the
number of stored propositions exceeds the active context bound?}

The answer is a directed graph with typed edges. We prove this from two
independent arguments, each sufficient on its own:

\begin{enumerate}
\item \textbf{Information preservation} (Theorem~\ref{thm:graph-info}).
   When a bounded system must make space for new propositions, the only
   operation that creates space without losing information is
   \emph{factoring}: extracting shared structure into a named node with
   typed edges. Factoring is graph construction. Discarding without
   adjacency information is blind; lossy compression without adjacency is
   blind discarding with extra steps.

\item \textbf{Retention dynamics} (Theorem~\ref{thm:graph-dynamics}).
   Any representation that supports sub-linear dependency queries ---
   answering ``what depends on this proposition?'' in less than $O(n)$
   time --- necessarily stores directed adjacency. Selective removal
   with cascade (invalidating a proposition and propagating to its
   dependents) requires typed edges to scope correctly. These are the
   operations any system performs when maintaining knowledge under
   continuous evidence arrival.
\end{enumerate}

The result is substrate-independent. It does not prescribe an
implementation (adjacency lists, edge tables, attention masks, synaptic
connections) --- it proves that the \emph{information content} of any
adequate representation is a directed graph, regardless of how that
graph is physically realized.

\subsection{Related work}

Graph-structured knowledge representations are widely used in practice:
knowledge graphs \cite{Hogan2021}, Bayesian networks \cite{Pearl1988},
description logics \cite{BaaderCalvaneseEtAl2003}, and the
RDF/OWL stack \cite{BernersLee2001} all represent knowledge as directed
labeled graphs. Guarino's formal ontology \cite{Guarino1998} grounds
this choice in the structure of the domain: entities, relations, and
constraints map naturally to nodes, edges, and types. These traditions
have demonstrated that graph structure is \emph{effective} for knowledge
representation. The question of whether it is \emph{necessary} --- as
opposed to convenient --- has received less formal attention.

The present paper provides the missing argument: under bounded active
context, graph structure is not a design preference but a
mathematical consequence of information preservation
(Theorem~\ref{thm:graph-info}) and sub-linear maintenance
(Theorem~\ref{thm:graph-dynamics}). The typed directed edges of
description logics and OWL are not an engineering convenience but the
minimal structure any system must maintain. This result retroactively
grounds decades of practice in a formal necessity argument.

Dupoux, LeCun, and Malik \cite{DupouxLeCunMalik2026} propose a
dual-system architecture (observation + action) with meta-control
routing, motivated by cognitive science. Their framework is
architectural and prescriptive: it argues that autonomous learning
\emph{should} integrate observation, action, and meta-control. The
present work is complementary: rather than proposing what intelligent
systems should contain, we derive what they \emph{must} contain under
bounded context.

The information bottleneck principle \cite{Tishby2000} establishes that
optimal representations compress input while preserving information
relevant to output. Our information-preservation argument
(Theorem~\ref{thm:graph-info}) addresses a different question: not what
to compress, but what structure the representation must have to compress
without information loss.

Simon's bounded rationality \cite{Simon1955} establishes that cognitive
agents satisfice rather than optimize. Our bounded context constraint
($C_n$) formalizes a specific consequence: when the bound is hard,
structure is not optional but necessary.

% ------------------------------------------------------------------
\section{Formal Framework}
\label{sec:framework}
% ------------------------------------------------------------------

We introduce the minimal definitions required for the two main results.

\begin{definition}[World State Space]
\label{def:world}
Let $W$ be a measurable space of possible world states. At time $t$, the
true state $w_t \in W$ is not directly observable by any entity.
\end{definition}

\begin{definition}[Proposition and Knowledge Base]
\label{def:knowledge}
A \emph{proposition} $p$ is a claim about $W$ with an associated
likelihood function $\ell_p : W \to [0,1]$. A \emph{knowledge base} $K$
is a collection of propositions. The posterior over world states given
$K$ is:
\[
P(w \mid K) \propto P_0(w) \cdot \prod_{p \in K} \ell_p(w)
\]
where $P_0$ is a prior. The \emph{uncertainty} of $K$ about $w$ is the
surprisal $U(w, K) = -\log P(w \mid K)$ \cite{Shannon1948, Cover2006}.
\end{definition}

\begin{definition}[Bounded Active Context]
\label{def:context}
For an entity $n$, let $C_n$ denote its \emph{active context capacity}:
the maximum number of propositions simultaneously available for
inference. Let
$\mathrm{ctx}(t) \subseteq K$ be the set of propositions loaded into
active context at time $t$, with $|\mathrm{ctx}(t)| \leq C_n$.

The \emph{inference quality} for problem $f$ is the relevance density:
\[
\rho(f) = \frac{|K^{[f]}|}{|\mathrm{ctx}(t)|}
\]
where $K^{[f]} \subseteq K$ is the set of propositions relevant to $f$.
$\rho = 1$ when active context holds exactly the relevant propositions;
$\rho \to 0$ when context is dominated by irrelevant content.
\end{definition}

\begin{definition}[Growth and Curation Regimes]
\label{def:regimes}
$K$ has a maximum capacity $K_{max}$ determined by substrate constraints.
When $|K| < K_{max}$, $K$ is in the \emph{growth regime}: propositions
accumulate freely. When $|K| = K_{max}$, $K$ is in the \emph{curation
regime}: adding a new proposition requires making space.
\end{definition}

\begin{definition}[Contextual Retrieval]
\label{def:retrieval}
Given problem $f$, \emph{contextual retrieval} is the operation that
loads $K^{[f]}$ into active context. The retrieval cost is the number of
propositions examined to identify $K^{[f]}$.
\end{definition}

% ------------------------------------------------------------------
\section{Graph Necessity from Information Preservation}
\label{sec:info}
% ------------------------------------------------------------------

\begin{definition}[Factoring]
\label{def:factoring}
Let $p_1, \ldots, p_m \in K$ share mutual information: there exists a
random variable $F^*$ (a function of $W$) such that
$I(F^*; \ell_{p_i}) > 0$ for all $i$, and the joint likelihood
decomposes as
\[
\ell_{p_i}(w) = g_i\!\bigl(\ell_{p_i}^{\text{res}}(w),\; F^*(w)\bigr)
\]
for some residual $\ell_{p_i}^{\text{res}}$ and reconstruction function
$g_i$. The decomposition need not be unique; any valid decomposition
suffices. In the general case, $g_i$ may be an arbitrary measurable
function --- the definition does not require multiplicative,
additive, or any other restricted functional form.
\emph{Factoring} is the operation that:
\begin{enumerate}
\item[(i)] Extracts $F^*$ into a single named, reusable unit.
\item[(ii)] Replaces the shared component in each $p_i$ with a
   reference to $F^*$.
\item[(iii)] Stores, for each reference, the reconstruction function
   $g_i$ specifying the role $F^*$ plays in recovering $p_i$.
\end{enumerate}
The total information is preserved: $F^*$ plus the residuals and their
reconstruction functions jointly recover all original $\ell_{p_i}$.

Note: this definition is stated without graph vocabulary. The output of
factoring is a set of units with references between them and role
annotations on each reference. We prove in
Theorem~\ref{thm:graph-info} that this structure necessarily constitutes
a directed graph with typed edges.

\textbf{Remark.} Factoring is not a novel operation introduced by this
paper. It is the central operation of description logics
\cite{BaaderCalvaneseEtAl2003}: the TBox/ABox separation extracts shared
conceptual structure (the TBox) from individual assertions (the ABox),
and concept subsumption hierarchies are literally shared components
referenced by multiple individuals. Upper ontologies such as BFO
\cite{Arp2015} and DOLCE \cite{Masolo2003} carry this further:
approximately 35 categories in BFO organize all of biomedical knowledge,
demonstrating that real-world knowledge bases have abundant shared
structure amenable to factoring. The definition above formalizes what
these traditions have practiced for decades.
\end{definition}

\begin{theorem}[Graph Necessity from Information Preservation]
\label{thm:graph-info}
Let $K$ be a knowledge base with bounded active context $C_n$
(Definition~\ref{def:context}) in the curation regime
(Definition~\ref{def:regimes}). Any representation of $K$ that
preserves information as $|K|$ grows beyond $C_n$ necessarily
maintains a directed graph with typed edges over propositions.
\end{theorem}

\begin{proof}
In the curation regime, adding any new proposition $p_{new}$ requires
making space. We enumerate the ways to make space and show that every
information-preserving option produces graph structure.

\textbf{Option A: Discard an existing proposition.}
Without explicit relational structure, the entity has no mechanism to
determine which existing propositions are redundant. Two propositions
$p_1, p_2$ that share a latent factor appear as independent entries;
there is no stored evidence that they share structure. The choice of
which to discard is therefore made without information about relational
dependencies. Discarding $p_{old}$ loses the information it encodes, and
potentially loses the information that $p_{old}$ has any relation to
other propositions --- a second-order loss. This is information loss.

\textbf{Option B: Factor --- extract shared structure
(Definition~\ref{def:factoring}).}
If propositions share a common informational component $F^*$, factoring
extracts $F^*$ as a single named node with typed edges, replacing
redundant copies with shared references. Space is created without loss:
the original information is reconstructable from $F^*$ plus the
residuals. Factoring also resolves the update problem: when new evidence
revises $F^*$, the single node is updated once, and the revision
propagates to all propositions that reference it via edges ---
$O(1)$ instead of $O(|K|)$.

By Definition~\ref{def:factoring}, factoring produces nodes and directed
typed edges. This is graph construction.

\textbf{Option C: Reorganization --- rewrite propositions into a more
compact equivalent encoding.}
Propositions $p_1, \ldots, p_m$ could be replaced by $q_1, \ldots, q_k$
($k < m$) that are jointly equivalent: the same posterior over $W$ is
induced. This is a change of basis in the information-theoretic sense.

If the reorganization preserves all dependency relationships ---
which propositions ground which --- then the result is an isomorphic
knowledge base with fewer nodes. But ``preserving dependency
relationships'' requires storing those relationships, which is storing
adjacency. If the reorganization does \emph{not} preserve dependency
relationships, then downstream propositions that depended on the
original $p_i$ are silently broken: they reference nodes that no longer
exist, with no forwarding information. The entity cannot determine what
broke without adjacency.

There is a special case: reorganization that extracts shared structure
across the $p_i$ into shared nodes referenced by the $q_j$. This is
factoring (Option~B). Reorganization that does not extract shared
structure but merely rewrites individual propositions more compactly
does not create space from shared structure --- it creates space from
per-proposition compression, which is lossless compression (Option~D).

\textbf{Option D: Lossless compression --- encode propositions more
compactly without information loss.}
Lossless compression reduces the physical footprint of a knowledge store
but does not change the hard capacity bound $C_n$. The bound is a
constraint on how many propositions can be \emph{simultaneously active
for inference}, not on how compactly they are encoded on disk.
A compressed bitstring must be decompressed before any proposition can be
accessed or any dependency traced; the decompressed form faces the same
curation pressure as the original. Compression is orthogonal to the
structural question the theorem addresses.

\textbf{Option E: Lossy compression --- replace propositions with a
summary that occupies less space.}
Compressing $p_1$ and $p_2$ into a summary $p_s$ that discards detail is
a form of discarding: the lost detail may ground other propositions or
actions. If $p_1$ has dependents --- propositions that reference it ---
then replacing $p_1$ with $p_s$ silently invalidates those dependents
unless the entity knows what depends on the lost detail. But knowing
what depends on $p_1$ \emph{is} knowing $p_1$'s adjacency --- its
edges. Without stored adjacency, lossy compression is blind discarding:
the entity cannot evaluate what breaks. With stored adjacency, the
entity can scope the damage, notify dependents, and compress safely ---
but storing adjacency is maintaining a graph. Lossy compression that
preserves correctness requires the same relational structure as
factoring. Lossy compression that ignores relational structure is
Option~A with extra steps.

\textbf{Exhaustiveness.} Options A--E cover the space of
space-creating operations: remove information (A, E), extract shared
structure (B), rewrite to equivalent form (C), or encode more compactly
(D). Of these, A and E lose information. D is orthogonal (it addresses
storage density, not the active-context bound). C either reduces to B (when shared structure is extracted)
or requires stored adjacency to preserve dependencies. Only B
(factoring) creates space, preserves information, and solves the
structural problem.

\textbf{The output of factoring is a directed graph with typed edges.}
Definition~\ref{def:factoring} is stated without graph vocabulary: it
produces named units with references and reconstruction roles. We now
show this structure has graph properties:

\emph{Nodes.} The named units --- shared components $F^*$ and residual
propositions $p_i^{\text{res}}$ --- are the nodes. Each is an
identifiable, individually addressable unit of information.

\emph{Directed edges.} Each reference from $p_i^{\text{res}}$ to $F^*$
is directed: $p_i$ depends on $F^*$ for reconstruction, but $F^*$ does
not depend on $p_i$. Revising $F^*$ requires updating the
reconstruction of every $p_i$ that references it; revising $p_i$ does
not require updating $F^*$. This asymmetry is inherent in the
extraction operation: the shared component is referenced by the
residuals, not vice versa. Any structure encoding asymmetric
dependencies between named units is a directed graph.

\emph{Typed edges.} The reconstruction role $g_i$ stored on each
reference distinguishes the \emph{kind} of dependency: whether $F^*$
\emph{grounds} $p_i$ (the residual's truth value depends on $F^*$),
\emph{evidences} $p_i$ (the residual's likelihood is informed by
$F^*$), or \emph{causes} $p_i$ (the residual's state is produced by
$F^*$). Without role annotation, cascade from an invalidated $F^*$
cannot distinguish dependents that require re-evaluation (inferentially
grounded through $F^*$) from units merely associated with $F^*$.
Untyped references force worst-case cascade over all references; typed
references scope cascade to dependency-bearing references only. Since
the reconstruction role $g_i$ is stored by
Definition~\ref{def:factoring}(iii), the edges are typed.

A collection of named units with directed, typed references between them
is a directed graph with typed edges, regardless of whether it is
implemented as adjacency lists, pointers, sparse matrices, or any other
medium.

\textbf{Remark on shared structure.} The theorem requires that
propositions share informational components. This is not an additional
assumption but a consequence of the knowledge base describing a
structured world. The proposition ``fire is hot'' is relevant to
cooking, camping, metallurgy, and fire safety --- any knowledge base
spanning these domains contains this shared component. A knowledge base
with \emph{zero} shared structure would be a collection of completely
independent facts about completely unrelated domains: each proposition
grounded in its own private ontology with no overlap. Such a knowledge
base faces no curation pressure (nothing is redundant, so nothing can
be factored) and requires no structural optimization. The theorem
addresses the non-degenerate case: knowledge bases about structured
worlds, which is every knowledge base that admits inference across
domains. The success of upper ontologies confirms that this is the
typical case: BFO \cite{Arp2015} organizes all of biomedical science
under approximately 35 shared categories; DOLCE \cite{Masolo2003}
provides cross-domain foundational categories reused across engineering,
geography, and law. These are not design choices but reflections of
shared structure that already exists in the domains they describe.

\textbf{Remark on compression.} Lossless compression changes how
compactly propositions are stored, not how many can be simultaneously
active. The bounded-context premise ($|K| > C_n$) is a constraint on
active inference capacity, not storage density. No encoding scheme
removes this bound; compression is orthogonal to the structural
question.
\end{proof}

% ------------------------------------------------------------------
\section{Graph Necessity from Retention Dynamics}
\label{sec:dynamics}
% ------------------------------------------------------------------

\begin{theorem}[Graph Necessity from Retention Dynamics]
\label{thm:graph-dynamics}
Let $K$ be a knowledge base supporting continuous knowledge maintenance
under evidence arrival. The maintenance operations are:
\begin{enumerate}
\item[(i)] \textbf{Addition with provenance}: insert a proposition $p$
   together with its evidential grounds --- which existing propositions
   $p$ depends on, and by what relation.
\item[(ii)] \textbf{Contextual retrieval}: given problem $f$, retrieve
   exactly $K^{[f]}$.
\item[(iii)] \textbf{Dependency query}: given proposition $p$, determine
   which propositions are grounded through $p$.
\item[(iv)] \textbf{Selective removal with cascade}: when $p$ is
   invalidated, remove $p$ and re-evaluate every proposition whose
   grounding depends on $p$, transitively, until $K$ stabilizes.
\end{enumerate}
Any representation achieving sub-linear cost on all four operations
simultaneously necessarily encodes directed adjacency between
propositions --- and any structure encoding directed adjacency over
propositions with typed relations \emph{is} a semantic graph, regardless
of implementation.
\end{theorem}

\begin{proof}
\textbf{Part 1: Dependency queries are necessary and continuous.}
Knowledge maintenance operates continuously: every new piece of evidence
potentially shifts the status of existing propositions. When evidence
revises or invalidates a proposition $p$, the system must evaluate every
proposition whose grounding passes through $p$. This is not a batch
operation that can be deferred --- acting on a proposition whose ground
has been invalidated risks incorrect inference. The dependency query
``what is grounded through $p$?'' is therefore invoked at the frequency
of evidence arrival.

\textbf{Part 2: Lower bound without adjacency.}
Consider a representation $R$ that stores $n$ propositions without
explicit directed relationships between them (a flat store, hash table,
embedding space, or any structure that does not record which propositions
depend on which). To answer the dependency query for $p$, $R$ must
examine each of the other $n - 1$ propositions to determine whether its
grounding involves $p$. Without stored adjacency, there is no mechanism
to restrict this examination: any proposition might depend on $p$, and
the only way to determine this is to check.

The cost of a single dependency query is therefore $\Omega(n)$ for any
representation without stored adjacency. Since dependency queries are
invoked at evidence-arrival frequency (Part~1), the amortized cost of
maintenance is $\Omega(n)$ per evidence event --- growing linearly with
$|K|$.

\textbf{Part 3: Cascade amplifies the cost.}
Selective removal is not a single deletion. When $p$ is removed, each
proposition $q$ grounded through $p$ must be re-evaluated: does $q$ have
alternative grounds, or should it also be removed? If $q$ also fails,
its dependents must be re-evaluated in turn. This is a transitive
closure on the dependency relation.

With explicit adjacency, cascade follows edges: cost is $O(|D_p|)$ where
$|D_p|$ is the transitive dependency set. Without adjacency, each level
of cascade requires an $\Omega(n)$ scan. Total cost: $\Omega(n \cdot L)$
where $L$ is the cascade depth, versus $O(|D_p|)$ in the graph where
$|D_p| \ll n$ for typical propositions with bounded connectivity.

\textbf{Remark.} Truth maintenance systems \cite{Doyle1979,DeKleer1986}
are a direct implementation of operations~(iii) and~(iv). A TMS stores
directed justifications (which propositions ground which) and propagates
belief revision through them when any justification is invalidated ---
precisely the dependency query and cascade operations this theorem
proves are necessary. The fact that AI research independently invented
TMS to solve this exact problem is empirical evidence for the theorem's
claim: without stored directed adjacency, maintenance under evidence
arrival is intractable.

\textbf{Part 4: Any sub-linear solution encodes adjacency.}
Suppose $R$ achieves $o(n)$ cost for dependency queries. Then $R$ must
store, for each proposition, information about which other propositions
depend on it --- otherwise the $\Omega(n)$ lower bound of Part~2
applies. This stored information is a set of directed relationships: for
each proposition $p$, a record of the propositions that reference $p$ as
part of their grounding.

This is precisely a directed adjacency structure. The nodes are
propositions. The edges are directed dependency relations. Any
representation that stores this information --- regardless of whether it
is implemented as adjacency lists, edge tables, pointers, or any other
mechanism --- \emph{is} a directed graph over propositions.

\textbf{Remark on approximate methods.}
Approximate retrieval mechanisms --- learned embeddings,
locality-sensitive hashing, approximate nearest-neighbor indices ---
achieve sub-linear performance on operation~(ii) (contextual retrieval)
without storing explicit directed adjacency. This does not circumvent
the theorem, because the theorem requires sub-linear cost on all four
operations \emph{simultaneously}. Approximate similarity over an
embedding space can identify propositions \emph{related to} a query, but
cannot answer ``what is \emph{grounded through} $p$?'' (operation~iii)
or propagate invalidation transitively through dependency chains
(operation~iv). These operations require knowing which propositions
depend on which --- not which are \emph{nearby} in a metric space.
Proximity and dependency are distinct relations: $p$ and $q$ may be
embedding-close yet inferentially independent, or embedding-distant yet
linked by a grounding chain. Any representation that achieves sub-linear
cost on dependency queries and cascade necessarily stores directed
dependency relationships, which is directed adjacency.

\textbf{Part 5: Typed edges are necessary for correct cascade.}
A bare adjacency structure (untyped edges) is insufficient. When $p$ is
invalidated, cascade must distinguish between propositions
\emph{inferentially grounded} through $p$ (which require re-evaluation)
and propositions merely \emph{associated} with $p$ (which do not). An
untyped edge conflates these: every neighbor must be treated as
potentially dependent, degrading cascade to the full neighborhood rather
than the true dependency set.

Typed edges --- carrying the relation type (``grounds,'' ``evidences,''
``causes,'' ``supports'') --- enable cascade to follow only
dependency-bearing edges, keeping cost proportional to $|D_p|$ rather
than $|\mathrm{adj}(p)|$. Since $|\mathrm{adj}(p)|$ can be much larger
than $|D_p|$, typed edges are necessary for efficient cascade.

\textbf{Remark on empirical convergence.}
The necessity of typed edges is confirmed by the empirical history of
knowledge representation. Early semantic networks used untyped
``is-related-to'' links; Woods \cite{Woods1975} demonstrated that these
were semantically ambiguous and operationally inadequate --- the system
could not distinguish inheritance from association, causation from
correlation. The field converged to typed relations: RDF introduced
typed predicates \cite{BernersLee2001}, OWL distinguished object
properties, data properties, and annotation properties with transitive,
symmetric, and inverse role characteristics
\cite{BaaderCalvaneseEtAl2003}. This trajectory --- starting untyped,
failing, converging to typed --- is exactly the progression Part~5
predicts must occur.

A directed graph with typed edges over propositions is a semantic graph.
\end{proof}

% ------------------------------------------------------------------
\section{Continuous-Weight Mechanisms Are Graph Implementations}
\label{sec:attention}
% ------------------------------------------------------------------

A natural objection: modern attention mechanisms achieve contextual
retrieval without explicit graph construction. Does this constitute a
non-graph alternative?

\begin{theorem}[Contextual Retrieval Requires Adjacency Computation]
\label{thm:attention}
Any mechanism that performs contextual retrieval over $n$ propositions
--- selecting which propositions are relevant to a given query --- must
compute pairwise relevance between the query and stored propositions.
This computation either (a) evaluates all $n$ propositions per query at
$O(n)$ or $O(n^2)$ cost (dynamic adjacency), or (b) pre-computes and
stores relevance relationships at $O(d)$ per query where $d$ is the
local degree (stored adjacency). In case (b), the stored relationships
constitute a directed graph. Case (a) is the cost of recomputing that
graph from scratch at every query.
\end{theorem}

\begin{proof}
Contextual retrieval for problem $f$ requires identifying $K^{[f]}$
--- the propositions relevant to $f$. This requires determining, for
each proposition $p \in K$, whether $p$ is relevant to $f$. Without
stored information about relevance relationships, every proposition must
be evaluated: cost $\Omega(n)$ per query.

Full self-attention is the canonical example:
$A_{ij} = \mathrm{softmax}(Q_i K_j^\top / \sqrt{d_k})$ evaluates all
$n^2$ pairwise relevances per query. This is not ``computing a graph''
in the trivial sense that any matrix can be called a graph --- it is
performing the specific computation (pairwise relevance determination)
that a pre-built graph would make unnecessary. The attention matrix is
dense: all $n^2$ entries are nonzero, encoding no structural
information. The computational cost is the cost of having no stored
structure.

Pre-computed adjacency --- stored information about which propositions
are relevant to which queries --- reduces retrieval to $O(d)$ where $d$
is the number of stored relevant relationships. This is a directed
graph: nodes are propositions, edges are relevance relationships, and
the direction encodes which proposition informs which query context.

Sparse attention mechanisms --- learned sparsity patterns, local
windows, routing functions --- reduce the $O(n^2)$ cost by learning
which pairwise relevances are worth computing: they are learning which
edges to store. The KV cache materializes computed relevances for reuse
across queries. Each optimization converts dynamic adjacency computation
into stored adjacency --- progressively building the graph that
Theorems~\ref{thm:graph-info} and~\ref{thm:graph-dynamics} prove is
necessary.

The result is not that ``attention is a graph'' (which would be
trivially true of any matrix) but that contextual retrieval requires
adjacency information, and the only question is whether that information
is computed dynamically ($O(n^2)$) or stored explicitly ($O(d)$). Stored
adjacency over propositions with typed relevance relationships is a
directed graph.
\end{proof}

\textbf{Remark (Scaling context windows).}
Growing $C_n$ directly does not solve the retrieval problem. For any
specific problem $f$, the relevance density $\rho(f) = |K^{[f]}| /
|\mathrm{ctx}(t)|$ (Definition~\ref{def:context}) decreases as context
fills with content not relevant to $f$ --- an effect documented
empirically as ``lost in the middle'' \cite{Liu2024}. The efficient
path is not to grow the context window but to grow the encoded knowledge
accessible via stored adjacency: filling the graph, not the context
window.

% ------------------------------------------------------------------
\section{Discussion}
\label{sec:discussion}
% ------------------------------------------------------------------

The two theorems establish graph necessity from independent premises.
Theorem~\ref{thm:graph-info} requires only the curation regime (bounded
capacity, new propositions arriving). Theorem~\ref{thm:graph-dynamics}
requires only continuous knowledge maintenance (evidence arrives,
dependencies must be traced) and does not assume lossless preservation
--- it applies equally to systems that tolerate information loss,
because the cost argument concerns maintenance operations, not
retention policy. Any system subject to both constraints ---
which includes every physical system that accumulates knowledge --- is
doubly committed to graph structure.

\subsection{The continuous-representation objection}

The strongest objection to this paper's results is that they
assume discrete propositions, while the most capable knowledge systems
--- large language models, biological neural networks --- store
knowledge in continuous distributed representations that do not
decompose into identifiable propositions. If there are no discrete
propositions, there is nothing to factor
(Theorem~\ref{thm:graph-info}), nothing to query dependencies over
(Theorem~\ref{thm:graph-dynamics}), and calling an attention matrix a
``graph'' (Theorem~\ref{thm:attention}) is a mathematical relabeling,
not a structural discovery. The objection: the theorems prove graph
necessity for an idealized system that no successful architecture
actually implements.

This objection is serious and requires a substantive response. We offer
three.

\textbf{(1) Continuous representations do encode discrete
propositions.}
The continuity of the substrate does not entail that the information
content is non-propositional. Meng et al.\ \cite{Meng2022} demonstrate
that specific factual associations in GPT-class models can be localized,
edited, and ablated at identified parameter subsets: ``The Eiffel Tower
is in Paris'' has a locatable representation that can be surgically
revised to ``The Eiffel Tower is in Rome'' without retraining. Geva
et al.\ \cite{Geva2021} show that feed-forward layers in transformers
function as key-value memories, with individual neurons activating for
identifiable semantic patterns. Li et al.\ \cite{Li2024} trace factual
recall to discrete circuits within continuous networks.

These results establish that continuous weight matrices encode
propositional content --- not approximately or metaphorically, but in a
sense precise enough to support surgical editing. The propositions are
\emph{superposed} (multiple propositions share parameters) and
\emph{distributed} (each proposition spans multiple parameters), but
they are identifiable, editable, and individually removable. Superposed
encoding is precisely the condition that makes factoring valuable: shared
structure across propositions is the input to factoring, and continuous
networks contain it abundantly.

\textbf{(2) The claim is about information content, not
implementation.}
Theorems~\ref{thm:graph-info} and~\ref{thm:graph-dynamics} prove that
the \emph{information content} of any adequate representation has graph
structure. They do not prescribe adjacency lists, edge tables, or
symbolic labels. A neural network whose weights encode directed
dependencies between propositions \emph{is} maintaining a graph in the
sense of Definition~\ref{def:knowledge} --- the graph exists in the
information, not in the data structure.

The objection that ``if any structured representation counts as a
graph, the claim is trivially true'' misidentifies the contribution.
The claim is not that representations have structure (which is trivial)
but that the structure must be \emph{directed}, \emph{typed}, and
\emph{adjacency-preserving} --- and that representations lacking these
specific properties lose information or pay $\Omega(n)$ maintenance
costs. Not all structure is graph structure. A flat embedding space has
structure (metric, topology) but not directed adjacency; a hash table
has structure (key-value mapping) but not typed dependencies. The
theorems identify \emph{which} structural properties are necessary, not
merely that some structure is required.

\textbf{(3) Catastrophic forgetting is the theorem's prediction for
systems without explicit adjacency.}
This is the strongest empirical evidence for the result.
Theorem~\ref{thm:graph-dynamics} proves that without stored directed
adjacency, selective removal with cascade --- updating one proposition
and propagating correctly to its dependents --- costs $\Omega(n)$ per
evidence event. The prediction: systems that store propositions in
continuous weights without explicit adjacency should exhibit
\emph{uncontrolled interference when knowledge is updated}.

This is precisely catastrophic forgetting
\cite{McCloskey1989,Kirkpatrick2017}: updating a neural network on new
information degrades previously learned information unpredictably.
The network cannot scope the update because it lacks the adjacency
to determine what depends on the modified parameters. Continual
learning remains an open problem in deep learning not because of
insufficient engineering but because the underlying representation
lacks the structural property the theorem identifies: directed
adjacency that enables scoped updates.

The response to the continuous-representation objection is therefore
not defensive but \emph{predictive}: the theorem identifies the specific
structural deficit (missing explicit adjacency), predicts the specific
failure mode (unscoped updates causing cascading interference), and the
most successful continuous-representation systems exhibit exactly that
failure mode. The objection that the theorems do not apply to neural
networks is refuted by the fact that neural networks suffer from
precisely the pathology the theorems predict for representations
without graph structure.

\textbf{(4) Knowledge graph embeddings are continuous implementations
of graph structure.}
A further data point: knowledge graph embedding methods (TransE
\cite{Bordes2013}, RotatE \cite{Sun2019}) project graph structure into
continuous vector spaces. The embeddings are continuous, but the
information content they encode --- which entities are related to which,
by what relation --- is directed typed adjacency. Inference tasks such
as link prediction require reconstructing adjacency from the embedding
space. These methods do not eliminate graph structure; they implement it
in a continuous substrate, which is exactly the relationship
Theorems~\ref{thm:graph-info}--\ref{thm:graph-dynamics} predict between
information content (graph) and implementation (free parameter).

\subsection{Scope and non-claims}

The theorems prove that the \emph{information content} of the
representation must be a directed graph. They do not prescribe a
specific graph formalism (property graph, RDF, Bayesian network) or
implementation (adjacency lists, sparse matrices, attention masks). The
graph is the abstract structure; the substrate is a free parameter.

\textbf{Implications for AI architecture.}
Theorem~\ref{thm:attention} reframes the relationship between
graph-based and neural approaches: they are not competing paradigms but
different implementations of the same structural necessity. Systems that
compute adjacency dynamically (full attention) pay $O(n^2)$; systems
that store it explicitly (graphs, sparse attention, KV caches) pay
$O(d)$. The trend in architecture design --- from dense attention toward
sparse, structured, and cached retrieval --- is convergence toward
explicit graph structure, driven by the same cost pressures the theorems
identify.

The result suggests that hybrid architectures combining explicit
knowledge graphs with neural retrieval are not an engineering
convenience but a response to a structural constraint: as $|K|$ grows,
systems must encode adjacency to maintain inference quality under bounded
context.

\textbf{Implications for scaling.}
The scaling remark following Theorem~\ref{thm:attention} challenges the
prevailing approach of growing context windows as the primary path to
more capable inference.
Under bounded context, larger windows with undifferentiated content
\emph{reduce} inference quality for any specific problem. The
alternative --- encoding knowledge into a graph and retrieving the
relevant subgraph per problem --- maintains $\rho = 1$ regardless of
total knowledge size. This is the difference between scaling the window
and scaling the knowledge.

\textbf{Biological evidence.}
Biological neural networks maintain sparse, structured connectivity
rather than dense all-to-all connections. Working memory is bounded
($C_n \approx 7 \pm 2$) while long-term knowledge grows without
apparent limit \cite{Miller1956}. Retrieval is associative and
context-driven --- consistent with graph traversal from a query node
rather than exhaustive scan. The theorems suggest this architecture is
not an accident of evolutionary history but a necessary consequence of
bounded context under information-preservation pressure.

% ------------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}
% ------------------------------------------------------------------

We have shown that graph structure over propositions is not a design
choice but a necessary consequence of two independently sufficient
constraints: information preservation under bounded capacity, and
sub-linear knowledge maintenance under continuous evidence arrival.
Continuous-weight mechanisms such as transformer self-attention implement
graph structure dynamically rather than constituting alternatives to it.
The result is substrate-independent: any system that accumulates
propositions under bounded active context and must preserve information
necessarily maintains a directed graph with typed edges, whether that
graph is realized as explicit edges, learned sparse attention, synaptic
connectivity, or any other medium.

\bibliographystyle{plain}
\begin{thebibliography}{99}

\bibitem{Arp2015}
R.~Arp, B.~Smith, and A.~D. Spear.
\newblock \emph{Building Ontologies with Basic Formal Ontology}.
\newblock MIT Press, 2015.

\bibitem{BaaderCalvaneseEtAl2003}
F.~Baader, D.~Calvanese, D.~L. McGuinness, D.~Nardi, and P.~F.
  Patel-Schneider, editors.
\newblock \emph{The Description Logic Handbook: Theory, Implementation, and
  Applications}.
\newblock Cambridge University Press, 2003.

\bibitem{BernersLee2001}
T.~Berners-Lee, J.~Hendler, and O.~Lassila.
\newblock The semantic web.
\newblock \emph{Scientific American}, 284(5):34--43, 2001.

\bibitem{Bordes2013}
A.~Bordes, N.~Usunier, A.~Garcia-Dur{\'a}n, J.~Weston, and
  O.~Yakhnenko.
\newblock Translating embeddings for modeling multi-relational data.
\newblock In \emph{Advances in Neural Information Processing Systems},
  volume~26, pages 2787--2795, 2013.

\bibitem{Cover2006}
T.~M. Cover and J.~A. Thomas.
\newblock \emph{Elements of Information Theory}.
\newblock Wiley-Interscience, 2nd edition, 2006.

\bibitem{DeKleer1986}
J.~de~Kleer.
\newblock An assumption-based {TMS}.
\newblock \emph{Artificial Intelligence}, 28(2):127--162, 1986.

\bibitem{DeLange2021}
M.~De~Lange, R.~Aljundi, M.~Masana, S.~Parisot, X.~Jia, A.~Leonardis,
  G.~Slabaugh, and T.~Tuytelaars.
\newblock A continual learning survey: Defying forgetting in classification
  tasks.
\newblock \emph{IEEE Transactions on Pattern Analysis and Machine Intelligence},
  44(7):3366--3385, 2021.

\bibitem{Doyle1979}
J.~Doyle.
\newblock A truth maintenance system.
\newblock \emph{Artificial Intelligence}, 12(3):231--272, 1979.

\bibitem{DupouxLeCunMalik2026}
E.~Dupoux, Y.~LeCun, and J.~Malik.
\newblock Why {AI} systems don't learn and what to do about it: Lessons on
  autonomous learning from cognitive science.
\newblock \emph{arXiv preprint arXiv:2603.15381}, 2026.

\bibitem{Geva2021}
M.~Geva, R.~Schuster, J.~Berant, and O.~Levy.
\newblock Transformer feed-forward layers are key-value memories.
\newblock In \emph{Proceedings of the 2021 Conference on Empirical Methods in
  Natural Language Processing}, pages 5484--5495, 2021.

\bibitem{Guarino1998}
N.~Guarino.
\newblock Formal ontology and information systems.
\newblock In \emph{Proceedings of FOIS'98}, pages 3--15. IOS Press, 1998.

\bibitem{Hogan2021}
A.~Hogan, E.~Blomqvist, M.~Cochez, C.~d'Amato, G.~de~Melo, C.~Gutierrez,
  S.~Kirrane, J.~E.~L. Gayo, R.~Navigli, S.~Neumaier, A.-C. Ngonga~Ngomo,
  A.~Polleres, S.~M. Rashid, A.~Rula, L.~Schmelzeisen, J.~Sequeda, S.~Staab,
  and A.~Zimmermann.
\newblock Knowledge graphs.
\newblock \emph{ACM Computing Surveys}, 54(4):1--37, 2021.

\bibitem{Kirkpatrick2017}
J.~Kirkpatrick, R.~Pascanu, N.~Rabinowitz, J.~Veness, G.~Desjardins,
  A.~A. Rusu, K.~Milan, J.~Quan, T.~Ramalho, A.~Grabska-Barwinska,
  D.~Hassabis, C.~Clopath, D.~Kumaran, and R.~Hadsell.
\newblock Overcoming catastrophic forgetting in neural networks.
\newblock \emph{Proceedings of the National Academy of Sciences},
  114(13):3521--3526, 2017.

\bibitem{Li2024}
K.~Li, O.~Press, and S.~Arora.
\newblock How does {GPT}-2 compute greater-than?: Interpreting mathematical
  abilities in a pre-trained language model.
\newblock In \emph{Advances in Neural Information Processing Systems}, 2024.

\bibitem{Masolo2003}
C.~Masolo, S.~Borgo, A.~Gangemi, N.~Guarino, and A.~Oltramari.
\newblock {WonderWeb} deliverable {D18}: Ontology library (final).
\newblock Technical report, ISTC-CNR, 2003.

\bibitem{Liu2024}
N.~F. Liu, K.~Lin, J.~Hewitt, A.~Paranjape, M.~Bevilacqua, F.~Petroni, and
  P.~Liang.
\newblock Lost in the middle: How language models use long contexts.
\newblock \emph{Transactions of the Association for Computational Linguistics},
  12:157--173, 2024.

\bibitem{McCloskey1989}
M.~McCloskey and N.~J. Cohen.
\newblock Catastrophic interference in connectionist networks: The sequential
  learning problem.
\newblock In G.~H. Bower, editor, \emph{Psychology of Learning and Motivation},
  volume~24, pages 109--165. Academic Press, 1989.

\bibitem{Meng2022}
K.~Meng, D.~Bau, A.~Andonian, and Y.~Belinkov.
\newblock Locating and editing factual associations in {GPT}.
\newblock In \emph{Advances in Neural Information Processing Systems},
  volume~35, pages 17359--17372, 2022.

\bibitem{Miller1956}
G.~A. Miller.
\newblock The magical number seven, plus or minus two: Some limits on our
  capacity for processing information.
\newblock \emph{Psychological Review}, 63(2):81--97, 1956.

\bibitem{Pearl1988}
J.~Pearl.
\newblock \emph{Probabilistic Reasoning in Intelligent Systems: Networks of
  Plausible Inference}.
\newblock Morgan Kaufmann, 1988.

\bibitem{Shannon1948}
C.~E. Shannon.
\newblock A mathematical theory of communication.
\newblock \emph{Bell System Technical Journal}, 27(3):379--423, 1948.

\bibitem{Simon1955}
H.~A. Simon.
\newblock A behavioral model of rational choice.
\newblock \emph{The Quarterly Journal of Economics}, 69(1):99--118, 1955.

\bibitem{Sun2019}
Z.~Sun, Z.-H. Deng, J.-Y. Nie, and J.~Tang.
\newblock {RotatE}: Knowledge graph embedding by relational rotation in complex
  space.
\newblock In \emph{Proceedings of the International Conference on Learning
  Representations}, 2019.

\bibitem{Tishby2000}
N.~Tishby, F.~C. Pereira, and W.~Bialek.
\newblock The information bottleneck method.
\newblock In \emph{Proceedings of the 37th Annual Allerton Conference on
  Communication, Control, and Computing}, pages 368--377, 2000.

\bibitem{Woods1975}
W.~A. Woods.
\newblock What's in a link: Foundations for semantic networks.
\newblock In D.~G. Bobrow and A.~Collins, editors, \emph{Representation and
  Understanding: Studies in Cognitive Science}, pages 35--82. Academic Press,
  1975.

\end{thebibliography}

\end{document}
