\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}
\hypersetup{hidelinks}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

\setlength{\parskip}{4pt plus 1pt minus 1pt}
\setlength{\parindent}{0pt}

\title{Knowledge as Normalization:\\Bounded Context Separates Shared\\Propositions from Action}
\author{Patrick D.\ McCarthy}
\date{}

\begin{document}
\maketitle

\begin{abstract}
When multiple action policies share propositions and active context is
bounded, storing each policy's propositions independently produces
$O(n \cdot k)$ context load, multiplicative in the number of policies
$n$ and propositions per policy $k$. We prove that bounded context
forces the extraction of shared propositions into a separate
structure~$K$, reducing context load to $O(n + k')$ where $k' = |K|$.
This is knowledge as normalization: $K$ precipitates from the action
space~$A$ the same way normal forms precipitate from redundant
relations in Codd's relational model. The normalization is not a
design choice but a scaling constraint imposed by bounded~$C_n$.
From the normalization theorem we derive the retention criterion
(retain when benefit is uncertain; prune only when benefit is
certainly zero) and $K/A$ inseparability (the content of $K$ is
constitutively determined by $A$). These results imply that proposals to give observation and action separate
training objectives violate a structural constraint imposed by bounded context.
\end{abstract}

% ------------------------------------------------------------------
\section{Introduction}
\label{sec:intro}
% ------------------------------------------------------------------

When a tumor severs a coupling channel, the cell loses both the
information that channel carried and the actions that information
supported \cite{McCarthy2026e}. A cell that can no longer sense
tissue-level growth signals can no longer execute tissue-coordinated
quiescence. Knowledge and action are inseparable: losing one entails
losing the other. The present paper proves that this inseparability
is a structural consequence of bounded context.

McCarthy \cite{McCarthy2026a} establishes that any entity accumulating
propositions under bounded active context necessarily maintains a
directed graph with typed edges over its propositions (where
``necessary'' means selected for under competition---see
\cite{McCarthy2026a}, Remark~1). That result
answers a structural question: \emph{what form} must knowledge take?
McCarthy \cite{McCarthy2026b} establishes that control architectures
converge to escalation under bounded context. The present paper asks the
next question: \emph{why does a separated knowledge structure emerge at
all?}

The answer is a scaling constraint. Consider an entity with $n$ action
policies, each requiring $k$ propositions on average, with overlap
fraction $\beta$ (the proportion of propositions shared across two or
more policies). Without a separated knowledge structure, each policy
carries its own copy of every required proposition. Total context load
is $O(n \cdot k)$. Under bounded active context $C_n$, this
architecture fails for any non-trivial repertoire: two overlapping
policies loaded simultaneously require $2k - |\text{overlap}|$ context
slots, and without separation the overlap is not recognized: both full
copies are loaded.

With a separated structure $K$, shared propositions are stored once and
referenced by multiple policies. Total context load drops to
$O(n + k')$ where $k' = |K|$ absorbs the shared propositions. The
per-action marginal cost approaches $O(1)$ as overlap increases.

This is Codd's insight \cite{Codd1970} applied to bounded cognition.
Codd showed that redundant copies across relations produce update
anomalies and wasted storage; normal forms extract shared dependencies
into referenced tables. The same pressure operates here: bounded
context is the capacity constraint, and $K$ is the normal form.

The paper proves a single theorem, the Normalization Theorem, and
derives corollaries establishing a retention criterion for propositions,
the inseparability of $K$ and $A$, and the asymptotic compactness of $K$
toward action-relevant propositions. These results engage prior work on
ecological perception \cite{Gibson1979}, bounded rationality
\cite{Simon1955}, the information bottleneck \cite{Tishby2000,
Tishby2011}, and active inference \cite{Friston2010,
ParrPezzuloFriston2022}.

% ------------------------------------------------------------------
\section{Formal Framework}
\label{sec:framework}
% ------------------------------------------------------------------

Definitions~\ref{def:world} and~\ref{def:knowledge} recapitulate
\cite{McCarthy2026a}; Definitions~\ref{def:action}--\ref{def:infogain}
are new.

\begin{definition}[World State Space]
\label{def:world}
Let $W$ be a measurable space of possible world states. At time~$t$, the
true state $w_t \in W$ is not directly observable by any entity.
\end{definition}

\begin{definition}[Propositions and Posterior]
\label{def:knowledge}
A \emph{proposition} $p$ about $W$ is a likelihood function
$\ell_p : W \to [0,1]$. A set of propositions $\mathcal{P}$ induces a
posterior via the joint likelihood $L_{\mathcal{P}}(w)$:
\[
P(w \mid \mathcal{P}) \;\propto\; P_0(w) \cdot L_{\mathcal{P}}(w)
\]
where $P_0$ is a prior and $L_{\mathcal{P}}$ encodes the joint
evidential contribution of all propositions in $\mathcal{P}$, with
arbitrary dependencies between them (Paper~1, \cite{McCarthy2026a}).
When propositions are conditionally independent given $w$,
$L_{\mathcal{P}}(w) = \prod_{p \in \mathcal{P}} \ell_p(w)$; the
normalization theorem does not require this special case. The uncertainty about $w$ is the surprisal
$U(w, \mathcal{P}) = -\log P(w \mid \mathcal{P})$ \cite{Shannon1948, Cover2006}.
We use $K$ to denote the separated set of shared propositions that
Theorem~\ref{thm:normalization} derives from the overlap structure of
action policies. At the definitional stage, $K$ is not assumed to
exist; the theorem establishes its necessity.
\end{definition}

\begin{definition}[Action Space]
\label{def:action}
Let $\Omega$ be a set of executable actions and $W_{\text{obs}} \subseteq W$
the observable component of the world state. The \emph{action space} is
\[
A \;\subseteq\; (\mathcal{P} \times W_{\text{obs}} \to \Delta \Omega)
\]
where $\Delta \Omega$ is the space of distributions over $\Omega$ and
$\mathcal{P}$ is the entity's proposition pool
(Definition~\ref{def:knowledge}). Each $a \in A$ selects an action
distribution given propositions and current observations. For each $a$,
the \emph{grounding set} $\mathcal{P}^{[a]} \subseteq \mathcal{P}$ is
the minimal subset of $\mathcal{P}$ such that $a$'s output is fully
determined by $\mathcal{P}^{[a]}$ and $W_{\text{obs}}$.
Theorem~\ref{thm:normalization} shows that the shared propositions
across action policies separate into a structure $K$; after that
separation, we write $K^{[a]}$ for the grounding set within $K$.

Action policies span a spectrum of propositional dependence. When
$\mathcal{P}^{[a]} = \emptyset$, $a$ depends only on
$W_{\text{obs}}$, a reflex or stimulus-response mapping that requires
no propositional knowledge. When $|\mathcal{P}^{[a]}| > 0$, $a$ is
\emph{knowledge grounded}: its output depends on stored propositions.
\end{definition}

\begin{definition}[Entity]
\label{def:entity}
An entity is a tuple $E = (K, A, \sigma, \Omega)$ where $K$ is the knowledge
graph (Definition~\ref{def:knowledge}), $A$ is the informed action space
(Definition~\ref{def:action}), $\sigma : W \to W_{\text{obs}}$ is the
sensing function, and $\Omega$ is the set of executable actions.
\end{definition}

\begin{definition}[Bounded Active Context]
\label{def:context}
$C_n$ denotes the active context capacity: the maximum number of
propositions simultaneously available for inference. Let
$\mathrm{ctx}(t) \subseteq \mathcal{P}$ be the propositions in active
context at time~$t$, with $|\mathrm{ctx}(t)| \leq C_n$. The relevance
density for action policy $a$ is:
\[
\rho(a) = \frac{|\mathcal{P}^{[a]} \cap \mathrm{ctx}(t)|}{|\mathrm{ctx}(t)|}
\]
\end{definition}

\begin{definition}[Survival Threshold]
\label{def:survival}
For each domain $d$, let $U_{\text{lethal}}^d$ be the uncertainty
threshold above which the entity does not survive: if $U(w, K) >
U_{\text{lethal}}^d$ for the relevant world states in domain $d$, the
entity's continued existence is not assured.
\end{definition}

\begin{definition}[Information Gain]
\label{def:infogain}
When the entity executes action $a$ and receives feedback encoded as
proposition $p_a$, the information gain is:
\[
G(a, K) = I(p_a ; W \mid K) \;\geq\; 0
\]
where $I(\cdot\,;\cdot \mid \cdot)$ is conditional mutual information
\cite{Cover2006}.
\end{definition}

% ------------------------------------------------------------------
\section{The Normalization Theorem}
\label{sec:theorem}
% ------------------------------------------------------------------

The theorem establishes that bounded active context forces the
separation of shared propositions from action policies.

\begin{definition}[Context Load]
\label{def:load}
For an entity with action space $A$ and knowledge representation $\mathcal{R}$,
the \emph{context load} $L(\mathcal{R}, A)$ is the number of proposition
slots required to simultaneously evaluate any pair of action policies.
\end{definition}

\begin{theorem}[Normalization of Shared Propositions]
\label{thm:normalization}
Let $E = (K, A, \sigma, \Omega)$ be an entity with bounded active context
$C_n$ (Definition~\ref{def:context}), $n = |A|$ knowledge-grounded
action policies, average proposition count $k$ per policy, and overlap
fraction $\beta > 0$, defined over the raw proposition pool:
$\beta = |\{p \in \mathcal{P} : |\{a \in A : p \in \mathcal{P}^{[a]}\}| \geq 2\}| \,/\, |\mathcal{P}|$,
the proportion of propositions referenced by two or more policies.
This is well-defined before $K$ exists, since it depends only on
$\mathcal{P}$ and the grounding sets $\mathcal{P}^{[a]}$. Then:
\begin{enumerate}
\item[(N1)] \textbf{Unnormalized failure.} Under an implicit-$K$
   architecture where each policy carries its own propositions, the
   context load for evaluating two overlapping policies is
   $L_{\text{impl}} = 2k - |\text{overlap}|$, and total proposition
   storage is $n \cdot k$. For any pair $(a_1, a_2)$ such that
   $L_{\text{impl}} > C_n$, the architecture cannot simultaneously
   evaluate both policies. More generally, as $n$ grows, the fraction
   of policy pairs that exceed $C_n$ increases, progressively
   restricting the entity's ability to compare and select actions.
\item[(N2)] \textbf{Normalized architecture.} Extracting shared
   propositions into a separated structure $K$ with each policy
   referencing shared propositions via graph edges produces context load
   $L_{\text{norm}} = |K^{[a_1]} \cup K^{[a_2]}|$ for any pair
   $(a_1, a_2)$, with total storage $|K| + n \cdot (1 - \beta) \cdot k$.
   Each shared proposition is loaded once regardless of how many policies
   reference it.
\item[(N3)] \textbf{Scaling.} $L_{\text{norm}} \leq L_{\text{impl}}$
   with strict inequality whenever $\beta > 0$. In the limit $\beta \to 1$,
   total storage approaches $|K| + n$ and per-action marginal context
   cost approaches $O(1)$.
\item[(N4)] \textbf{Anomaly elimination.} Under implicit-$K$, updating
   a shared proposition $p$ requires modifying $n$ copies, producing the
   update anomaly: one copy is revised while others retain the stale
   value, yielding inconsistent action policies. Under normalized $K$,
   updating $p$ is $O(1)$ and all referencing policies see the revision
   immediately.
\end{enumerate}
Therefore, under bounded $C_n$ with $\beta > 0$, the separated $K$
architecture is necessary for scaling the action repertoire beyond
$C_n / k$ policies.
\end{theorem}

\begin{proof}
The proof proceeds by counting context slots under each architecture.

\textbf{Part (N1): Unnormalized context load.}
Under the implicit-$K$ architecture, each action policy $a_i$ stores
its own copy of every proposition it requires: the set
$K_i^{\text{local}}$ with $|K_i^{\text{local}}| = k$. When two
policies $a_1, a_2$ must be evaluated simultaneously (e.g., to compare
expected outcomes), the entity must load $K_1^{\text{local}}$ and
$K_2^{\text{local}}$ into active context. Since the copies are
independent, the overlap between $K_1^{\text{local}}$ and
$K_2^{\text{local}}$ is not recognized by the retrieval system: both
full copies are loaded, requiring $2k$ context slots minus any
structural deduplication the system performs. In the worst case
(no deduplication), $L_{\text{impl}} = 2k$. With perfect structural
deduplication, $L_{\text{impl}} = 2k - |\text{overlap}|$. In either
case, total storage across all policies is $n \cdot k$, and when
$n \cdot k > C_n$, the entity cannot represent its full repertoire.

\textbf{Part (N2): Normalized context load.}
Under the normalized architecture, shared propositions are stored once
in $K$ and each policy $a_i$ is represented as a function of $K^{[a_i]}
\subseteq K$ and $W_{\text{obs}}$. When two policies $a_1, a_2$ are
evaluated simultaneously, the entity loads $K^{[a_1]} \cup K^{[a_2]}$
into context. Shared propositions in $K^{[a_1]} \cap K^{[a_2]}$ are
loaded once. Context load is $|K^{[a_1]} \cup K^{[a_2]}| = |K^{[a_1]}|
+ |K^{[a_2]}| - |K^{[a_1]} \cap K^{[a_2]}|$. Total storage is
$|K| + n \cdot (1 - \beta) \cdot k$, where the second term accounts
for policy-specific (unshared) propositions.

\textbf{Part (N3): Scaling comparison.}
$L_{\text{norm}} = |K^{[a_1]} \cup K^{[a_2]}| \leq |K^{[a_1]}| +
|K^{[a_2]}|$, with strict inequality when
$K^{[a_1]} \cap K^{[a_2]} \neq \emptyset$ (i.e., $\beta > 0$).
In the limit $\beta \to 1$, nearly all propositions are shared:
$|K^{[a_i]}|$ is dominated by $K$, and the per-policy marginal
storage (the unshared component) approaches $O(1)$, giving total
storage $|K| + n \cdot O(1) = O(|K| + n)$.

\textbf{Part (N4): Update anomaly.}
Under implicit-$K$, a shared proposition $p$ appears in
$K_i^{\text{local}}$ for every policy $a_i$ that uses it. Updating $p$
(e.g., upon receiving new evidence) requires identifying and modifying
all $n$ copies. If any copy is missed, $a_i$ operates on stale
information while $a_j$ operates on the revision. This is exactly
Codd's update anomaly \cite{Codd1970}: redundant storage of the same
fact causes inconsistency upon partial update. Under normalized $K$,
$p$ appears once; updating it is $O(1)$ and all policies referencing
$p$ via graph edges see the update immediately.

\textbf{Necessity.}
The separation is not a design preference but a scaling constraint. Any
entity with $n \cdot k > C_n$ and $\beta > 0$ that does not extract
shared propositions will either (a)~exceed its context bound, or
(b)~restrict its action repertoire to $\lfloor C_n / k \rfloor$
policies. Since survival pressure requires maintaining action policies
across multiple domains (Definition~\ref{def:survival}), and domain
coverage grows with $n$, the context bound forces normalization for any
entity whose action repertoire is not trivially small.
\end{proof}

% ------------------------------------------------------------------
\section{Corollaries}
\label{sec:corollaries}
% ------------------------------------------------------------------

The normalization theorem establishes \emph{that} shared propositions
separate. The corollaries establish \emph{what determines their
content}.

\subsection{Retention criterion}

Once $K$ exists as a separated structure, the question becomes: which
propositions does $K$ retain? The answer follows from the cost
asymmetry inherent in bounded context under survival pressure.

\begin{lemma}[Proposition Absence Cost]
\label{lem:absence}
Let $p \in K$ be a proposition grounding some action policy $a \in A$ in
domain~$d$. Then:
\begin{enumerate}
\item[(i)] Removing $p$ from $K$ weakly increases expected uncertainty:
$H(W \mid K \setminus \{p\}) \geq H(W \mid K)$.
\item[(ii)] If $p$ is the sole grounding for a critical action in
domain~$d$, then removal may cause $U(w, K \setminus \{p\}) >
U_{\text{lethal}}^d$ for survival-relevant world states.
\item[(iii)] The cost of absent $p$ when needed,
$C_{\text{absent}}$, is bounded below by entity non-survival and
therefore dominates all finite operational costs.
\end{enumerate}
\end{lemma}

\begin{proof}
\textbf{Part (i).} Within the entity's model, the posterior is
$P(w \mid K) \propto P_0(w) \cdot L_K(w)$, where $L_K$ incorporates the
constraints from all propositions in $K$. Removing proposition $p$
removes a constraint: $L_{K \setminus \{p\}}$ is less restrictive than
$L_K$, so the posterior is weakly broader and $H(W \mid K \setminus
\{p\}) \geq H(W \mid K)$. This is model-internal monotonicity: it
holds because the entity treats its own propositions as valid
constraints.

\textbf{Model vs.\ reality.} Model-internal monotonicity does not
guarantee that the entity's model tracks reality. A proposition may be
wrong: the likelihood function may be mis-specified, the evidence may be
noise, or the world may have shifted since the evidence was collected.
In such cases, the entity's model entropy decreases (it becomes more
confident) while its true uncertainty about $W$ may increase. The
retention criterion (Corollary~\ref{cor:retention}) operates on model
entropy because that is all the entity has access to; it cannot step
outside its own model to evaluate its propositions against ground truth.
Whether model entropy tracks true entropy over time is a question about
learning quality, not about retention logic: an entity whose learning
mechanism has positive expected accuracy (the propositions it acquires
are more often right than wrong) will see its model converge toward
reality; an entity with systematically misleading propositions will not.
Paper~4 \cite{McCarthy2026d} formalizes this as the requirement
$\mathbb{E}_t[\eta_M] > 0$.
The retention criterion is correct given the entity's model; the model's
fidelity to reality is a separate property.

\textbf{Part (ii).} If $p$ is the sole grounding for the unique
policy maintaining $U(w, K) \leq U_{\text{lethal}}^d$, removing $p$
renders that policy ungrounded. No other policy suffices by assumption.

\textbf{Part (iii).} When $U > U_{\text{lethal}}^d$, the entity does
not survive in domain~$d$. Non-survival terminates all future utility,
so $C_{\text{absent}} \gg c_{\text{retain}}$ for any finite retention
cost.
\end{proof}

\begin{corollary}[Retention Criterion]
\label{cor:retention}
Let $E$ be an entity with bounded $C_n$ and normalized $K$
(Theorem~\ref{thm:normalization}). The retention criterion for
propositions in $K$ is asymmetric:
\begin{enumerate}
\item[(R)] \textbf{Retention}: $p$ is retained when there exists $a \in A$
applicable to $p$, or when the survival benefit of $p$ is uncertain.
\item[(P)] \textbf{Pruning}: $p$ is pruned only when the survival benefit
of $p$ is certainly zero across all reachable future trajectories.
\item[(D)] \textbf{Displacement} (curation regime, $|K| = K_{\max}$):
$p$ is displaced by candidate $q$ when $q$ has strictly higher net
marginal value: expected survival benefit minus maintenance cost.
\end{enumerate}
\end{corollary}

\begin{proof}
Define the option value of retaining $p$:
\[
\mathrm{OV}(p) = P\!\bigl(\exists\,\tau^* : B(p, \tau^*) > 0\bigr)
\;\cdot\; \mathbb{E}\bigl[B(p, \tau^*) \mid B(p, \tau^*) > 0\bigr]
\]
where $B(p, \tau^*)$ is the benefit of having $p$ at future time
$\tau^*$ along some reachable trajectory. This is a real option in the
sense of Dixit and Pindyck \cite{DixitPindyck1994}: retaining $p$
preserves the option to act on $p$-grounded policies.

The net value of retaining $p$ is $\mathrm{NV}_{\text{retain}} =
\mathrm{OV}(p) - c_{\text{retain}}(p, K)$, where $c_{\text{retain}}$
includes retrieval interference, edge maintenance, and context
consumption. The net value of pruning is
$\mathrm{NV}_{\text{prune}} = -c_{\text{prune}} -
P(\exists\,\tau^*) \cdot C_{\text{absent}}$.

By Lemma~\ref{lem:absence}(iii), $C_{\text{absent}} \gg
c_{\text{retain}}$ whenever $p$ grounds any survival-relevant action.
For propositions with uncertain relevance, the expected cost of wrongly
pruning ($C_{\text{absent}}$, unbounded) dominates the expected cost of
wrongly retaining ($c_{\text{retain}}$, bounded). Therefore:
\[
\mathrm{NV}_{\text{retain}} > \mathrm{NV}_{\text{prune}}
\quad\text{whenever}\quad
P\!\bigl(\exists\,\tau^* : B(p, \tau^*) > 0\bigr) > 0
\]
Retention dominates whenever there is any positive probability of future
benefit. Pruning is correct only when $P(\exists\,\tau^*) = 0$, certain
zero benefit, which is unverifiable for most propositions. In the
curation regime ($|K| = K_{\max}$), every addition requires displacement
by marginal value comparison.

The richer cost structure sharpens the result. For critical propositions,
$C_{\text{absent}}$ dominates $c_{\text{retain}}$ regardless of
maintenance cost. For non-critical propositions, as $|K|$ grows,
interference and edge costs increase, raising the retention threshold.
When $c_{\text{retain}}(p, K) > \mathrm{OV}(p)$, displacement is
correct: the entity replaces $p$ with a candidate of higher net value.
``Aggressive forgetting'' in bounded systems is displacement, not
pruning.

Note: the entity cannot classify propositions into ``survival-relevant''
and ``not'' with certainty. It can form imperfect ordinal rankings of
expected marginal value. The retention criterion (R) operates on the
binary question (default to retention); the displacement criterion (D)
operates on the ordinal question (displace the lowest-ranked occupant).
\end{proof}

\subsection{Inseparability and indexed knowledge}

\begin{corollary}[$K/A$ Inseparability]
\label{cor:inseparability}
The content of $K$ is constitutively determined by $A$. Two entities
$E_1 = (K_1, A_1, \sigma, \Omega_1)$ and $E_2 = (K_2, A_2, \sigma,
\Omega_2)$ with identical sensing functions but different action spaces
$A_1 \neq A_2$ will, in the curation regime, maintain different knowledge
graphs $K_1 \neq K_2$.
\end{corollary}

\begin{proof}
By Theorem~\ref{thm:normalization}, $K$ contains the shared propositions
across $A$. Different action spaces share different propositions:
if $a \in A_1 \setminus A_2$, propositions exclusively grounding $a$
have positive marginal value for $E_1$ but zero marginal value for
$E_2$. Under the displacement criterion
(Corollary~\ref{cor:retention}(D)), $E_2$ displaces these propositions
in favor of candidates relevant to $A_2$, while $E_1$ retains them.
Provided $A_1 \setminus A_2$ contains actions whose grounding
propositions are exclusive to those actions, $K_1 \neq K_2$.

The inseparability is at the \emph{optimization} level: systems may
implement $K$ and $A$ as distinct modules, but cannot optimize them
independently. The training objective for $K$ must reference $A$
(which propositions are action-relevant), and vice versa (which actions
are groundable given current $K$).
\end{proof}

\subsection{Compactness and forgetting}

\begin{corollary}[Asymptotic Compactness]
\label{cor:compactness}
Under curation pressure, $K$ converges toward the action-relevant
subset $\bigcup_{a \in A} K^{[a]}$.
\end{corollary}

\begin{proof}
A proposition $p$ with no connection to any current or foreseeable
action has zero estimated marginal value but positive maintenance cost
$c_{\text{retain}}(p, K) > 0$. In the growth regime ($|K| < K_{\max}$),
$p$ is retained because certain zero benefit cannot be verified
(Corollary~\ref{cor:retention}(P)). In the curation regime, any
candidate $q$ connected to some $a \in A$ has strictly higher estimated
marginal value than $p$, so $p$ is displaced
(Corollary~\ref{cor:retention}(D)). Over time, propositions with no
estimated action relevance are progressively displaced. The retained set
converges toward $\bigcup_{a \in A} K^{[a]}$ asymptotically, with the
rate depending on the accuracy of the entity's marginal-value estimates
and the frequency of curation events.
\end{proof}

\begin{corollary}[Curation and Forgetting]
\label{cor:forgetting}
Under bounded $K$ ($|K| = K_{\max}$), forgetting is correct behavior.
Curation displaces propositions of lower marginal value when higher-value
candidates arrive.
\end{corollary}

\begin{proof}
By Corollary~\ref{cor:retention}(D), displacement occurs when a
candidate $q$ has strictly higher net marginal value than the current
occupant $p$. This is not information loss but information improvement:
the entity's net survival benefit improves with each displacement.
\end{proof}

\textbf{Remark (Free context).}
Define the \emph{free context} $C_{\mathrm{free}}(t) = C_n -
|\mathrm{ctx}_{\mathrm{unresolved}}(t)|$: the portion of bounded context
available for novel problems after existing knowledge maintenance.
Every proposition retained in $K$ reduces $C_{\mathrm{free}}$ by its
retrieval interference and edge maintenance overhead. Curation under
displacement (Corollary~\ref{cor:retention}(D)) is therefore a
$C_{\mathrm{free}}$-maximizing operation: displacing low-value
propositions frees context for higher-value candidates.

% ------------------------------------------------------------------
\section{Strong Inseparability: The Encoding Continuum}
\label{sec:continuum}
% ------------------------------------------------------------------

\begin{remark}[$K$ and $A$ Are Not Natural Kinds]
\label{rem:continuum}
Given Theorem~\ref{thm:normalization} ($K$ precipitates from $A$) and
the encoding hierarchy $E(c,r)$ of \cite{McCarthy2026b} (a continuous
space parameterized by runtime cost $c$ and reversibility $r$), $K$ and
$A$ are not disjoint categories but regions on $E(c,r)$. The encoding
hierarchy is empirically instantiated by the complementary learning
systems of McClelland et al.\ \cite{McClelland1995}: hippocampus (high
cost, high reversibility, fast episodic learning) and neocortex (low
cost, low reversibility, slow statistical consolidation). Memory
consolidation is movement along $E(c,r)$, operating in both directions.
\end{remark}

\textbf{Argument.}
We exhibit both directions of flow.

\textbf{$K \to A$ (declarative to procedural).}
A proposition $p$ grounding action $a$ begins at encoding level $L_n$
(active reasoning): explicitly represented, revisable. As the entity
repeatedly applies $a$ successfully, $p$ migrates to lower encoding
levels. At $L_{n-1}$ (learned heuristic), $p$ is retrievable but not
explicitly stated. At $L_1$ (developmental encoding), $p$ is compiled
into the execution of $a$. At $L_0$ (heritable encoding), $p$ is
implicit in the wiring that produces $a$. This is the
declarative-to-procedural transition observed in skill acquisition
\cite{FittsPosner1967}: the novice follows explicit rules ($K$ at
$L_n$); the expert executes embodied competence ($A$ at $L_1$).

\textbf{$A \to K$ (procedural to declarative).}
Novel information often enters as pure action. A child touches a hot
stove and pulls back. The first encoding is $A$: \emph{don't touch
that}. No proposition about temperature or tissue damage is represented.
Only later does the entity derive the propositional structure: ``the
stove is hot,'' ``hot things burn,'' ``heat transfers from higher to
lower temperature.'' Each proposition is derived from the action, not
the other way around.

\textbf{Bidirectionality.}
At no point in either direction does information cross a boundary from
``knowledge'' to ``action.'' The encoding level determines whether
information appears propositional (high $c$, high $r$) or procedural
(low $c$, low $r$), but the information content is continuous through the
hierarchy. The entity tuple $E = (K, A, \sigma, \Omega)$ is notational
convenience: $K$ is the high-encoding region of a single structure, $A$
is the low-encoding region, and information flows in both directions.

\textbf{Implication.} The system does not ``decide'' whether information
belongs in $K$ or $A$. The encoding dynamics decide. Some information
enters as explicit propositions and compiles into action ($K \to A$).
Some enters as pure action and generates propositions only on reflection
($A \to K$). Some never becomes propositional at all: the expert's
``feel'' for a material, the athlete's timing. These are $A$ without
$K$: action knowledge that was never derived from and may never produce
explicit propositions.

This strengthens the normalization result. The theorem proves the weak
version: $K$ separates from $A$ as the normalized shared propositions.
The continuum result proves the strong version: $K$ and $A$ are the same
structure viewed at different encoding depths, with bidirectional flow.
Separating them architecturally requires classifying each piece of
information as ``world model'' or ``action policy'' at acquisition, but
for $A \to K$ information, the classification is wrong at entry; for
$K \to A$ information, the classification becomes wrong over time.

\textbf{Remark on modularity of mind.} The strongest version of the
objection that $K$ and $A$ are natural kinds comes from Fodor's
modularity thesis \cite{Fodor1983}. Fodor argues that mental modules
are informationally encapsulated: they process input without access to
information held by other modules. Carruthers \cite{Carruthers2006}
extends the argument to central cognition.

The encapsulation is real. The axis of encapsulation is not. Fodor's
modules are encapsulated along the encoding hierarchy $E(c,r)$: a
processor at $L_1$ cannot access representations at $L_n$. The visual
module occupies a region of $E(c,r)$, not a side of a $K/A$ partition.
The same module holds both knowledge (edge statistics, color constancy
priors) and action (saccade targeting, vergence control) at its encoding
level. Encapsulation runs \emph{between encoding levels}, not between
knowledge and action. Physical modularity---hippocampus, neocortex,
motor cortex, cerebellum---is $E(c,r)$ physically instantiated, not
evidence for $K/A$ as a natural partition.

% ------------------------------------------------------------------
\section{Objections and Scope}
\label{sec:objections}
% ------------------------------------------------------------------

\subsection{The overlap assumption}

The normalization theorem requires $\beta > 0$: at least some
propositions are shared across action policies. Is this assumption
restrictive?

In practice, $\beta > 0$ is the norm for any entity with a
non-trivial action repertoire. Physical laws, spatial relations,
object identities, social norms, and causal regularities ground
multiple action policies simultaneously. An entity whose action
policies share \emph{no} propositions would have a purely reflexive
repertoire---each action depending only on $W_{\text{obs}}$, with
no propositional grounding at all. Such an entity has $K = \emptyset$
and the normalization question does not arise. For any entity with
$K \neq \emptyset$ and $|A| > 1$, overlap is the expected condition.

\subsection{The scope of survival pressure}

The retention asymmetry (Corollary~\ref{cor:retention}) depends on
$C_{\text{absent}}$ dominating all finite costs, which requires
survival pressure. But most knowledge does not appear
survival-critical.

The answer turns on what the entity can know \emph{in advance} about
which propositions are survival-critical. Under genuine uncertainty
about future domains of encounter, the entity cannot partition its
knowledge into ``survival-relevant'' and ``not.'' Any proposition
might ground a survival-critical action in an unforeseen domain. The
proposition ``Paris is the capital of France'' is survival-irrelevant
on an ordinary Tuesday and survival-critical when navigating an
unfamiliar city to reach a hospital.

The cost of wrongly classifying a proposition as non-critical
($C_{\text{absent}}$) dominates the cost of wrongly classifying it
as critical ($c_{\text{retain}}$), by the same asymmetry the
corollary establishes. The conservative criterion therefore extends
to all propositions whose future relevance is uncertain---which, for
any finite entity, is nearly all propositions.

For entities genuinely not under survival pressure---archival databases,
static knowledge stores---the retention asymmetry weakens.
$C_{\text{absent}}$ is finite, and the symmetric marginal value
comparison of the curation regime becomes the primary decision rule.

\subsection{The unfalsifiable pruning criterion}

The pruning condition (P) requires $P(\exists\,\tau^*) = 0$: certain
zero benefit across all reachable trajectories. No finite agent can
verify this. Does this make pruning operationally vacuous?

The unfalsifiability is a feature, not a defect. It explains three
empirical phenomena:

\textbf{(1) Knowledge hoarding in the growth regime.} Biological
systems accumulate enormous quantities of seemingly useless information
before capacity forces curation. This is the behavior
Corollary~\ref{cor:retention} prescribes: in the growth regime, the
unfalsifiable pruning criterion means \emph{never voluntarily discard}.

\textbf{(2) Forgetting is always forced, never voluntary.} Biological
forgetting is driven by capacity pressure rather than deliberate
pruning \cite{AndersonSchooler1991}. The corollary explains why:
the pruning condition cannot be verified, so forgetting happens only
via displacement (D).

\textbf{(3) The asymmetry operates within curation.} The displacement
criterion is not symmetric ``higher value wins.'' The burden of proof
falls on displacement: the challenger must be strictly better. Ties go
to the incumbent.

% ------------------------------------------------------------------
\section{Engagement with Prior Work}
\label{sec:discussion}
% ------------------------------------------------------------------

\textbf{Codd \cite{Codd1970}.}
The normalization theorem is the knowledge-system analogue of Codd's
relational normalization. Codd identified that redundant storage of
data across relations produces update anomalies and wasted capacity;
normal forms extract shared dependencies into referenced tables. We
prove that the same pressure operates on bounded cognitive systems:
shared propositions across action policies produce redundant context
load and update inconsistency; $K$ is the normal form that eliminates
both. The contribution is showing that normalization is not merely a
good engineering practice but a necessary consequence of bounded context.

\textbf{Separate training objectives for observation and action.}
Corollary~\ref{cor:inseparability} shows that separate
training objectives for observation and action modules violate a
structural constraint: the content of the world model must reference
the action space (which propositions are action-relevant), and the
action space must reference the world model (which actions are
groundable). Separate modules with a shared loss is compatible;
separate training objectives is not.

\textbf{Information Bottleneck \cite{Tishby2000, Tishby2011}.}
The information bottleneck retains information relevant to an output
variable subject to a rate constraint, with threshold controlled by
Lagrange multiplier $\beta$. Our retention criterion
(Corollary~\ref{cor:retention}) recovers a version of this trade-off
but derives the threshold rather than positing it: under survival
pressure, $C_{\text{absent}}$ is bounded below by entity non-survival,
producing a maximally conservative retention rule. The normalization
theorem adds a structural result the information bottleneck lacks:
not just \emph{which} information to retain, but \emph{how} to
organize it (shared propositions separated from policies).

\textbf{Active inference \cite{Friston2010, ParrPezzuloFriston2022}.}
Active inference argues that perception and action are jointly optimized
under variational free energy minimization. This is the closest existing
framework to $K/A$ inseparability. What
Theorem~\ref{thm:normalization} adds is the organizational constraint
under hard capacity limits. When context is unbounded, all information
can be retained and the coupling is a soft preference. Under bounded
$C_n$, the coupling becomes a hard constraint: propositions must be
displaced, and the displacement criterion must reference the action
space. Active inference tells the entity what to do; the normalization
theorem tells it how to organize what it knows while doing it.

\textbf{Gibson \cite{Gibson1979}.}
Gibson's affordances---properties of the environment specified relative to
an organism's action capabilities---are knowledge indexed by action under
a different name. Corollary~\ref{cor:inseparability} formalizes Gibson's
insight. What the normalization theorem adds is a structural account:
affordances are not merely perceived but \emph{factored out} from action
policies as shared propositions, and this factoring is forced by bounded
capacity.

\textbf{Anderson and Schooler \cite{AndersonSchooler1991}.}
They demonstrated that memory retention statistics mirror environmental
statistics. This is an empirical prediction of
Corollary~\ref{cor:retention}: if retention is indexed by action
relevance, and action relevance tracks environmental statistics, then
retention curves should mirror environmental frequency
distributions---precisely what they observe.

\textbf{Simon \cite{Simon1955}.}
The retention criterion is a satisficing criterion in Simon's sense.
The entity does not compute the optimal knowledge set (a combinatorial
problem) but applies a simple rule: retain if benefit is not certainly
zero, displace if a better candidate arrives. When the benefit function
is submodular, the greedy criterion achieves a $(1 - 1/e)$-approximation
to optimal \cite{Nemhauser1978}. In the general case, the guarantee does
not apply, but the criterion remains a well-defined satisficing
heuristic.

% ------------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}
% ------------------------------------------------------------------

Knowledge as normalization operates at two levels.

The \emph{structural} level (Theorem~\ref{thm:normalization}):
bounded active context with overlapping action policies forces the
extraction of shared propositions into a separated structure $K$,
converting multiplicative context load ($O(n \cdot k)$) to additive
load ($O(n + k')$) and eliminating update anomalies. This is not a
design choice but a scaling constraint. The entity that does not
normalize cannot scale its action repertoire beyond $C_n / k$ policies.

The \emph{content} level (Corollaries~\ref{cor:retention}--\ref{cor:inseparability}):
the propositions retained in $K$ are constitutively determined by
$A$. The retention criterion is asymmetric (retain when benefit is
uncertain; prune only when certainly zero) and derived from the cost
structure of bounded context under survival pressure, not from a
design parameter. Systems may implement $K$ and $A$ as separate
modules, but cannot give them independent training objectives without
accepting information loss or reconstruction cost.

The \emph{strong} inseparability (Remark~\ref{rem:continuum}): $K$
and $A$ are not natural kinds but regions on the encoding hierarchy
$E(c,r)$, with information flowing in both directions. The boundary
is not merely blurry; it is incoherent as a classification criterion.

The knowledge graph precipitates from the action space as the
normalization of shared propositions. It is not a store of declarative
facts separate from procedural skills; it is the shared structure
underlying both, factored out to avoid redundancy under bounded
capacity.

\section*{Acknowledgments}

The author used Claude (Anthropic) for drafting assistance, literature review, and \LaTeX{} preparation. All theoretical content, proofs, and arguments are the author's own.

\bibliographystyle{plain}
\begin{thebibliography}{99}

\bibitem{Codd1970}
E.~F. Codd.
\newblock A relational model of data for large shared data banks.
\newblock \emph{Communications of the ACM}, 13(6):377--387, 1970.

\bibitem{AndersonSchooler1991}
J.~R. Anderson and L.~J. Schooler.
\newblock Reflections of the environment in memory.
\newblock \emph{Psychological Science}, 2(6):396--408, 1991.

\bibitem{Cover2006}
T.~M. Cover and J.~A. Thomas.
\newblock \emph{Elements of Information Theory}.
\newblock Wiley-Interscience, 2nd edition, 2006.

\bibitem{Carruthers2006}
P.~Carruthers.
\newblock \emph{The Architecture of the Mind: Massive Modularity and the
  Flexibility of Thought}.
\newblock Oxford University Press, 2006.

\bibitem{DixitPindyck1994}
A.~K. Dixit and R.~S. Pindyck.
\newblock \emph{Investment Under Uncertainty}.
\newblock Princeton University Press, 1994.



\bibitem{Fodor1983}
J.~A. Fodor.
\newblock \emph{The Modularity of Mind}.
\newblock MIT Press, 1983.

\bibitem{Friston2010}
K.~Friston.
\newblock The free-energy principle: A unified brain theory?
\newblock \emph{Nature Reviews Neuroscience}, 11(2):127--138, 2010.

\bibitem{ParrPezzuloFriston2022}
T.~Parr, G.~Pezzulo, and K.~J. Friston.
\newblock \emph{Active Inference: The Free Energy Principle in Mind, Brain, and
  Behavior}.
\newblock MIT Press, 2022.

\bibitem{FittsPosner1967}
P.~M. Fitts and M.~I. Posner.
\newblock \emph{Human Performance}.
\newblock Brooks/Cole, 1967.

\bibitem{Gibson1979}
J.~J. Gibson.
\newblock \emph{The Ecological Approach to Visual Perception}.
\newblock Houghton Mifflin, 1979.

\bibitem{McClelland1995}
J.~L. McClelland, B.~L. McNaughton, and R.~C. O'Reilly.
\newblock Why there are complementary learning systems in the hippocampus and
  neocortex: Insights from the successes and failures of connectionist models
  of learning and memory.
\newblock \emph{Psychological Review}, 102(3):419--457, 1995.

\bibitem{McCarthy2026a}
P.~D. McCarthy.
\newblock Graph structure is necessary for information preservation under
  bounded context.
\newblock Technical report, 2026.

\bibitem{McCarthy2026b}
P.~D. McCarthy.
\newblock Convergence of control architectures to escalation under bounded context.
\newblock Working paper, 2026.

\bibitem{McCarthy2026d}
P.~D. McCarthy.
\newblock Evaluation-driven descent, encoding permanence, and the structural
  invariants of intelligence.
\newblock Manuscript, 2026.

\bibitem{Nemhauser1978}
G.~L. Nemhauser, L.~A. Wolsey, and M.~L. Fisher.
\newblock An analysis of approximations for maximizing submodular set
  functions --- {I}.
\newblock \emph{Mathematical Programming}, 14(1):265--294, 1978.

\bibitem{Shannon1948}
C.~E. Shannon.
\newblock A mathematical theory of communication.
\newblock \emph{Bell System Technical Journal}, 27(3):379--423, 1948.

\bibitem{Simon1955}
H.~A. Simon.
\newblock A behavioral model of rational choice.
\newblock \emph{The Quarterly Journal of Economics}, 69(1):99--118, 1955.

\bibitem{Tishby2000}
N.~Tishby, F.~C. Pereira, and W.~Bialek.
\newblock The information bottleneck method.
\newblock In \emph{Proceedings of the 37th Annual Allerton Conference on
  Communication, Control, and Computing}, pages 368--377, 2000.

\bibitem{Tishby2011}
N.~Tishby and N.~Zaslavsky.
\newblock Deep learning and the information bottleneck principle.
\newblock In \emph{IEEE Information Theory Workshop}, pages 1--5, 2015.

\bibitem{McCarthy2026e}
P.~D. McCarthy.
\newblock Coupling channels, severance hierarchy, and the prognostic
structure of cancer: an empirical test of bounded-context theory.
\newblock \emph{SSRN preprint}, 2026.

\end{thebibliography}

\end{document}
