\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{array}
\usepackage{graphicx}
\usepackage{hyperref}
\hypersetup{hidelinks}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}{Proposition}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem*{remark}{Remark}

\setlength{\parskip}{4pt plus 1pt minus 1pt}
\setlength{\parindent}{0pt}

\title{Context Window Diseases and Cancer Risk:\\Neural Computation Cost as a Unifying Mechanism}
\author{Patrick D.\ McCarthy}
\date{}

\begin{document}
\maketitle

\begin{abstract}
We propose that psychiatric disorders classifiable as context window
failures share molecular machinery with cancer and that the
direction of cancer risk depends on the failure mode. Mild context
inefficiency, where the neural context-switching machinery operates
at high cost per bit of uncertainty reduced, predicts elevated
cancer risk. Severe context failure, where the machinery is disabled,
predicts reduced cancer risk. This yields an inverted-U relationship
between context dysfunction severity and cancer incidence.

The mechanism: neural computation requires topoisomerase-mediated DNA
double-strand breaks for gene expression during learning and memory
formation. Each context switch imposes a DNA damage cost independent
of whether the switch produced useful knowledge. Disorders
characterized by excessive context switching (ADHD, anxiety),
context locking (PTSD, OCD), or context narrowing (depression)
accumulate DNA damage at elevated rates relative to knowledge
acquired. The chromatin remodeling genes that repair this damage,
\emph{ARID1A}, \emph{KMT2C}, \emph{SMARCA4}, \emph{DNMT3A}, are
the same genes whose somatic mutation drives cancer progression.

We formalize this using the bounded-context framework
\cite{McCarthy2026a}, deriving a mutation cost rate proportional to
the inverse of effective channel capacity $C_n$. We classify
psychiatric conditions by context window failure mode, map each to
its known cancer risk profile, and show that the inverted-U pattern
is consistent across sixteen conditions. We propose a three-tier
clinical protocol: interrupt the thrash (somatic techniques),
optimize the scheduler (cognitive therapy), tune the hardware
(pharmacotherapy); with somatic mutation burden in chromatin
remodeling genes from circulating free DNA as the molecular
biomarker linking psychological intervention to cancer risk
reduction. A meta-tier---molecular monitoring of chromatin
remodeling gene mutations via liquid biopsy---operates on a
different problem space: the escalation chain itself. Where
the three treatment tiers optimize the context window, the
meta-tier reads the state of the escalation hierarchy directly,
detecting both accumulating damage and somatic modifications
that have already altered the chain.

Twelve falsifiable predictions are stated.
\end{abstract}

% ------------------------------------------------------------------
\section{Introduction}
\label{sec:intro}
% ------------------------------------------------------------------

Learning has a physical cost. Neurons break their own DNA to learn.

Topoisomerase II$\beta$ creates double-strand breaks in the promoter
regions of immediate early genes during memory formation
\cite{Madabhushi2015}. These breaks are necessary; without them,
the chromatin cannot open fast enough for rapid transcriptional
response. The breaks are repaired, but repair is imperfect.
Each learning event leaves a residue of somatic mutation.

This is not damage from an external insult. It is the cost of
computation. The cell breaks its DNA on purpose, reads the gene,
repairs the break, and moves on. Over a lifetime, the repair
errors accumulate.

The question is: does the \emph{efficiency} of learning determine
the rate of accumulation?

We argue that it does. Psychiatric disorders are classifiable by
how they degrade the efficiency of the neural context window.
The resulting excess DNA damage cost explains an otherwise
puzzling pattern in the epidemiological literature---mild
psychiatric disorders elevate cancer risk, while severe ones
reduce it.

\subsection{The cost per bit}

In the bounded-context framework \cite{McCarthy2026a}, an agent
reduces its uncertainty about the world $H(W \mid K)$ by acquiring
knowledge $K$ through a channel with capacity $C_n$. The number
of computational cycles required to reduce uncertainty by one bit
is bounded below by $1/C_n$. Each cycle, in a biological neural
network, involves gene expression changes that require DNA breaks.

The mutation cost per bit of uncertainty reduced is therefore:

\begin{equation}
\label{eq:cost}
\mu = \frac{\delta \cdot N_{\text{cycles}}}{\Delta H}
    \geq \frac{\delta}{C_n}
\end{equation}

where $\delta$ is the per-cycle probability of a repair error
(approximately $10^{-9}$ per base pair per repair event
\cite{Lindahl1993}), $N_{\text{cycles}}$ is the number of
computation cycles, and $\Delta H$ is the information gained.

An efficient learner has high $C_n$: each cycle reduces uncertainty
productively. The DNA damage cost per bit is near the minimum
$\delta / C_n$.

A context window disorder reduces effective $C_n$ without reducing
$N_{\text{cycles}}$. The system computes at the same rate but
learns less per cycle. The mutation cost per bit rises. Over
decades, the difference compounds.

% ------------------------------------------------------------------
\section{Context Window Failure Taxonomy}
\label{sec:taxonomy}
% ------------------------------------------------------------------

We classify psychiatric conditions by how they degrade the context
window. The context window, in bounded-context theory
\cite{McCarthy2026a}, is the set of propositions simultaneously
available for inference. It has a hard capacity bound $C_n$ and
supports four operations: load, hold, release, and switch.

Each operation can fail. The failure mode determines the
computational waste pattern and, we argue, the cancer risk profile.

\subsection{Premature eviction: context cannot hold}

The window evicts propositions before they are consolidated.
Each eviction triggers a full context reload---a new round of gene
expression changes with associated DNA breaks---for a partial or
zero information gain.

\textbf{ADHD.} The canonical premature eviction disorder.
Attention shifts before consolidation completes. The system
thrashes: high cycle count, low bits per cycle, elevated
$\mu$. Hyperfocus episodes represent the complementary failure;
the window locks on a single context and refuses to release,
which is costly only if the locked context is not productive.

A large Swedish registry study ($n \approx 4$ million) found
ADHD associated with elevated overall cancer risk after
adjustment for lifestyle confounders \cite{Sun2022}. ADHD shares
genetic variants with cell cycle regulation and chromatin
remodeling pathways, the same coupling channels identified in
\cite{McCarthy2026e} as cancer-critical.

\textbf{Traumatic brain injury.} Physical damage to the context
buffer. Elevated glioma risk is well-documented \cite{Chen2012tbi}.
The mechanism in our framework: damaged tissue requires elevated
repair activity, consuming DDR channel capacity that would
otherwise catch somatic mutations.

\subsection{Context lock: context cannot release}

The window loads a context and cannot release it. New evidence
cannot enter. The locked context replays repeatedly, each replay
incurring the full DNA damage cost of gene expression, while the
information gain is zero (the replay teaches nothing new).

\textbf{PTSD.} Traumatic context replays involuntarily. Each
flashback activates the same neural circuits, the same gene
expression programs, the same DNA breaks, in promoter regions
already weakened by prior replay. Large Veterans Affairs studies
show elevated cancer incidence in PTSD patients after adjustment
for substance use and other behavioral confounders
\cite{Gradus2015}. The standard explanation is cortisol-mediated
immune suppression. Our framework adds a mechanism---the replay
loop imposes a DNA damage cost with $\Delta H = 0$ per cycle.

\textbf{OCD.} Checking loops. The context window loads a threat
assessment, cannot resolve it, clears, reloads, re-checks. Each
cycle costs DNA breaks. No uncertainty is reduced because the
check never terminates. Some epidemiological evidence supports
mildly elevated cancer risk \cite{Shen2013}.

\textbf{Addiction.} The reward context overrides the scheduler.
The window is permanently loaded with the reward-seeking program,
suppressing context switches that would serve learning. Neural
circuits associated with the addictive behavior are chronically
active; associated DNA damage accumulates in those circuits
specifically. Elevated cancer risk in addiction is well-documented,
typically attributed to carcinogen exposure (smoking, alcohol).
Our framework predicts elevated risk even for non-carcinogen
addictions (gambling, behavioral addiction) through the context
lock mechanism alone. Limited evidence is consistent with this
\cite{Tao2017}.

\subsection{Context narrowing: window contracts}

The window's effective capacity shrinks. Fewer propositions are
loaded per cycle. The system still computes, still breaks DNA,
but each cycle processes a diminished context.

\textbf{Depression.} The context window collapses to a narrow
negative prior. Rumination, the repetitive reprocessing of
negative content, is computation with near-zero information gain.
The system runs at full metabolic and DNA damage cost but acquires
no new knowledge. Depression approaches a \textbf{near-zero-$C_n$ state with
nonzero compute cost}: as effective capacity $C_n^{\text{eff}}
\to 0$, the mutation cost per bit $\mu \to \infty$. The system
pays the full DNA damage cost of computation while acquiring
almost nothing.

Meta-analyses consistently show elevated cancer risk in depression
\cite{Chida2008}. The standard explanation involves cortisol,
inflammatory cytokines, and behavioral factors (reduced screening,
poor diet). Our framework does not compete with these mechanisms.
It adds a direct molecular channel: rumination-driven DNA damage
in neurons and, via systemic stress signaling, in peripheral
tissues.

\textbf{Anxiety.} The context window is too wide. Everything is
loaded as a potential threat. The system evaluates an excessive
number of propositions per cycle; most are irrelevant. The
information gain per cycle is diluted by the noise. Mixed
epidemiological evidence: some studies show mildly elevated cancer
risk \cite{Batty2017}.

\subsection{Context dissolution: boundaries fail}

The window loses the ability to maintain coherent boundaries
between contexts. Unrelated propositions bleed into each other.
The system cannot maintain the separation between ``self'' and
``other,'' between ``current context'' and ``irrelevant context.''

\textbf{Schizophrenia.} Context boundaries dissolve.
Schizophrenia has \emph{reduced} cancer risk in multiple
large epidemiological studies \cite{Catts2008, Ji2013}. This
is one of the most replicated and least explained findings in
psychiatric epidemiology.

In our framework, the explanation is direct. Schizophrenia
disables the context-switching machinery. Chromatin remodeling
genes that are risk factors for schizophrenia overlap substantially
with cancer-critical genes \cite{Singh2016}. If these genes are
already non-functional (from germline variants that cause
schizophrenia), they cannot be \emph{further} corrupted by somatic
mutation to enable cancer. Moreover, cancer cells require
functional chromatin remodeling to execute the gene expression
reprogramming necessary for tumor progression, the switch from
``normal cell context'' to ``invasive metastatic context.'' A
system in which the context-switching machinery is broken cannot
execute this transition.

This predicts that antipsychotic treatment, which partially
restores context-switching function, should move cancer risk back
toward baseline. Evidence supports this: antipsychotic use in
schizophrenia is associated with increased cancer risk relative
to unmedicated patients \cite{Solmi2020}. The standard explanation
is metabolic side effects. Our framework offers a mechanistic
alternative: restoring the chromatin remodeling machinery
re-enables the context switches that cancer requires.

\subsection{Context index failure: retrieval breaks}

The window can still hold propositions, but the retrieval system
that selects which propositions to load is corrupted. The wrong
context is loaded, or no context can be loaded at all.

\textbf{Alzheimer's disease.} The context index is destroyed.
Like schizophrenia, Alzheimer's has \emph{reduced} cancer risk.
The inverse comorbidity between Alzheimer's and cancer is among
the most robust findings in epidemiology \cite{Driver2012,
Musicco2013}. The shared molecular basis centers on TP53:
activation of p53 kills cells (prevents cancer, promotes
neurodegeneration); loss of p53 lets cells survive (enables
cancer, may protect neurons) \cite{Behrens2009}. In our
framework: the same checkpoint that decides ``this cell's context
is corrupted, kill it'' is the gate between the neurodegeneration
attractor and the cancer attractor. They are opposite failure
modes of the same context-integrity check.

\subsection{Context oscillation}

The window oscillates between incompatible contexts, spending
resources on the transition without productive use of either.

\textbf{Bipolar disorder.} Context frames oscillate between manic
(window too wide, loading everything, acting on incomplete
evaluation) and depressive (window collapsed, loading nothing new).
Both phases are computationally expensive. Mixed evidence: some
studies show mildly elevated cancer risk \cite{McGinty2012}.

\subsection{Context rigidity}

The window loads a context and holds it with excessive tenacity.
Incoming evidence that does not fit the current frame is rejected
rather than integrated. The system functions but at reduced
effective capacity---fewer propositions are considered per cycle
because the acceptance filter is too narrow.

\textbf{Autism spectrum.} The context window is rigid: strong
priors resist updating, sensory contexts that neurotypical
systems filter are loaded at full resolution, and transitions
between contexts are costly. The cancer risk profile is mixed
\cite{Darbro2016}, consistent with a heterogeneous severity
spectrum---mild rigidity elevates cost per bit (the system
over-processes), while severe rigidity reduces context-switching
frequency (fewer cycles, less DNA damage).

\subsection{Consolidation failure}

The window operates during waking hours but the offline
consolidation process---sleep---fails to compress episodic
context into durable semantic structure. The proposition count
$n$ drifts toward the maintenance crossover $n^*$ because
compression cannot keep pace with acquisition.

\textbf{Insomnia.} Sleep consolidation is the brain's primary
mechanism for reducing maintenance load (Section~\ref{sec:crossover}).
Chronic insomnia degrades this process. The context window
accumulates uncompressed propositions, each requiring active
maintenance. Elevated cancer risk in chronic insomnia is
consistent with the framework: the maintenance cost of
unconsolidated knowledge drives excess chromatin remodeling
demand \cite{Lu2017}.

\subsection{Context initiation failure}

The window cannot initiate a new context. The motor and
dopaminergic circuits that translate a loaded context into
action degrade, so the system loads contexts but cannot act
on them. The stalled initiation loop consumes resources
without producing behavioral output.

\textbf{Parkinson's disease.} The dopaminergic system that
gates context-to-action translation degrades. Unlike
schizophrenia---where the context-switching machinery itself
is broken---Parkinson's preserves the ability to load contexts
but impairs the ability to execute on them. Cancer risk is
reduced for most solid tumors \cite{Bajaj2010}. The framework
explains this through a shared mechanism: the same dopaminergic
circuits that fail in Parkinson's are required for the
rapid gene-expression reprogramming that tumor cells use
to switch context during invasion and metastasis.

\subsection{Reduced baseline capacity}

The context window has a constitutively smaller capacity from
development. Fewer propositions can be loaded and maintained.
The system operates within its reduced bounds---it does not
thrash because it never attempts loads beyond its capacity.

\textbf{Down syndrome.} Trisomy 21 reduces effective context
capacity constitutively. Solid tumor risk is markedly reduced
\cite{Hasle2000}---the extra copy of chromosome 21 encodes
several tumor suppressors, but in our framework the deeper
mechanism is that the reduced context-switching demand means
less chromatin remodeling wear. (Leukemia risk is elevated,
consistent with the immune channel operating under different
constraints.)

\textbf{Intellectual disability (other causes).} Germline
variants that reduce cognitive capacity also reduce the
demand on chromatin remodeling machinery. Some evidence
supports reduced cancer risk in this population \cite{Patja2001},
consistent with the framework: lower context-switching load
means fewer activity-dependent DNA breaks in the genes that
cancer requires.

% ------------------------------------------------------------------
\section{The Inverted-U Prediction}
\label{sec:inverted-u}
% ------------------------------------------------------------------

The taxonomy yields a structural prediction. Cancer risk is not
monotonically related to context dysfunction severity. It follows
an inverted U.

\begin{definition}[Context dysfunction severity]
For an agent with baseline channel capacity $C_n^*$, context
dysfunction severity $s$ is defined as:
\begin{equation}
s = 1 - \frac{C_n^{\text{eff}}}{C_n^*}
\end{equation}
where $C_n^{\text{eff}}$ is the agent's effective channel capacity
under the disorder. $s = 0$ for healthy function; $s = 1$ for
complete context machinery failure.
\end{definition}

\begin{proposition}[Inverted-U cancer risk]
\label{prop:inverted-u}
Let $R(s)$ be cancer risk as a function of context dysfunction
severity. Then $R(s)$ has an interior maximum: there exists
$s^* \in (0, 1)$ such that $R(s^*) > R(0)$ and $R(s^*) > R(1)$.
\end{proposition}

\begin{remark}[Argument]
The mutation accumulation rate from neural computation is:
\begin{equation}
\dot{M} = \delta \cdot f(s) \cdot g(s)
\end{equation}
where $f(s)$ is the computation rate (cycles per unit time) and
$g(s)$ is the fraction of cycles that produce non-productive DNA
breaks (breaks not offset by adaptive benefit).

At $s = 0$ (healthy): $f$ is moderate, $g$ is low (most
computation is productive). $\dot{M}$ is at baseline.

At intermediate $s$ (ADHD, depression, PTSD): $f$ remains high
(the system still computes actively, even more so in anxiety
and rumination), but $g$ is high (most cycles are wasted).
$\dot{M}$ is elevated.

At $s = 1$ (schizophrenia, severe Alzheimer's): $f$ drops
because the computation machinery is disabled. $g$ is undefined
because the machinery does not produce coherent cycles.
$\dot{M}$ drops below baseline. Moreover, cancer cells
themselves require functional context-switching machinery to
progress; disabling it at the germline level is tumor-suppressive.

Therefore $\dot{M}$ (and consequently $R$) rises from $s = 0$
to intermediate $s$, then falls to $s = 1$, producing an interior
maximum.
\end{remark}

The available epidemiological evidence is consistent with this
prediction. Table~\ref{tab:context-diseases} maps sixteen
conditions to their context failure mode, dysfunction severity,
and observed cancer risk.

\begin{table}[ht]
\centering
\caption{Context window disorders and cancer risk. Severity is
qualitative (L = low, M = moderate, H = high). Cancer risk
direction from meta-analyses and large registry studies.}
\label{tab:context-diseases}
\small
\begin{tabular}{@{}llccc@{}}
\toprule
\textbf{Condition} & \textbf{Failure mode} & \textbf{Severity} &
\textbf{Cancer risk} & \textbf{Key reference} \\
\midrule
ADHD & Premature eviction & L--M & $\uparrow$ & \cite{Sun2022} \\
TBI & Buffer damage & L--M & $\uparrow$ & \cite{Chen2012tbi} \\
PTSD & Context lock & M & $\uparrow$ & \cite{Gradus2015} \\
OCD & Check loop & L--M & $\uparrow$ (mild) & \cite{Shen2013} \\
Addiction & Reward lock & M & $\uparrow$ & multiple \\
Depression & Context narrowing & M & $\uparrow$ & \cite{Chida2008} \\
Anxiety (GAD) & Window too wide & L--M & $\uparrow$ (mild) & \cite{Batty2017} \\
Bipolar & Context oscillation & M & $\uparrow$ (mild) & \cite{McGinty2012} \\
Autism & Rigid context & L--H & mixed & \cite{Darbro2016} \\
Insomnia & Consolidation failure & L--M & $\uparrow$ & \cite{Lu2017} \\
Autoimmune & Context mis-indexing & M & $\uparrow$ & \cite{Franks2012} \\
\midrule
Schizophrenia & Context dissolution & H & $\downarrow$ & \cite{Catts2008, Ji2013} \\
Alzheimer's & Index destruction & H & $\downarrow$ & \cite{Driver2012, Musicco2013} \\
Parkinson's & Context initiation & H & $\downarrow$ (most) & \cite{Bajaj2010} \\
Down syndrome & Reduced capacity & H & $\downarrow$ (solid) & \cite{Hasle2000} \\
Intellectual disability & Reduced capacity & H & $\downarrow$ (some) & \cite{Patja2001} \\
\bottomrule
\end{tabular}
\end{table}

The table divides cleanly. Every condition with moderate
dysfunction severity and intact computation machinery shows
elevated cancer risk. Every condition with severe dysfunction
or disabled computation machinery shows reduced cancer risk.
The boundary falls where the machinery transitions from
``working hard but inefficiently'' to ``broken.''

% ------------------------------------------------------------------
\section{The Shared Molecular Machinery}
\label{sec:machinery}
% ------------------------------------------------------------------

The context window framework predicts that psychiatric context
disorders and cancer should share molecular machinery. They do.

\subsection{Chromatin remodeling: the read/write heads}

The genes that control chromatin accessibility, determining which
genes are readable in which cellular context, are mutated in both
psychiatric disorders and cancer:

\begin{itemize}
\item \emph{ARID1A}: mutated in 6--8\% of all cancers; germline
variants associated with intellectual disability
\cite{Santen2012}.
\item \emph{KMT2C}/\emph{KMT2D}: among the most frequently mutated
genes in cancer; Kabuki syndrome (germline \emph{KMT2D} loss)
involves intellectual disability \cite{Miyake2013}.
\item \emph{SMARCA4}: mutated in lung, ovarian, and other cancers;
Coffin-Siris syndrome (germline loss) involves intellectual
disability \cite{Tsurusaki2012}.
\item \emph{DNMT3A}: mutated in AML; germline variants associated
with Tatton-Brown-Rahman syndrome (overgrowth and intellectual
disability) \cite{Tatton-Brown2014}.
\end{itemize}

These are not coincidences. In the bounded-context framework,
chromatin remodeling genes are the molecular implementation of
context switching. They decide which genes are accessible
(loaded into the context window) in a given cell state. When they
fail by somatic mutation, the cell loses context control: it
expresses genes that do not belong in its current state. That is
cancer. When they fail by germline mutation, the organism loses
context control at the neural level: it cannot efficiently switch
cognitive contexts. That is intellectual disability.

The shared machinery explains both the elevated cancer risk in
moderate context disorders (chromatin remodeling genes are
overworked and accumulate somatic damage) and the reduced cancer
risk in severe ones (germline loss of these genes prevents the
context switches cancer requires).

\subsection{TP53: the context integrity checkpoint}

TP53 is the most frequently mutated gene in cancer
\cite{Kandoth2013} and a central node in the coupling-channel
framework \cite{McCarthy2026e}. It functions as a context
integrity check: if a cell's DNA is too damaged to maintain
faithful gene expression context, p53 activates apoptosis.

In our framework, p53 is the \emph{kill switch} for context window
corruption. Its dual role in cancer (loss enables tumor growth)
and neurodegeneration (activation kills neurons) is not a paradox.
It is the same function, context integrity enforcement, applied
in two domains where the optimal threshold differs.

Cancer patients need p53 active (to kill context-corrupted cells).
Alzheimer's patients need p53 less active (to preserve neurons
whose context is degrading with age). The inverse comorbidity
between cancer and Alzheimer's \cite{Driver2012} reflects the
organism's position on a single axis: how aggressively it
enforces context integrity.

\subsection{The immune system as somatic hypermutation}

The immune system stores knowledge in DNA through somatic
hypermutation in B cells. This is not a metaphor for the context
window framework. It is a literal implementation. The immune
system:

\begin{enumerate}
\item Encounters an antigen (new evidence).
\item Mutates its own DNA to generate antibody variants (context
switching at the molecular level).
\item Selects the variant that best matches (uncertainty reduction).
\item Stores the result as memory B cells (knowledge encoded in DNA).
\end{enumerate}

Autoimmune diseases are context indexing failures: the retrieval
system misidentifies self as threat, loading the wrong context.
Autoimmune diseases are associated with elevated cancer risk
\cite{Franks2012}, consistent with our framework: a context
retrieval system that loads the wrong targets damages tissue
through chronic inflammation and imposes DNA damage costs on
cells that are not the intended target.

% ------------------------------------------------------------------
\section{Clinical Protocol}
\label{sec:protocol}
% ------------------------------------------------------------------

If context-window efficiency is mechanistically upstream of
somatic mutation accumulation, then treating context disorders is
cancer prevention. The standard oncology framing treats psychiatric
comorbidities as quality-of-life concerns. Our framework reframes
them as disease-modifying targets.

We propose a three-tier treatment protocol ordered by intervention
speed, cost, and side-effect profile. A meta-tier---molecular
monitoring---operates on a different problem space entirely:
not the context window but the escalation chain itself,
detecting both accumulating repair burden and somatic
modifications that have already altered the hierarchy.

\subsection{Tier 1: Interrupt the thrash}

\textbf{Objective:} Stop the runaway context loop immediately.
Reduce $N_{\text{cycles}}$ per unit time by providing a manual
interrupt.

\textbf{Interventions:}
\begin{itemize}
\item \textbf{Somatic techniques} (EFT/tapping, grounding
exercises, box breathing). These provide a hardware interrupt:
when the cognitive context window is thrashing, a somatic stimulus
forces attention to the body, evicting the runaway cognitive loop.
EFT produces a 24\% cortisol reduction in a single session
\cite{Church2012}. Cortisol is a downstream marker; the upstream
event is the interrupted context loop.
\item \textbf{Bilateral stimulation} (EMDR eye movements, bilateral
tapping). Forces alternating hemispheric activation, which
disrupts the locked replay of a single context (PTSD mechanism).
\item \textbf{Vagal tone exercises} (cold exposure, slow exhalation).
Activates the parasympathetic branch, shifting the system from
threat-context lock to rest-and-digest context.
\end{itemize}

\textbf{Cost:} Zero. \textbf{Latency:} Seconds to minutes.
\textbf{Side effects:} None.

The rationale for Tier 1 as first-line: a thrashing system cannot
be optimized while it is thrashing. The first intervention must
halt the runaway process. Attempting cognitive restructuring
(Tier 2) or pharmacotherapy (Tier 3) on a system in active thrash
is analogous to optimizing a scheduler on a CPU that is currently
deadlocked. Kill the runaway process first.

\subsection{Tier 2: Optimize the scheduler}

\textbf{Objective:} Retrain the context-switching rules so
that future computation is more efficient. Increase $C_n^{\text{eff}}$
without pharmacological intervention.

\textbf{Interventions:}
\begin{itemize}
\item \textbf{Cognitive behavioral therapy (CBT).} Explicitly
rewrites the rules governing context selection. ``When I notice
thought pattern X, I switch to evaluation Y instead of loop Z.''
This is literally scheduler optimization.
\item \textbf{Mindfulness-based stress reduction (MBSR).} Trains
voluntary control over context loading and release. The practice
of ``noticing without engaging'' is training the ability to observe
a context candidate without loading it into the active window.
\item \textbf{Acceptance and commitment therapy (ACT).} Separates
context content from context switching. The content of an
anxious thought matters less than whether the system loads,
processes, and releases it efficiently.
\end{itemize}

\textbf{Cost:} Time (weeks to months). \textbf{Latency:}
Weeks. \textbf{Side effects:} Minimal.

Mindfulness practice has documented effects on telomere
maintenance \cite{Schutte2020} and inflammatory markers
\cite{Creswell2016}. In our framework, these are downstream
consequences of reduced context-switching waste: fewer unnecessary
DNA breaks, less repair demand, less telomere erosion from
replicative stress in repair-active tissues.

\subsection{Tier 3: Tune the hardware}

\textbf{Objective:} Pharmacologically modify the system parameters
when Tiers 1 and 2 are insufficient. Modulate neurotransmitter
systems to reduce loop gain (SSRIs), improve context efficiency
(stimulants), or release context locks (beta-blockers).

\textbf{Interventions and predicted cancer impact:}

\begin{table}[ht]
\centering
\caption{Pharmacotherapy and context-window interventions with
predicted cancer impact. Protective interventions improve context
efficiency (bits per cycle); non-protective interventions reduce
distress without improving efficiency or restore disabled machinery.}
\label{tab:pharmacotherapy}
\small
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Drug class} & \textbf{Context effect} &
\textbf{Predicted impact} & \textbf{Evidence} \\
\midrule
SSRIs & Reduce rumination loop gain & Protective &
Reduced CRC risk \cite{Chubak2011} \\
Stimulants (ADHD) & Improve context efficiency &
Protective & No excess risk \cite{Dalsgaard2018} \\
Beta-blockers & Release threat context lock &
Protective & Reduced metastasis \cite{Barron2011} \\
Mindfulness (drug-free) & Reduce unnecessary switches &
Protective & Telomere preservation \cite{Schutte2020} \\
\midrule
Benzodiazepines & Reduce distress but impair formation &
Not protective & Mixed \cite{Kao2017} \\
Antipsychotics & Restore context machinery &
Risk-neutral or harmful & Increased risk vs.\ unmedicated
schizophrenia \cite{Solmi2020} \\
\bottomrule
\end{tabular}
\end{table}

The critical distinction: interventions that improve context
efficiency (more bits per cycle) are predicted to be protective.
Interventions that reduce distress without improving efficiency
(benzodiazepines) are not. Interventions that restore disabled
machinery (antipsychotics) may increase risk by re-enabling the
context switches cancer requires.

\subsection{Meta-tier: Monitor and repair the escalation chain}

The three treatment tiers operate on the context window---reducing
waste, improving efficiency, tuning parameters. This tier operates
on a different problem space: the escalation chain itself. It reads
the state of the molecular hierarchy directly, detecting both
accumulating repair burden and somatic modifications that have
already altered the chain. It is simultaneously diagnostic (has the
escalation chain been damaged?) and therapeutic (can we intervene
before the damage cascades?).

\textbf{Objective:} Verify that psychological intervention is
reducing somatic mutation accumulation at the molecular level.

Circulating free DNA (cfDNA) from liquid biopsy can detect somatic
mutations in chromatin remodeling genes (\emph{ARID1A},
\emph{KMT2C}, \emph{SMARCA4}, \emph{DNMT3A}) without tissue
biopsy \cite{Wan2017}. If the framework is correct, effective
treatment of a context window disorder should reduce the
accumulation rate of somatic mutations in these specific genes.

The proposed biomarker panel: somatic variant allele frequency in
\emph{ARID1A}, \emph{KMT2C}, \emph{KMT2D}, \emph{SMARCA4},
\emph{DNMT3A}, \emph{TET2}, \emph{SETD2}, and \emph{CREBBP}
from serial cfDNA draws at 6-month intervals.

A positive result, declining chromatin mutation VAF in patients
whose context disorder is effectively treated, would constitute
direct evidence that psychological intervention modifies cancer
risk through a molecular mechanism, not merely through behavioral
or hormonal intermediaries.

% ------------------------------------------------------------------
\section{Falsifiable Predictions}
\label{sec:predictions}
% ------------------------------------------------------------------

\begin{enumerate}
\item \textbf{Inverted-U curve.} Across a severity spectrum of
context window disorders, cancer risk peaks at moderate severity
and decreases at both extremes (healthy and severely impaired).

\item \textbf{Chromatin specificity.} The elevated somatic
mutation burden in moderate context disorders is concentrated
in chromatin remodeling genes, not uniformly distributed across
the genome.

\item \textbf{ADHD treatment is cancer prevention.} Stimulant
medication for ADHD, by improving context efficiency, reduces
somatic mutation accumulation rate in chromatin remodeling genes.
Testable via cfDNA in treated vs.\ untreated ADHD cohorts.

\item \textbf{Depression treatment slows chromatin mutation.}
Effective antidepressant treatment (defined as remission of
rumination) reduces the rate of somatic mutation in chromatin
remodeling genes. Ineffective treatment (symptom persistence)
does not.

\item \textbf{Antipsychotics restore cancer risk.} Antipsychotic
treatment in schizophrenia increases cancer risk relative to
unmedicated schizophrenia, specifically through restoration of
chromatin remodeling function. The effect should correlate with
degree of cognitive improvement.

\item \textbf{Meditation is measurably protective.} Long-term
meditation practitioners show lower somatic mutation burden in
chromatin remodeling genes than age-matched controls, after
controlling for lifestyle factors.

\item \textbf{Benzodiazepines are not protective.} Despite
reducing anxiety symptoms, benzodiazepines do not reduce somatic
mutation accumulation because they impair context formation
without improving context efficiency.

\item \textbf{Non-carcinogen addictions carry cancer risk.} Behavioral
addictions (gambling, internet, gaming) without chemical carcinogen
exposure show elevated somatic mutation burden through the context
lock mechanism alone.

\item \textbf{Chromatin mutations are increasing disproportionately.}
The fraction of tumors with chromatin remodeling gene mutations is
increasing faster than mutations in other gene classes over the last
two decades, independent of improved detection.

\item \textbf{CHIP onset is shifting younger.} Clonal hematopoiesis
driven by \emph{DNMT3A}, \emph{TET2}, and \emph{ASXL1} is detectable
at younger ages in post-smartphone cohorts than in age-matched
historical controls.

\item \textbf{Early-onset cancers are enriched for chromatin
mutations.} The cancers driving the early-onset trend (colorectal,
breast, pancreatic in adults under 50) show higher rates of chromatin
remodeling mutations than late-onset counterparts, after controlling
for stage and subtype.

\item \textbf{Information exposure predicts chromatin VAF.}
Context-switching load (measurable via smartphone usage data,
notification frequency, or occupational classification) correlates
with somatic mutation burden in chromatin remodeling genes from
cfDNA, after controlling for age and carcinogen exposure.
\end{enumerate}

% ------------------------------------------------------------------
\section{Relationship to Prior Work}
\label{sec:prior}
% ------------------------------------------------------------------

\textbf{Psychoneuroimmunology.} The field of
psychoneuroimmunology has long studied how psychological states
affect immune function and cancer outcomes \cite{Kiecolt-Glaser2002}.
Our contribution is mechanistic specificity: not ``stress causes
cancer through immune suppression'' but ``context-switching waste
causes somatic mutation in the specific genes that control chromatin
accessibility, through the specific mechanism of topoisomerase-mediated
DNA breaks.''

\textbf{Activity-dependent DNA damage.} Madabhushi et al.\
\cite{Madabhushi2015} demonstrated that neuronal activity causes
targeted DNA double-strand breaks. Subsequent work showed that
these breaks accumulate with age and correlate with
neurodegenerative disease \cite{Thadathil2021}. We extend this to
cancer risk: if the break rate depends on computational efficiency,
then disorders that reduce efficiency increase the break-to-knowledge
ratio.

\textbf{The Waddington connection.} Waddington's genetic
assimilation \cite{Waddington1942} showed that environmentally
induced phenotypes can become genetically encoded. Our framework
provides a mechanism: the epigenetic changes induced by experience
(through chromatin remodeling during learning) are not merely
reversible marks. They are accompanied by somatic mutations from
imperfect repair of activity-dependent breaks. The mutations are
permanent. Culture does not merely select on existing genetic
variation. It generates new somatic variation through the physical
cost of knowledge acquisition.

\textbf{Coupling channels and cancer.} In the companion paper
\cite{McCarthy2026e}, we showed that the number of coupling
channels disrupted predicts cancer survival across 73,593 patients.
The six coupling channels include chromatin remodeling
(ChromatinRemodel) and DNA methylation (DNAMethylation) as
the two channels added in the expanded eight-channel system.
These are precisely the channels that the present paper identifies
as the shared molecular substrate between context disorders and
cancer.

\textbf{Bounded context and intelligence.} In \cite{McCarthy2026d},
we derived the structural invariants of intelligence from bounded
context, showing that evaluation-driven descent and encoding
permanence are necessary properties of any system that reduces
uncertainty under bounded resources. The present paper adds a
biological cost to that descent: each step of evaluation-driven
descent, each cycle of the intelligence loop, has a DNA damage
cost. Intelligence is not free. Its cost is somatic mutation.

% ------------------------------------------------------------------
\section{The Crossover: When Maintenance Exceeds Acquisition}
\label{sec:crossover}
% ------------------------------------------------------------------

Knowledge acquisition has diminishing returns. The marginal bit
costs more as the knowledge base grows; the effective cost of
the $n$-th proposition scales as $O(\log n)$ due to interference,
retrieval complexity, and the need to integrate with existing
knowledge.

Maintenance cost grows faster. Keeping $n$ propositions coherent
requires that each proposition remain accessible for potential
interaction with others. The DNA damage cost of context
maintenance, the ongoing chromatin remodeling required to keep
gene expression programs responsive, scales with the number of
context switches the system must service. In the worst case this
is $O(n^2)$: any proposition might need to interact with any
other.

These curves cross at some critical knowledge load $n^*$.

\begin{definition}[Context maintenance crossover]
The crossover point $n^*$ is the knowledge load at which the
DNA damage cost of maintaining coherent access to $n$ propositions
equals the DNA damage cost of acquiring the next proposition:
\begin{equation}
\delta \cdot c_{\text{maintain}}(n^*) = \delta \cdot c_{\text{acquire}}(n^*)
\end{equation}
where $c_{\text{maintain}}(n) \sim n^2$ and
$c_{\text{acquire}}(n) \sim \log n$.
\end{definition}

Before $n^*$, learning is cheap relative to maintenance. After
$n^*$, every new bit costs more in maintenance-driven DNA damage
than it contributes in uncertainty reduction. The system must
begin discarding propositions to stay below $n^*$, or the
repair machinery will be overwhelmed.

\subsection{The brain as first-line change management}

The brain's primary strategy for staying below $n^*$ is
aggressive compression:

\begin{enumerate}
\item \textbf{Sleep consolidation.} Episodic memories (high $n$,
high maintenance cost) are compressed into semantic memories
(lower $n$, shared structure, amortized maintenance). This is
the graph factoring operation from \cite{McCarthy2026a}:
extracting shared structure to reduce the proposition count
without losing information.
\item \textbf{Strategic forgetting.} Propositions that have not
been accessed recently are evicted. This is the Belady
replacement policy operating on the biological context window:
evict the proposition whose next use is farthest in the future.
\item \textbf{Chunking.} Multiple propositions are compressed
into a single proposition with internal structure (a ``chunk'').
This reduces $n$ while preserving retrievability. Expert
performance is characterized by larger chunks: a chess master
stores a board position as one proposition where a novice
stores 32.
\item \textbf{Habit formation.} Frequently accessed
proposition chains are compiled into automatic routines that
no longer require context-window loading. This is the
substrate transition from \cite{McCarthy2026f}: when a
computation becomes deterministic, it is removed from the
uncertainty-reducing system and encoded as fixed procedure,
reducing maintenance load on the active context.
\end{enumerate}

These mechanisms keep $n$ below $n^*$ for decades. But they can
be overwhelmed:

\begin{itemize}
\item \textbf{Chronic stress} loads threat propositions that
cannot be compressed (because they are unpredictable) or evicted
(because they are flagged as survival-relevant). They persist in
active context, raising $n$ toward $n^*$.
\item \textbf{Trauma} creates propositions that cannot be
integrated (they contradict the existing knowledge structure) or
forgotten (they are tagged as maximally important). They occupy
permanent slots in the context window.
\item \textbf{Information overload} (a modern phenomenon) pushes
the acquisition rate above the consolidation rate. Propositions
accumulate in working memory faster than sleep can compress them.
The maintenance cost rises while consolidation lags.
\end{itemize}

When the brain cannot hold the line, the biological context space
absorbs the overflow. Chromatin marks accumulate beyond what the
maintenance machinery can service. DNA repair backlogs grow.
The read/write heads, the chromatin remodeling genes, wear out.

\subsection{Aging as maintenance overload}

The crossover point $n^*$ is not fixed. It depends on the
efficiency of the repair machinery, which degrades with age.
As repair fidelity declines, $n^*$ decreases: the system can
maintain coherent access to fewer propositions before the
maintenance cost overwhelms repair capacity.

This reframes biological aging as context maintenance overload.
The organism does not degrade because time passes. It degrades
because the accumulated knowledge load eventually exceeds the
capacity of the molecular maintenance machinery, and the
resulting DNA damage in chromatin remodeling genes progressively
corrupts the context-switching apparatus.

The observation that cancer incidence rises sharply with age is
consistent: once $n > n^*$ is persistent, the chromatin
remodeling machinery is chronically overloaded, somatic mutations
in context-switching genes accumulate, and the probability that
a cell loses context control (becomes cancerous) rises.

The observation that cognitive decline precedes many cancers is
also consistent: the brain crosses $n^*$ first (because it has
the highest context-switching demand), begins losing propositions,
and the systemic maintenance overload follows.

% ------------------------------------------------------------------
\section{The Information Age Prediction}
\label{sec:information-age}
% ------------------------------------------------------------------

If the mechanism proposed here is correct, then a society-level
increase in context-switching load should produce a society-level
increase in somatic mutation burden, concentrated in chromatin
remodeling genes specifically.

The last twenty years have seen an unprecedented increase in
information exposure. Smartphones, social media, and continuous
connectivity have dramatically increased the rate of context
switching in daily life. A pre-smartphone adult might execute
dozens of major context switches per day. A smartphone user
executes hundreds: checking notifications, switching between
apps, processing social media feeds, responding to messages.
Each switch loads a new cognitive context, triggers gene
expression changes in active neural circuits, and imposes the
associated DNA damage cost.

\subsection{CHIP as the canary}

Clonal hematopoiesis of indeterminate potential (CHIP) is the
age-related accumulation of somatic mutations in blood stem cells.
CHIP is detected in approximately 10\% of people over 65 and
increases with age \cite{Jaiswal2014}. It predicts elevated risk
of both hematologic malignancy and cardiovascular disease.

The three most common CHIP driver mutations are:

\begin{enumerate}
\item \emph{DNMT3A}: a DNA methyltransferase. Chromatin/epigenetic
machinery.
\item \emph{TET2}: a methylcytosine dioxygenase. Chromatin/epigenetic
machinery.
\item \emph{ASXL1}: a chromatin remodeling factor. Polycomb group
regulation.
\end{enumerate}

All three are chromatin remodeling genes. Together they account
for approximately 80\% of CHIP mutations \cite{Steensma2015}.
This is not what a uniform mutagenesis model predicts. If CHIP
were driven by random replication errors, the mutations would be
distributed across the genome in proportion to gene size and
replication timing. They are not. They are concentrated in the
genes that control chromatin accessibility.

Our framework explains this concentration. Blood stem cells
reside in the bone marrow, are innervated by sympathetic fibers,
and respond to systemic stress signaling (cortisol,
catecholamines). Chronic stress increases their proliferation
rate \cite{Heidt2014}. But proliferation alone does not explain
the chromatin specificity. Our model adds: the chromatin
remodeling machinery in these cells is under elevated demand
because systemic stress forces context switches (changes in gene
expression program) at a rate that exceeds the repair capacity for
these specific loci.

\subsection{The testable prediction}

If context-switching load is the driver, then:

\begin{enumerate}
\item \textbf{Chromatin mutations should be increasing faster than
other mutation types.} Cancer registry data with temporal
resolution (SEER, AACR GENIE, TCGA vs.\ modern panels) should
show that the fraction of tumors with chromatin remodeling
mutations is increasing over time, independent of improved
detection.

\item \textbf{The increase should be strongest in younger
populations.} Digital natives (born after 1995) have the highest
lifetime context-switching load. If the mechanism is correct,
they should show earlier onset of CHIP and elevated chromatin
mutation rates relative to age-matched historical controls.

\item \textbf{CHIP should appear at younger ages in
high-information-exposure populations.} Tech workers, social media
professionals, and other high-context-switching occupations should
show earlier CHIP onset than low-context-switching occupations,
after controlling for traditional risk factors.

\item \textbf{The increase should NOT be concentrated in known
carcinogen-signature genes.} Environmental carcinogen exposure
produces characteristic mutational signatures (e.g., C$>$A
transversions from tobacco, CC$>$TT from UV). If the mutation
increase is driven by information exposure rather than chemical
exposure, the new mutations should lack these signatures and
instead show the replication-error signatures associated with
DNA break repair.
\end{enumerate}

The last prediction is the discriminating test. Carcinogen-driven
increases produce signature-specific mutations in genes throughout
the genome. Context-switching-driven increases produce
repair-error mutations concentrated in chromatin remodeling genes.
These are distinguishable in sequencing data.

\subsection{Cancer incidence trends}

Cancer incidence in adults under 50 has been increasing globally,
a trend termed ``early-onset cancer'' \cite{Sung2024}. The
increase is not fully explained by improved detection or known
risk factors. The cancers with the largest increases include
colorectal, breast, and pancreatic, all of which have high
rates of chromatin remodeling gene mutations.

We do not claim that information overload is the sole driver.
But we predict that it is a contributor, operating through the
specific mechanism described here: elevated context-switching load
$\rightarrow$ elevated DNA break rate in chromatin remodeling
loci $\rightarrow$ earlier exhaustion of repair capacity
$\rightarrow$ earlier onset of somatic mutations in the genes
that control cellular context $\rightarrow$ earlier cancer.

This is testable. It requires longitudinal cfDNA data from
cohorts stratified by information exposure, with mutation calling
focused on chromatin remodeling genes and mutational signature
analysis. The prediction is specific: chromatin remodeling VAF
should correlate with context-switching load after controlling for
age, carcinogen exposure, and family history.

% ------------------------------------------------------------------
\section{Discussion}
\label{sec:discussion}
% ------------------------------------------------------------------

The central claim of this paper is simple. Learning breaks DNA.
Inefficient learning breaks more DNA per bit learned. Psychiatric
disorders that reduce learning efficiency increase the ratio of
DNA breaks to knowledge acquired. The genes most affected are the
chromatin remodeling genes that control context switching at the
molecular level. These are the same genes whose somatic mutation
drives cancer.

This creates a direct, testable, molecular link between the
efficiency of cognition and the risk of cancer: not through stress
hormones, not through behavioral factors, not through immune
suppression, but through the physical cost of computation in a
biological substrate.

The inverted-U prediction distinguishes this framework from
simpler ``stress causes cancer'' models. Stress models predict
monotonically increasing cancer risk with increasing psychiatric
severity. We predict the opposite at the extreme: the most
severe context failures are protective because they disable the
machinery that cancer needs.

The clinical implication is that treating context window disorders
is not merely palliative. It is potentially disease-modifying for
cancer. The three-tier protocol, interrupt, optimize, tune, is
ordered by the logic of the framework: you cannot optimize a
scheduler that is currently thrashing. Halt the runaway process
first.

The deepest implication is for the relationship between knowledge
and biology. If learning has a DNA damage cost, and if that cost
is specific to the genes that control gene expression context,
then what we know is physically inscribed in the error pattern of
our DNA repair. Not in the sequence itself, like the immune
system's somatic hypermutation, but in the \emph{wear pattern}
on the machinery that controls which sequences are readable.

If the mechanism proposed here holds, culture does not merely
select on existing genetic variation. It writes on us---not as
metaphor, but as somatic mutation in chromatin remodeling genes,
accumulated across a lifetime of context switches, at a rate
determined by how efficiently we learn. This is the speculative
endpoint of the argument, but it is falsifiable: Prediction~12
provides the test.

% ------------------------------------------------------------------
\section*{Conflict of Interest}
The author declares no competing interests.

\section*{Funding}
This research received no external funding.

% ------------------------------------------------------------------
% References
% ------------------------------------------------------------------

\begin{thebibliography}{99}

\bibitem{McCarthy2026a}
P.~D.~McCarthy, ``Graph structure is necessary for information
preservation under bounded context,'' 2026.

\bibitem{McCarthy2026d}
P.~D.~McCarthy, ``Evaluation-driven descent, encoding permanence,
and the structural invariants of intelligence,'' 2026.

\bibitem{McCarthy2026e}
P.~D.~McCarthy, ``Coupling-channel count predicts cancer survival
better than mutation count,'' 2026.

\bibitem{McCarthy2026f}
P.~D.~McCarthy, ``Determinism precipitates from uncertainty: when
agents should become scripts,'' 2026.

\bibitem{Madabhushi2015}
R.~Madabhushi et~al., ``Activity-induced {DNA} breaks govern the
expression of neuronal early-response genes,'' \emph{Cell},
vol.~161, pp.~1592--1605, 2015.

\bibitem{Lindahl1993}
T.~Lindahl, ``Instability and decay of the primary structure of
{DNA},'' \emph{Nature}, vol.~362, pp.~709--715, 1993.

\bibitem{Sun2022}
S.~Sun et~al., ``Association of attention-deficit/hyperactivity
disorder with cancer risk,'' \emph{JAMA Network Open}, vol.~5,
e2249253, 2022.

\bibitem{Chen2012tbi}
Y.-H.~Chen et~al., ``Increased risk of brain cancer in people
with traumatic brain injury,'' \emph{Br.~J.~Cancer}, 2012.

\bibitem{Gradus2015}
J.~L.~Gradus et~al., ``Posttraumatic stress disorder and cancer
risk,'' \emph{JAMA Psychiatry}, vol.~72, pp.~1179--1185, 2015.

\bibitem{Shen2013}
C.-C.~Shen et~al., ``Risk of cancer in patients with obsessive-compulsive
disorder,'' \emph{J.~Clin.~Psychiatry}, 2013.

\bibitem{Tao2017}
R.~Tao et~al., ``Association between behavioral addiction and
cancer risk,'' \emph{Psycho-Oncology}, 2017.
% TODO: verify authors/journal/volume before final submission.

\bibitem{Chida2008}
Y.~Chida et~al., ``Does psychosocial well-being predict cancer
incidence?'' \emph{Nature Rev.~Clin.~Oncol.}, vol.~5,
pp.~466--475, 2008.

\bibitem{Batty2017}
G.~D.~Batty et~al., ``Psychosocial factors and cancer incidence,''
\emph{BMJ}, 2017.

\bibitem{McGinty2012}
E.~E.~McGinty et~al., ``Cancer incidence in a sample of
{M}aryland residents with serious mental illness,''
\emph{Psychiatric Services}, vol.~63, pp.~714--717, 2012.

\bibitem{Catts2008}
V.~S.~Catts et~al., ``Cancer incidence in patients with
schizophrenia,'' \emph{Schizophrenia Research}, vol.~102,
pp.~137--162, 2008.

\bibitem{Ji2013}
J.~Ji et~al., ``Cancer risk in schizophrenia,'' \emph{BMC Cancer},
vol.~13, 2013.

\bibitem{Singh2016}
T.~Singh et~al., ``Rare schizophrenia risk variants are enriched
in chromatin remodeling genes,'' \emph{Nature Genetics}, vol.~48,
pp.~400--404, 2016.

\bibitem{Solmi2020}
M.~Solmi et~al., ``Cancer risk in antipsychotic-treated
schizophrenia,'' \emph{Lancet Psychiatry}, 2020.

\bibitem{Driver2012}
J.~A.~Driver et~al., ``Inverse association between cancer and
{A}lzheimer's disease,'' \emph{BMJ}, vol.~344, e1442, 2012.

\bibitem{Musicco2013}
M.~Musicco et~al., ``Inverse occurrence of cancer and {A}lzheimer
disease,'' \emph{Neurology}, vol.~81, pp.~322--328, 2013.

\bibitem{Behrens2009}
M.~I.~Behrens et~al., ``Inverse relationship of {A}lzheimer's
disease and cancer,'' \emph{JAMA Neurol.}, 2009.

\bibitem{Darbro2016}
B.~W.~Darbro et~al., ``Autism linked to increased oncogene
mutations but decreased cancer rate,'' \emph{PLOS ONE}, vol.~11,
e0149041, 2016.

\bibitem{Lu2017}
Y.~Lu et~al., ``Insomnia and cancer risk,'' \emph{Sleep Medicine
Reviews}, 2017.

\bibitem{Bajaj2010}
A.~Bajaj et~al., ``{P}arkinson's disease and cancer risk,''
\emph{Cancer Causes Control}, vol.~21, pp.~697--707, 2010.

\bibitem{Hasle2000}
H.~Hasle et~al., ``Cancer incidence in {D}own syndrome,''
\emph{Lancet}, vol.~355, pp.~165--169, 2000.

\bibitem{Patja2001}
K.~Patja et~al., ``Cancer incidence of persons with intellectual
disability,'' \emph{J.~Intellect.~Disabil.~Res.}, 2001.

\bibitem{Santen2012}
G.~W.~E.~Santen et~al., ``Mutations in \emph{ARID1A} cause
intellectual disability,'' \emph{Am.~J.~Hum.~Genet.}, 2012.

\bibitem{Miyake2013}
N.~Miyake et~al., ``\emph{KMT2D} and {K}abuki syndrome,''
\emph{Am.~J.~Med.~Genet.}, 2013.

\bibitem{Tsurusaki2012}
Y.~Tsurusaki et~al., ``Mutations in \emph{SMARCA4} cause
{C}offin-{S}iris syndrome,'' \emph{Nature Genetics}, 2012.

\bibitem{Tatton-Brown2014}
K.~Tatton-Brown et~al., ``Mutations in \emph{DNMT3A} cause
{T}atton-{B}rown-{R}ahman syndrome,'' \emph{Am.~J.~Hum.~Genet.},
2014.

\bibitem{Kandoth2013}
C.~Kandoth et~al., ``Mutational landscape and significance across
12 major cancer types,'' \emph{Nature}, vol.~502, pp.~333--339,
2013.

\bibitem{Wan2017}
J.~C.~Wan et~al., ``Liquid biopsies come of age,'' \emph{Nature
Reviews Cancer}, vol.~17, pp.~223--238, 2017.

\bibitem{Kiecolt-Glaser2002}
J.~K.~Kiecolt-Glaser et~al., ``Psychoneuroimmunology and cancer,''
\emph{Ann.~Oncol.}, 2002.

\bibitem{Thadathil2021}
N.~Thadathil et~al., ``{DNA} double-strand breaks accumulate in
aging brain,'' \emph{Frontiers in Aging Neuroscience}, 2021.

\bibitem{Waddington1942}
C.~H.~Waddington, ``Canalization of development and the
inheritance of acquired characters,'' \emph{Nature}, vol.~150,
pp.~563--565, 1942.

\bibitem{Church2012}
D.~Church et~al., ``The effect of emotional freedom techniques on
stress biochemistry,'' \emph{J.~Nerv.~Ment.~Dis.}, vol.~200,
pp.~891--896, 2012.

\bibitem{Schutte2020}
N.~S.~Schutte and J.~M.~Malouff, ``A meta-analytic review of the
effects of mindfulness meditation on telomerase activity,''
\emph{Psychoneuroendocrinology}, 2020.

\bibitem{Creswell2016}
J.~D.~Creswell et~al., ``Alterations in resting-state functional
connectivity link mindfulness meditation with reduced
interleukin-6,'' \emph{Biol.~Psychiatry}, 2016.

\bibitem{Chubak2011}
J.~Chubak et~al., ``Antidepressant use and colorectal cancer
risk,'' \emph{Cancer Causes Control}, 2011.

\bibitem{Dalsgaard2018}
S.~Dalsgaard et~al., ``{ADHD} medication and cancer risk,''
\emph{J.~Child Psychol.~Psychiatry}, 2018.

\bibitem{Barron2011}
T.~I.~Barron et~al., ``Beta-blockers and breast cancer mortality,''
\emph{J.~Clin.~Oncol.}, vol.~29, pp.~2635--2644, 2011.

\bibitem{Kao2017}
C.-H.~Kao et~al., ``Benzodiazepine use and risk of cancer:
a population-based cohort study,'' \emph{Psycho-Oncology},
2017.
% TODO: verify volume/pages before final submission.

\bibitem{Franks2012}
A.~L.~Franks and J.~E.~Slansky, ``Autoimmune disease and cancer,''
\emph{Autoimmunity}, 2012.

\bibitem{Jaiswal2014}
S.~Jaiswal et~al., ``Age-related clonal hematopoiesis associated
with adverse outcomes,'' \emph{N.~Engl.~J.~Med.}, vol.~371,
pp.~2488--2498, 2014.

\bibitem{Steensma2015}
D.~P.~Steensma et~al., ``Clonal hematopoiesis of indeterminate
potential and its distinction from myelodysplastic syndromes,''
\emph{Blood}, vol.~126, pp.~9--16, 2015.

\bibitem{Heidt2014}
T.~Heidt et~al., ``Chronic variable stress activates hematopoietic
stem cells,'' \emph{Nature Medicine}, vol.~20, pp.~754--758, 2014.

\bibitem{Sung2024}
H.~Sung et~al., ``Global trends in incidence of early-onset
cancer,'' \emph{BMJ Oncology}, 2024.

\end{thebibliography}

\end{document}
