\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}
\hypersetup{hidelinks}
\usepackage{booktabs}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}{Definition}

\setlength{\parskip}{4pt plus 1pt minus 1pt}
\setlength{\parindent}{0pt}

<<<<<<< HEAD
\title{Escalation Dominates Top-Down Allocation\\Under Bounded Context}
=======
\title{Convergence of Control Architectures\\to Escalation Under Bounded Context}
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
\author{Patrick D.\ McCarthy}
\date{}

\begin{document}
\maketitle

\begin{abstract}
<<<<<<< HEAD
Top-down allocation --- where a supervisory process directs
subproblems to lower levels --- is viable when the action repertoire
$|A|$ fits within bounded active context $C_n$. We prove that once
$|A|$ exceeds a crossover point $|A|^* = C_n / 2\alpha$, top-down
allocation degrades inference quality monotonically, whether the
architecture is flat ($O(|A|)$ overhead) or hierarchical ($O(b \cdot
\log_b |A|)$ overhead for branching factor $b$). The crossover is
permanent: recovery requires growing $C_n$ with $|A|$, which no
bounded system can satisfy. Escalation --- bottom-up failure signaling
where higher levels engage only when lower levels fail --- achieves
$O(1)$ steady-state context cost and \emph{improves} with capability
growth, as successfully encoded actions relieve rather than burden
the highest level. Since intelligence requires $|A| \gg C_n$, every
intelligent system operates beyond the crossover. Escalation dominance
is not a universal claim but a scaling result: top-down wins early,
escalation wins at every scale that matters. This is a formal critique
of precision allocation models including the Free Energy Principle
=======
Top-down allocation---where a supervisory process directs subproblems
to lower levels---is viable when the concurrent delegation count $k$
fits within bounded active context $C_n$. We prove that as the action repertoire $|A|$ grows under bounded
$C_n$, concurrent supervisory demands $k$ grow because total domain
knowledge $|K|$ grows with $|A|$, exposing more dependency chains to
evidence revision. Any control architecture that remains viable must
progressively shed supervisory tracking of routine actions, offloading
them to autonomous lower levels that signal upward only on failure. This shedding \emph{is} escalation.
This monotone increase in the escalation fraction is not a design
choice but a scaling consequence:
maintaining top-down tracking incurs $O(k \cdot \alpha)$ context
overhead, which fills $C_n$ as capability grows, degrading inference
quality for novel problems. The crossover occurs when $k \cdot \alpha
\geq C_n / 2$---for biological working memory, at 2--3 concurrent
delegations. Beyond the crossover, the only recovery is to stop
tracking routine actions, which is to adopt escalation for those
actions. Hierarchical decomposition delays the crossover (each level
sees only $O(b)$ subordinates) but does not eliminate it: the number
of management levels grows with $|A|$, consuming encoding capacity.
As hierarchical top-down becomes more autonomous at lower levels, it
converges toward escalation. The pressure is one-directional:
escalation never converges toward top-down, because it never
introduces anticipatory delegation context. This result refines
precision allocation models including the Free Energy Principle
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
\cite{Friston2010,Friston2017} and the System~M routing architecture
of Dupoux, LeCun, and Malik \cite{DupouxLeCunMalik2026}.
\end{abstract}

% ------------------------------------------------------------------
\section{Introduction}
\label{sec:intro}
% ------------------------------------------------------------------

Every system that acts under uncertainty must solve a control problem:
which problems receive active reasoning, and which are handled by
automatic, pre-encoded responses? This question is distinct from the
content of reasoning; it concerns the \emph{architecture of
<<<<<<< HEAD
attention}. Under bounded active context $C_n$ --- a hard limit on the
number of propositions simultaneously available for inference --- the
control architecture is not a free parameter but a survival-critical
design choice.

Two competing architectures present themselves:
=======
attention}. Under bounded active context $C_n$---a hard limit on the
number of propositions simultaneously available for inference, the
formal analogue of Simon's bounded rationality
\cite{Simon1955,Simon1962}---the control architecture is not a free
parameter but a survival-critical design choice.

Two complementary mechanisms address this problem:
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025

\begin{enumerate}
\item \textbf{Top-down allocation.} A supervisory process at the
   highest encoding level $L_n$ monitors all subsystems, evaluates
   relevance, and routes problems to appropriate levels. This
   architecture is implicit in precision weighting under the Free
   Energy Principle \cite{Friston2010,Friston2017}, where higher
   cortical levels set the gain on prediction errors at lower levels.
   It is explicit in the System~M proposal of Dupoux, LeCun, and Malik
   \cite{DupouxLeCunMalik2026}, where a meta-controller directs when
   Systems~A (observation) and B (action) engage. Kahneman's popular
   account of System~1 and System~2 \cite{Kahneman2011} implicitly
   assumes top-down control: System~2 ``decides'' when to override
   System~1.

\item \textbf{Escalation.} Lower levels handle problems autonomously.
<<<<<<< HEAD
   Higher levels engage only when lower levels fail --- that is, when
=======
   Higher levels engage only when lower levels fail, that is, when
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
   uncertainty cannot be bounded at the current level. This
   architecture is implicit in Global Workspace Theory
   \cite{Baars1988,Dehaene2011}, where local processors broadcast to
   the global workspace only when local resolution fails.
\end{enumerate}

<<<<<<< HEAD
At small scale --- when $|A| \leq C_n$, so the entire action repertoire
fits in active context --- top-down allocation is viable and may even be
preferable: the supervisor can hold complete state with room to spare,
and the benefit of optimal routing outweighs the modest overhead.
Intelligence, however, requires $|A| \gg C_n$: an entity must know far
more than it can hold in mind at once. This paper proves that once
$|A|$ exceeds $C_n$, top-down allocation degrades and escalation
dominates --- and that the crossover is permanent. The result is not
that top-down is wrong but that it cannot scale, and scaling is what
intelligence demands.

The proof proceeds as follows. Section~2 introduces the encoding
hierarchy $E(c,r)$ and the formal definitions required for the result.
Section~3 establishes the escalation principle (Theorem~1): informed
actions are handled at the lowest sufficient level. Section~4 proves
the central result (Theorem~2): top-down allocation degrades inference
quality as capability grows, with recovery requiring $C_n$ to grow
linearly with capability --- a condition no bounded system can satisfy.
Section~5 compares the two architectures directly. Section~6 engages
with prior work.
=======
At small scale, when the entire action repertoire fits in active
context, top-down allocation is viable and may be preferable: the
supervisor can hold complete state with room to spare, and the benefit
of optimal routing outweighs the modest overhead. Intelligence,
however, requires $|A| \gg C_n$: an entity must know far more than it
can hold in mind at once. This paper proves that as $|A|$ grows under
bounded $C_n$, any viable control architecture converges toward
escalation properties. The convergence is not a design preference but
a scaling consequence: maintaining top-down tracking incurs context
overhead that grows with capability, eventually filling $C_n$ and
degrading the very inference the supervisor is meant to optimize.
As in \cite{McCarthy2026a}, ``necessary'' here means necessary for
persistence under competition---selective necessity, not logical
impossibility of alternatives. The
only recovery is to stop tracking routine actions---which is to adopt
escalation for those actions.

The proof proceeds as follows. Section~2 introduces the encoding
hierarchy $E(c,r)$ and the formal definitions required for the result.
Section~3 defines the escalation policy: informed
actions are handled at the lowest sufficient level. Section~4 proves
the central result (Theorem~2): top-down allocation incurs $O(k
\cdot \alpha)$ context cost, and any mechanism that keeps top-down
viable at scale necessarily converges toward escalation. Section~5
compares the two architectures directly. Section~6 engages with prior
work.
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025

This paper is the second in a series. Paper~1 \cite{McCarthy2026a}
established that any system accumulating knowledge under bounded
context necessarily maintains graph structure. The present paper
<<<<<<< HEAD
addresses how that structure is \emph{controlled} --- how problems are
=======
addresses how that structure is \emph{controlled}: how problems are
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
routed through the encoding hierarchy.

% ------------------------------------------------------------------
\section{Formal Framework}
\label{sec:framework}
% ------------------------------------------------------------------

We introduce the minimal definitions required for the main results.
<<<<<<< HEAD
We inherit the world state space $W$, proposition, knowledge base $K$,
=======
We inherit the world state space $W$, the notion of proposition, knowledge base $K$,
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
uncertainty $U(w,K) = -\log P(w \mid K)$ \cite{Cover2006}, and bounded
active context $C_n$ from \cite{McCarthy2026a}. An \emph{informed
action} $a \in A$ is an action grounded in propositions from $K$; we
write $K^{[a]} \subseteq K$ for the set of propositions that ground $a$.

\begin{definition}[Encoding Hierarchy]
\label{def:hierarchy}
An \emph{encoding hierarchy} is a continuous two-dimensional space
$E(c,r)$ parameterized by:
\begin{itemize}
\item \textbf{Runtime cost} $c \geq 0$: the computational cost of
   accessing and modifying the encoding.
\item \textbf{Reversibility} $r \in [0,1]$: the ease with which the
   encoding can be revised. $r = 1$ is fully reversible; $r = 0$ is
   permanent.
\end{itemize}

Boundaries within this continuous space are \emph{interfaces}
$L_0, L_1, \ldots, L_n$. The number of interfaces and their positions
are substrate-determined: biological systems, artificial systems, and
institutional systems partition $E(c,r)$ differently.
\end{definition}

Table~\ref{tab:hierarchy} illustrates the hierarchy with representative
interfaces. The precise number and character of levels varies by
substrate; what is invariant is the monotonic relationship between
runtime cost and reversibility.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Interface} & \textbf{Description} & \textbf{Runtime Cost} &
\textbf{Reversibility} \\
\midrule
$L_n$ & Active reasoning / working memory & $O(\text{compute})$ &
Fully reversible ($r \approx 1$) \\
$L_{n-1}$ & Learned heuristics / cached policies & $O(1)$ retrieval &
Revisable with effort \\
$\vdots$ & \emph{substrate-dependent intermediate levels} & & \\
$L_1$ & Developmental encoding / early learning & Near-zero &
Difficult to revise \\
$L_0$ & Heritable / constitutional encoding & $0$ &
Permanent ($r \approx 0$) \\
\bottomrule
\end{tabular}
\caption{Representative encoding hierarchy. Cost increases and
reversibility increases from $L_0$ to $L_n$. The number of interfaces
is substrate-determined.}
\label{tab:hierarchy}
\end{table}

\begin{definition}[Active Context Capacity and Relevance Density]
\label{def:context}
The \emph{active context capacity} $C_n$ is bounded: $|\mathrm{ctx}(L_n,
t)| \leq C_n$. For an informed action $a$, the \emph{relevance
density} at $L_n$ is:
\[
<<<<<<< HEAD
\rho(L_n, f) = \frac{|K^{[a]}|}{|\mathrm{ctx}(L_n, t)|}
=======
\rho(L_n, a) = \frac{|K^{[a]}|}{|\mathrm{ctx}(L_n, t)|}
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
\]
where $K^{[a]} \subseteq K$ is the knowledge relevant to $a$.
$\rho = 1$ is maximum signal: active context holds exactly what $a$
requires. $\rho \to 0$ is noise-dominated: context is filled with
content irrelevant to $a$.
\end{definition}

\begin{definition}[Delegation Overhead]
\label{def:delegation}
Under \emph{top-down allocation}, $L_n$ actively manages lower-level
processes. For each active delegation, $L_n$ must hold in context:
\begin{enumerate}
\item[(i)] The \emph{subproblem specification}: what was delegated.
\item[(ii)] The \emph{expected return signal}: what a successful
   result looks like.
\item[(iii)] The \emph{integration context}: how the result will be
   incorporated into the current reasoning state.
\end{enumerate}
Let $\alpha > 0$ be the \emph{minimum context cost per delegation}:
the irreducible number of context slots consumed by maintaining one
active delegation. $\alpha > 0$ because each delegation requires at
least one proposition specifying what was delegated.
\end{definition}

% ------------------------------------------------------------------
\section{Escalation}
\label{sec:escalation}
% ------------------------------------------------------------------

<<<<<<< HEAD
\begin{theorem}[Escalation]
\label{thm:escalation}
Under the encoding hierarchy (Definition~\ref{def:hierarchy}), the
cost-minimizing control policy handles each action $a$ at the lowest
level $L_i$ such that the knowledge encoded at $L_i$ is sufficient to
bound uncertainty for $a$. Escalation to $L_{i+1}$ occurs if and only
if $L_i$ fails to bound uncertainty.
\end{theorem}

\begin{proof}
By Definition~\ref{def:hierarchy}, higher levels have strictly greater
runtime cost: $c(L_{i+1}) > c(L_i)$. This monotonicity is not
incidental but reflects the architecture of complexity: Simon
\cite{Simon1962} shows that stable complex systems are
\emph{nearly decomposable} --- they process information locally, with
infrequent communication between subsystems. The encoding hierarchy
formalizes this structure: lower levels are cheaper precisely because
they are more specialized and less revisable. We show that handling $a$
at any level higher than the lowest sufficient level is strictly
dominated.

Let $L_i$ be the lowest level such that $U(w, K^{[a]}_{L_i}) \leq
\epsilon$ for the action-relevant threshold $\epsilon$. Consider the
alternative of handling $a$ at $L_j$ ($j > i$):

\begin{enumerate}
\item \textbf{Cost.} $c(L_j) > c(L_i)$ by the monotonicity of cost in
   the hierarchy. The uncertainty reduction is the same (both levels
   achieve $U \leq \epsilon$), so the excess cost $c(L_j) - c(L_i) > 0$
   yields zero marginal benefit. The higher-level policy is strictly
   dominated on cost.

\item \textbf{Context.} If $j = n$ (the active reasoning level), then
   handling $a$ at $L_n$ consumes active context capacity $C_n$. Since
   $C_n$ is bounded and $L_n$ is the only level that handles genuinely
   novel problems (problems no lower level can resolve), each unnecessary
   use of $L_n$ reduces $\rho$ for the next novel problem. The
   higher-level policy is strictly dominated on context conservation.
\end{enumerate}

Under any cost-minimizing criterion --- whether natural selection,
engineering optimization, or resource-rational design --- the dominated
policy is eliminated. The unique cost-minimizing policy is: handle $a$
at the lowest sufficient level; escalate if and only if that level
fails.

\textbf{Remark.} This result does not claim that all systems
\emph{implement} escalation. Path dependence, architectural
constraints, and design choices can produce systems that use higher
levels than necessary. The theorem establishes that such systems pay a
strictly positive cost premium: they are not cost-optimal under bounded
context. The claim is normative (escalation is optimal) rather than
descriptive (all systems escalate).
\end{proof}
=======
\begin{definition}[Escalation Policy]
\label{def:escalation}
Under the encoding hierarchy (Definition~\ref{def:hierarchy}), the
\emph{escalation policy} handles each action $a$ at the lowest level
$L_i$ such that the knowledge encoded at $L_i$ is sufficient to bound
uncertainty for $a$. Escalation to $L_{i+1}$ occurs if and only if
$L_i$ fails to bound uncertainty.
\end{definition}

\textbf{Observation.} The escalation policy is cost-minimizing under
monotone cost. Since $c(L_{i+1}) > c(L_i)$ by
Definition~\ref{def:hierarchy}, handling $a$ at any level higher than
the lowest sufficient level pays strictly more for the same uncertainty
reduction. If $j = n$, handling $a$ at $L_n$ also consumes bounded
active context, reducing $\rho$ for the next novel problem. The
escalation policy is strictly dominant on both cost and context
conservation. This is immediate from the monotone cost structure and
requires no proof.

The substantive question is not whether escalation is cost-minimizing
---it is, by construction---but whether bounded systems \emph{converge
toward} escalation as they scale. That is the content of
Theorem~\ref{thm:topdown}.

\textbf{Remark.} The escalation policy does not claim that all systems
\emph{implement} escalation. Path dependence, architectural
constraints, and design choices can produce systems that use higher
levels than necessary. Such systems pay a strictly positive cost
premium. The claim is normative (escalation is cost-optimal) rather
than descriptive (all systems escalate).
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025

\begin{corollary}[Context Window as Frontier]
\label{cor:frontier}
$L_n$ is the frontier of unbounded uncertainty, not the seat of
intelligence. A problem reaching $L_n$ is evidence of failure at every
lower level. An entity that routes most problems through $L_n$ is not
<<<<<<< HEAD
``thinking more'' --- it is failing to encode.
\end{corollary}

=======
``thinking more''; it is failing to encode.
\end{corollary}

\textbf{Remark (Free context).}
The quantity that escalation preserves is \emph{free context}:
$C_{\mathrm{free}}(t) = C_n -
|\mathrm{ctx}_{\mathrm{unresolved}}(t)|$, the active context capacity
available for novel problems after existing commitments. Under
top-down allocation, delegation overhead reduces
$C_{\mathrm{free}}$ by $k \cdot \alpha$
(Theorem~\ref{thm:topdown}). Under escalation, routine actions
consume zero $L_n$ capacity, so $C_{\mathrm{free}}$ is bounded below
only by the current novel problem. Escalation is the control
architecture that maximizes $C_{\mathrm{free}}$; this is the
mechanism by which capability growth \emph{relieves} rather than
burdens the highest encoding level. Escalation protects
$C_{\mathrm{free}}$: graph structure creates it by compressing
shared propositions, knowledge curation maintains it by displacing
low-value propositions, and escalation shields it from supervisory
overhead.

>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
\begin{corollary}[Simultaneous Multi-Level Activity]
\label{cor:simultaneous}
All levels of the hierarchy operate concurrently. $L_0$ handles
heritable responses, $L_1$ handles developmental encodings, and
<<<<<<< HEAD
intermediate levels handle their respective domains ---
=======
intermediate levels handle their respective domains,
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
simultaneously. $L_n$ handles only the residual: problems that no
lower level can resolve. The escalation architecture does not
serialize processing through a hierarchy; it parallelizes across levels,
with upward propagation occurring only on failure.
\end{corollary}

% ------------------------------------------------------------------
<<<<<<< HEAD
\section{Top-Down Allocation Degrades Inference Quality}
\label{sec:topdown}
% ------------------------------------------------------------------

\begin{theorem}[Top-Down Allocation Degrades Inference Quality]
=======
\section{Convergence to Escalation Under Bounded Context}
\label{sec:topdown}
% ------------------------------------------------------------------

\begin{theorem}[Convergence to Escalation at Scale]
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
\label{thm:topdown}
Under top-down allocation, the supervisory level $L_n$ must maintain
delegation state for managed subprocesses
(Definition~\ref{def:delegation}). Let $k(t)$ be the number of
concurrent delegations at time $t$. The relevance density for a novel
<<<<<<< HEAD
problem $f_{\mathrm{novel}}$ is:
\[
\rho(L_n, f_{\mathrm{novel}}) =
\frac{|K^{[f_{\mathrm{novel}}]}|}{|K^{[f_{\mathrm{novel}}]}| + k(t)
\cdot \alpha}
\]
If $k(t)$ grows with $|A|$, then $\rho \to 0$ as $|A| \to \infty$.
Recovery requires growing $C_n$ with $k(t)$ --- a condition no bounded
system can satisfy.
=======
problem $a_{\mathrm{novel}}$ is:
\[
\rho(L_n, a_{\mathrm{novel}}) =
\frac{|K^{[a_{\mathrm{novel}}]}|}{|K^{[a_{\mathrm{novel}}]}| + k(t)
\cdot \alpha}
\]
\textbf{Assumption} (concurrent demand growth): $k(t) \to \infty$ as
$|A| \to \infty$. Section~\ref{sec:crossover} justifies this from the
dependency structure of $K$. Given this assumption, as $|A|$ grows
under bounded $C_n$:
\begin{enumerate}
\item[(i)] $\rho \to 0$: inference quality degrades monotonically.
\item[(ii)] Recovery requires either growing $C_n$ (impossible under
   boundedness) or reducing $k$---that is, shedding supervisory
   tracking of actions that lower levels can handle autonomously.
\item[(iii)] The shed actions are, by definition, operating under
   escalation: handled at the lowest sufficient level, with upward
   signaling only on failure.
\end{enumerate}
Let $\phi(t) = |\{a \in A(t) : a \text{ operates under escalation}\}|
/ |A(t)|$ be the fraction of the action repertoire under escalation.
Since at most $k^* = C_n / (2\alpha)$ actions can be under active
top-down supervision simultaneously (Section~\ref{sec:crossover}),
\[
\phi(t) \;\geq\; \frac{|A(t)| - k^*}{|A(t)|} \;=\; 1 - \frac{k^*}{|A(t)|}
\]
As $|A| \to \infty$ under bounded $C_n$, $\phi \to 1$. The bound
$k^*$ is fixed by substrate ($C_n$ and $\alpha$ are constants), so
the escalation fraction approaches unity not because every action is
shed but because the number of actions that \emph{can} remain under
top-down supervision is bounded while the total repertoire grows
without bound.
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
\end{theorem}

\begin{proof}
Under top-down allocation, $L_n$ serves two simultaneous functions:
<<<<<<< HEAD
(a) reasoning about the novel problem $f_{\mathrm{novel}}$, requiring
$|K^{[f_{\mathrm{novel}}]}|$ context slots, and (b) maintaining
=======
(a) reasoning about the novel problem $a_{\mathrm{novel}}$, requiring
$|K^{[a_{\mathrm{novel}}]}|$ context slots, and (b) maintaining
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
delegation state for $k(t)$ concurrently managed subprocesses, each
consuming $\alpha$ context slots (Definition~\ref{def:delegation}).

The total context consumed is:
\[
<<<<<<< HEAD
|\mathrm{ctx}(L_n, t)| = |K^{[f_{\mathrm{novel}}]}| + k(t) \cdot \alpha
\]

Since $\alpha > 0$, any growth in $k(t)$ increases the denominator
while the numerator $|K^{[f_{\mathrm{novel}}]}|$ is bounded per
problem. Therefore $\rho$ decreases monotonically with $k(t)$.

The critical question is whether $k(t)$ grows with $|A|$. We show that
it must under any architecture that merits the name ``top-down'':

\textbf{Why $k$ grows with $|A|$.} Top-down allocation is defined by
the supervisory level \emph{initiating} and \emph{tracking}
delegations (Definition~\ref{def:delegation}). A supervisor that
delegates but does not track (fire-and-forget) holds no return-integration context and
cannot evaluate, redirect, or integrate results --- it is not
supervising. A supervisor that tracks only a fixed subset of $|A|$
while ignoring the rest has abandoned top-down control of the
ignored actions, which are then self-managing (i.e., operating under
escalation). In either case, the system is no longer top-down for the
untracked actions.

For the portion of $A$ that remains under genuine top-down
supervision, $k(t)$ grows with the number of supervised actions. Under
flat top-down, $k = |A_{\text{supervised}}|$. Under hierarchical
top-down with branching factor $b$, each supervisory level tracks $b$
subordinates across $\lceil \log_b |A_{\text{supervised}}| \rceil$
levels, giving $k = b \cdot \log_b |A_{\text{supervised}}|$. In both
cases $k$ grows with the supervised repertoire.

The critical threshold occurs when $k(t) \cdot \alpha \geq C_n -
|K^{[f_{\mathrm{novel}}]}|$: delegation overhead has consumed all
context capacity not occupied by the current problem's knowledge.

To maintain $\rho$ at a fixed level as $|A|$ grows, $C_n$ must grow
at least linearly with $k(t)$. But $C_n$
is bounded by substrate (Definition~\ref{def:context}). Therefore
top-down allocation necessarily degrades inference quality as the
supervised repertoire grows under bounded context.
\end{proof}

\subsection{The crossover: when top-down stops working}
\label{sec:crossover}

Top-down allocation is viable when delegation overhead fits comfortably
within $C_n$. Define the \emph{crossover point} as the supervised
repertoire size at which delegation overhead consumes half of active
context:
\[
k(|A|^*) \cdot \alpha = \frac{C_n}{2}
\]
For flat top-down ($k = |A|$), this gives $|A|^* = C_n / 2\alpha$. For
hierarchical top-down ($k = b \cdot \log_b |A|$), the crossover is
$b \cdot \log_b |A|^* = C_n / 2\alpha$, which gives a larger $|A|^*$
--- hierarchy delays the crossover but does not eliminate it, because
$\log_b |A|$ still grows without bound.

Below $|A|^*$, the supervisor retains at least $C_n/2$ slots for
reasoning --- adequate for most problems. Above $|A|^*$, delegation
overhead dominates: every new supervised capability \emph{reduces} the
system's ability to reason about novel problems.

\textbf{Estimating the crossover.} The minimum delegation cost $\alpha$
is at least 1: specifying ``what was delegated'' requires at least one
proposition in context. A tighter estimate: each delegation requires a
specification, an expected return, and an integration slot
(Definition~\ref{def:delegation}), giving $\alpha \geq 3$. Under
Cowan's \cite{Cowan2001} revised estimate of working memory capacity
($C_n \approx 4$, not Miller's $7 \pm 2$), the flat crossover is
$|A|^* = C_n / 2\alpha \approx 4/6 < 1$. Even with Miller's more
generous bound ($C_n \approx 7$), $|A|^* \approx 7/6 \approx 1$.
The crossover for biological working memory occurs at \emph{one
delegation} --- a supervisor that manages even a single subprocess has
already consumed most of its reasoning capacity. This is why human
``multitasking'' degrades performance: the delegation overhead of
tracking two tasks consumes the context needed to reason about
either.

This crossover is permanent under any top-down architecture. Once
$k(|A|) \cdot \alpha > C_n / 2$, the only recovery is to grow $C_n$
--- which is bounded by substrate --- or to abandon top-down
supervision for part of $A$, which is to adopt escalation for those
actions.

The crossover clarifies the scope of the result. Top-down allocation
is not wrong --- it is a \emph{small-scale} architecture. For an entity
whose entire repertoire fits in active context, top-down supervision is
efficient and may be optimal. But intelligence requires $|A| \gg C_n$:
an entity that can only act on what it can simultaneously hold in mind
is not intelligent but reactive. Every physical system that we
recognize as intelligent --- from insects routing foraging with stigmergic
signals to humans operating with $\sim$7 active context slots across
thousands of learned skills --- operates well beyond the crossover.
=======
|\mathrm{ctx}(L_n, t)| = |K^{[a_{\mathrm{novel}}]}| + k(t) \cdot \alpha
\]

Since $\alpha > 0$, any growth in $k(t)$ increases the denominator
while the numerator $|K^{[a_{\mathrm{novel}}]}|$ is bounded per
problem. This establishes Part~(i): $\rho$ decreases monotonically
with $k(t)$.

\textbf{Part (ii): the dilemma at scale.} By premise, $k(t)$ grows
with $|A|$. Under flat top-down, the supervisor tracks all active
processes directly. Under hierarchical top-down with branching factor
$b$, each supervisory level tracks $b$ subordinates across
$\lceil \log_b |A| \rceil$ levels. In both architectures, the
supervisory load at $L_n$ grows with the repertoire.

The critical threshold occurs when $k(t) \cdot \alpha \geq C_n -
|K^{[a_{\mathrm{novel}}]}|$: delegation overhead has consumed all
context capacity not occupied by the current problem's knowledge. To
maintain $\rho$ at a fixed level, $C_n$ must grow at least linearly
with $k(t)$. But $C_n$ is bounded by substrate. The only alternative
is to reduce $k$---to stop tracking actions that lower levels can
handle without supervision.

\textbf{Part (iii): shedding tracking is escalation.} A supervisor
that delegates but does not track (fire-and-forget) holds no
return-integration context and cannot evaluate, redirect, or integrate
results; it is not supervising those actions. Those actions are
self-managing: handled at the lowest sufficient level, with upward
signaling only on failure. This is precisely the escalation policy
(Definition~\ref{def:escalation}).

The convergence follows: as $|A|$ grows, a viable system must shed
top-down tracking of an increasing fraction of its repertoire. The
shed fraction operates under escalation. The supervisor retains
top-down control only for the residual that no lower level can
handle---genuinely novel problems. The system has not abandoned
top-down allocation; it has converged to a hybrid where escalation
handles the routine and top-down handles the novel. This hybrid is
the only architecture that remains viable under bounded $C_n$ as
$|A| \to \infty$.
\end{proof}

\subsection{The crossover: when top-down yields to escalation}
\label{sec:crossover}

Top-down allocation is viable when delegation overhead fits comfortably
within $C_n$. The crossover is defined in terms of \emph{concurrent
delegations} $k$, not total repertoire $|A|$. An entity with
10{,}000 learned actions does not supervise all 10{,}000
simultaneously; $k$ is the number of subprocesses the supervisor is
actively tracking at time $t$. The crossover occurs when:
\[
k^* \cdot \alpha = \frac{C_n}{2}
\quad\Longrightarrow\quad
k^* = \frac{C_n}{2\alpha}
\]

Below $k^*$, the supervisor retains at least $C_n/2$ slots for
reasoning about novel problems. Above $k^*$, delegation overhead
consumes the majority of $C_n$. The specific fraction $C_n/2$ is a
convention, not a derived quantity; the qualitative result---a finite
crossover exists---holds for any positive fraction. The quantitative
estimate ($k^* = 2$--$3$ for humans) is sensitive to this choice.

\textbf{Estimating the crossover.} The minimum delegation cost $\alpha$
is at least 1: specifying ``what was delegated'' requires at least one
proposition in context. Under Cowan's \cite{Cowan2001} revised estimate
of working memory capacity ($C_n \approx 4$), the crossover is
$k^* = 4/2 = 2$ concurrent delegations. Under Miller's more generous
bound ($C_n \approx 7$), $k^* = 3.5$. A human supervisor can
actively track 2--3 concurrent delegations before reasoning quality
degrades. This aligns with the empirical observation that human
``multitasking'' degrades performance sharply beyond two concurrent
tasks: the delegation overhead of tracking a third task consumes the
context needed to reason about any of them.

\textbf{Justification of the concurrent demand growth assumption.}
Theorem~\ref{thm:topdown} assumes $k(t) \to \infty$ as
$|A| \to \infty$. We justify this with a simple model of evidence
cascades.

Let $|K|$ denote the number of propositions and $\bar{d}$ the average
out-degree (number of dependents per proposition). Under positive
novelty rate $\nu > 0$, evidence events arrive at rate $\nu$, each
hitting one proposition and cascading to its dependents. Each cascade
produces $\bar{d}$ propositions requiring re-evaluation. If the mean
resolution time per proposition is $\tau$, the expected number of
concurrently unresolved items is:
\[
\mathbb{E}[k] \;=\; \nu \cdot \bar{d} \cdot \tau
\]
A supervisor could batch multiple unresolved items into a single
delegation (``handle all re-evaluations in domain $X$''). Batching
reduces the number of tracked delegations but does not eliminate the
overhead: specifying each batch still costs at least $\alpha$ per
batch, and verifying that a batch was handled correctly requires
loading its scope into context. Batching changes the constant in
$k \cdot \alpha$ but not the scaling behavior as $|A| \to \infty$.

The key: $\bar{d}$ grows with $|K|$. As $|A|$ grows, overlapping
action policies share more propositions, and those shared propositions
are extracted into $K$ (the normalization that any bounded system
converges toward, since maintaining redundant copies exceeds $C_n$ as
the repertoire scales). Paper~1 \cite{McCarthy2026a} establishes that
edge density increases as $K$ matures (Corollary~1): $\bar{d}$ is
monotonically non-decreasing in $|K|$. Since $|K|$ grows with $|A|$
and $\bar{d}$ grows with $|K|$, we have:
\[
\mathbb{E}[k] \;=\; \nu \cdot \bar{d}(|K|) \cdot \tau
\;\xrightarrow{|A| \to \infty}\; \infty
\]
provided $\nu > 0$ and $\tau > 0$ (resolution is not instantaneous).
The growth is driven by edge densification: a richer knowledge base has
a larger surface area for evidence to hit, producing more concurrent
unresolved items per unit time.

Under pure top-down, all concurrent unresolved items would require
supervisory tracking, producing $k \gg k^*$. The only way to keep
$k \leq k^*$ is to ensure that most concurrent demands are handled
autonomously at lower levels---that is, under escalation. The larger
$|A|$ grows, the larger the fraction that must operate under
escalation to keep $k$ bounded. This is the convergence of
Theorem~\ref{thm:topdown}(iii): escalation is not adopted as a
design choice but forced by the scaling arithmetic.

The crossover clarifies the scope of convergence. Top-down allocation
is not wrong; it is viable at small scale, when concurrent demands fit
within $C_n$. But intelligence requires $|A| \gg C_n$: an entity that
can only act on what it can simultaneously hold in mind is not
intelligent but reactive. Every physical system that we recognize as
intelligent---from insects routing foraging with stigmergic signals to
humans operating with ${\sim}7$ active context slots across thousands of
learned skills---appears to have converged, consistent with the
cost-minimization pressure of the escalation policy
(Definition~\ref{def:escalation}):
the vast majority of its repertoire operates under escalation, with
top-down engagement reserved for novel coordination.

\textbf{Time-multiplexing converges to escalation.}
An objection: the supervisor need not hold all $k$ delegations
simultaneously. It could round-robin, checking one subordinate at a
time, clearing context between checks. This eliminates the simultaneous
$k \cdot \alpha$ load. But consider what the supervisor must maintain
across checks: a \emph{delegation list}---which subordinates were
assigned which tasks---so it knows whom to check next. This list is the
delegation context; storing it costs $k \cdot \alpha$ whether held in
active context or reloaded from long-term storage (with context-switch
cost $\gamma$ per reload). As $k$ grows, the total cycle time $k \cdot
(\alpha + \gamma)$ grows, and the supervisor falls behind evidence
arrival (rate $\nu$). The crossover still exists.

The only way to eliminate the delegation list entirely is to stop
tracking what was delegated: the supervisor holds zero delegation state
and processes signals \emph{as they arrive from below}. But a
supervisor with no delegation list that processes upward signals on
arrival is not time-multiplexing top-down control---it is escalation
(Section~\ref{sec:comparison}). Time-multiplexing does not
avoid the convergence; it is a way station on the path toward it.
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025

\textbf{Cell division as illustration.} A single-celled organism can
``manage'' its repertoire top-down: the entire state fits in one
context. A multicellular organism with $10^{13}$ cells cannot manage
each cell individually through a central controller.

Biology's actual solution is a \emph{hybrid}: local autonomy (cells
monitor nutrient gradients, contact inhibition, and growth factor
concentrations) combined with systemic signaling (hormonal regulation,
tumor suppressors, immune surveillance). The systemic signals are
<<<<<<< HEAD
genuine top-down control --- but they are \emph{broadcast} signals
=======
genuine top-down control, but they are \emph{broadcast} signals
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
(growth hormone affects all cells of the right type), not individual
delegations (no central process tracks each cell's state). Broadcasting
is $O(1)$ in delegation context: the signal is sent once regardless of
the number of recipients. Individual delegation would require tracking
$10^{13}$ cells, which is well beyond any bounded context.

The hybrid architecture is precisely the crossover result: top-down
broadcast control for system-wide coordination (below the crossover for
broadcast), combined with local escalation for individual cell
decisions (above the crossover for per-cell management). When the
<<<<<<< HEAD
systemic signaling fails --- when a cell ignores growth suppression
signals --- the immune system engages: escalation on failure.
=======
systemic signaling fails, when a cell ignores growth suppression
signals, the immune system engages: escalation on failure.
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025

\subsection{The hierarchical top-down objection}

The strongest objection to Theorem~\ref{thm:topdown} is that it attacks
<<<<<<< HEAD
flat top-down allocation --- a strawman. Real supervisory systems are
=======
flat top-down allocation, a strawman. Real supervisory systems are
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
\emph{hierarchical}: a top level manages $b$ intermediaries, each
managing $b$ sub-intermediaries, so that overhead at any single level is
$O(b)$, not $O(|A|)$. Does hierarchical decomposition rescue top-down
allocation?

No. The hierarchical decomposition is more subtle than the flat case,
but it does not rescue top-down allocation. We address the strongest
version of the objection directly.

<<<<<<< HEAD
\textbf{(1) $L_n$ sees only $O(b)$ overhead --- but the hierarchy has
a different cost.}
The strongest version of the hierarchical objection notes that $L_n$
manages only $b$ direct subordinates, so overhead at $L_n$ is
$b \cdot \alpha$ --- constant in $|A|$. The $\log_b |A|$ levels of
=======
\textbf{(1) $L_n$ sees only $O(b)$ overhead, but the hierarchy has
a different cost.}
The strongest version of the hierarchical objection notes that $L_n$
manages only $b$ direct subordinates, so overhead at $L_n$ is
$b \cdot \alpha$, constant in $|A|$. The $\log_b |A|$ levels of
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
overhead are distributed across intermediate levels, each with its own
bounded context. No single level overflows. This is correct and is a
genuine advantage of hierarchical decomposition over flat top-down.

<<<<<<< HEAD
However, the cost is not eliminated --- it is \emph{redistributed}.
=======
However, the cost is not eliminated; it is \emph{redistributed}.
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
The hierarchy requires $\lceil \log_b |A| \rceil$ supervisory levels,
each of which devotes $b \cdot \alpha$ of its context to delegation
state. The number of management-dedicated levels grows with $|A|$.

\textbf{(2) Management levels displace knowledge levels.}
This is the deeper cost, and it has a precise analogue in organizational
economics. Williamson \cite{Williamson1967} proves that each management
layer in a hierarchical organization introduces ``control loss'': a
fraction $\alpha_W < 1$ of the intended directive is lost at each level,
so that $n$ management levels transmit only $\alpha_W^n$ of the
top-level intent. This limits the viable depth of hierarchical control
and, consequently, organization size. The same structure applies here:
each intermediary in a hierarchical top-down system is itself a
supervisory process that must hold delegation context, and the
cumulative cost of these management layers limits the encoding
capacity available for knowledge.

In an escalation architecture, every encoding level is a
\emph{knowledge level}: it holds proven actions and propositions, not
routing state. In hierarchical top-down, $\lceil \log_b |A| \rceil$
levels are \emph{management levels}: they hold delegation state. As
$|A|$ grows, the number of management levels grows (logarithmically),
consuming encoding capacity that escalation would use for knowledge.
Williamson's result shows this is not merely wasteful but structurally
limiting: the management overhead places an upper bound on the
effective depth of hierarchical control, just as it places an upper
bound on firm size.

The cost is not context overflow at any single level but a growing
fraction of the entity's encoding hierarchy devoted to management rather
than knowledge. An entity with 10 encoding levels and $|A|$ requiring
4 management levels has only 6 levels for knowledge; escalation would
use all 10. The management overhead is in \emph{levels consumed}, not
\emph{context slots consumed at $L_n$}.

\textbf{(3) The causal direction is the fundamental difference.}
The deepest distinction between hierarchical top-down and escalation is
not the cost asymptotic but the \emph{causal direction of engagement}:

\begin{itemize}
\item In hierarchical top-down, the upper level \emph{initiated} the
   delegation. It is waiting for a result it requested. It must hold
   return-integration context: what was delegated, what a successful
   result looks like, and how to incorporate it. This context is held
   \emph{whether or not the subprocess encounters difficulty}. A
   supervisor waiting for ten reports holds ten slots regardless of
   whether any report is late.

\item In escalation, the upper level \emph{did not initiate} anything.
   It holds zero context for problems handled below. It loads context
<<<<<<< HEAD
   only when a failure signal arrives from below --- and in steady
=======
   only when a failure signal arrives from below, and in steady
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
   state, failures are rare because lower levels handle most problems.
   There is nothing to ``wait for'' because nothing was requested.
\end{itemize}

<<<<<<< HEAD
This is not a quantitative difference in overhead --- it is a structural
=======
This is not a quantitative difference in overhead; it is a structural
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
difference in what context is held and why. Hierarchical top-down holds
\emph{anticipatory context} (waiting for results that may or may not
arrive). Escalation holds \emph{reactive context} (responding to
failures that have already occurred). Anticipatory context scales with
the number of delegations; reactive context scales with the failure rate.
As the system matures and lower levels improve, the failure rate
decreases, so escalation overhead \emph{decreases with capability}. In
hierarchical top-down, the number of delegations \emph{increases with
capability}: more capability means more subprocesses, each requiring a
delegation slot.

<<<<<<< HEAD
\textbf{Remark on convergence.} Hierarchical top-down that is
sufficiently deep, with intermediaries that handle most problems without
escalating, begins to \emph{resemble} escalation. The intermediary that
resolves a problem without reporting upward is functioning as an
escalation level: it holds no delegation context at the level above.
The convergence is one-directional: as hierarchical top-down becomes
more autonomous at lower levels, it converges toward escalation.
Escalation never converges toward top-down, because it never introduces
anticipatory delegation context.
=======
\textbf{Convergence of hierarchical top-down.} This is the central
point. Hierarchical top-down that is sufficiently deep, with
intermediaries that handle most problems without escalating, begins to
\emph{resemble} escalation. The intermediary that resolves a problem
without reporting upward is functioning as an escalation level: it
holds no delegation context at the level above. The convergence is
one-directional: as hierarchical top-down becomes more autonomous at
lower levels, it converges toward escalation. Escalation never
converges toward top-down, because it never introduces anticipatory
delegation context. Hierarchical decomposition is not an alternative
to escalation; it is the \emph{intermediate regime} through which
top-down converges toward escalation as $|A|$ grows.
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025

\subsection{Event-driven ``top-down'' and causal direction}

A fourth objection: an event-driven top-down supervisor registers
callbacks, holds zero delegation state, and is invoked by subsystems on
completion or failure. This achieves escalation's $O(1)$ steady-state
cost while retaining ``top-down authority.''

<<<<<<< HEAD
The issue is not one of labeling --- it is not that we absorb every
=======
The issue is not one of labeling; it is not that we absorb every
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
efficient architecture under the name ``escalation'' to make the thesis
unfalsifiable. The issue is the \emph{causal direction of the signal}.
The structural distinction between top-down and escalation is:

\begin{itemize}
\item \textbf{Top-down}: the upper level initiates. The signal goes
   \emph{downward}. The upper level says ``do this'' and waits for the
   result.
\item \textbf{Escalation}: the lower level signals. The signal goes
   \emph{upward}. The lower level says ``I cannot handle this'' and the
   upper level engages reactively.
\end{itemize}

In the event-driven architecture, who initiates the callback? The
<<<<<<< HEAD
\emph{subsystem}. The subsystem detected the condition --- whether
completion or failure --- and sent the signal \emph{upward}. The upper
=======
\emph{subsystem}. The subsystem detected the condition, whether
completion or failure, and sent the signal \emph{upward}. The upper
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
level did not poll, did not check, did not anticipate. It was invoked
from below. This is bottom-up signaling regardless of what the
architecture is called.

<<<<<<< HEAD
The claimed top-down authority --- the ability to initiate, cancel, or
redirect --- is unavailable to a supervisor holding zero state. A
supervisor with no delegation context does not know what is happening
below and cannot redirect what it is not tracking. When the callback
arrives, the supervisor must \emph{reload context} to interpret the
result --- exactly what an escalation level does when a failure signal
=======
The claimed top-down authority, the ability to initiate, cancel, or
redirect, is unavailable to a supervisor holding zero state. A
supervisor with no delegation context does not know what is happening
below and cannot redirect what it is not tracking. When the callback
arrives, the supervisor must \emph{reload context} to interpret the
result, exactly what an escalation level does when a failure signal
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
arrives.

The point is not definitional but empirical: if you trace the causal
arrow in the event-driven architecture, it points upward. The
architecture may have been \emph{designed} as top-down (the supervisor
originally assigned the task), but in steady-state operation, the
information flow is bottom-up. The label ``top-down'' describes the
initial assignment; the ongoing control structure is escalation.

\subsection{Hidden costs of escalation during learning}

A second objection: escalation achieves $O(1)$ steady-state cost, but
<<<<<<< HEAD
this assumes lower levels are competent. During learning --- when lower
levels frequently fail --- escalation produces frequent failure signals
=======
this assumes lower levels are competent. During learning, when lower
levels frequently fail, escalation produces frequent failure signals
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
at $L_n$, burdening it with escalation context.

This is correct but not fatal to the argument. During learning, both
architectures burden $L_n$: top-down burdened by delegation overhead
plus the novel problems themselves; escalation burdened only by the
novel problems (the failures that escalate). The difference: under
top-down, the overhead is \emph{additive} to the learning load
(delegation context plus problem context). Under escalation, the
overhead \emph{is} the learning load (failure context only). Escalation
<<<<<<< HEAD
during learning is not $O(1)$ --- it is $O(\text{failure rate} \cdot
|K^{[a]}|)$ --- but it never exceeds the intrinsic difficulty of the
=======
during learning is not $O(1)$; it is $O(\text{failure rate} \cdot
|K^{[a]}|)$, but it never exceeds the intrinsic difficulty of the
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
problems themselves. Top-down adds coordination overhead on top of
intrinsic difficulty.

As learning proceeds and lower levels encode proven actions, escalation
load decreases monotonically. Top-down delegation overhead does not
<<<<<<< HEAD
decrease with learning --- it grows with the number of managed processes,
which increases as the system becomes more capable.

\subsection{Empirical dominance of dense attention}

A third objection: transformers with full self-attention --- the purest
top-down architecture, where every token attends to every other --- have
=======
decrease with learning; it grows with the number of managed processes,
which increases as the system becomes more capable.

\subsection{Empirical success of dense attention}

A third objection: transformers with full self-attention, the purest
top-down architecture, where every token attends to every other, have
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
dominated alternative architectures for a decade. If top-down truly
degraded, why does dense attention win?

The answer is in the trajectory, not the snapshot. Dense attention
<<<<<<< HEAD
dominates \emph{today} at \emph{current scales}. But the trend is
=======
succeeds \emph{today} at \emph{current scales}. But the trend is
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
toward escalation-like structure:

\begin{itemize}
\item \textbf{Sparse attention} (learned sparsity patterns, local
<<<<<<< HEAD
   windows, routing) reduces $O(n^2)$ by learning which edges to keep
   --- learned graph pruning \cite{McCarthy2026a}.
\item \textbf{Mixture-of-experts} routes tokens to specialized
   sub-networks rather than evaluating all parameters for all tokens ---
   escalation at the parameter level.
\item \textbf{Retrieval-augmented generation} retrieves relevant context
   from an external store rather than holding everything in the context
   window --- graph traversal from a query node.
\item \textbf{``Lost in the middle''} \cite{Liu2024}: empirical evidence
   that inference quality degrades as context fills with irrelevant
   content --- precisely the $\rho$ degradation of
   Theorem~\ref{thm:topdown}.
\end{itemize}

The industry is converging toward escalation. Dense attention is the
starting point, not the destination. The cost pressures the theorem
identifies are driving the convergence.
=======
   windows, routing) reduces $O(n^2)$ by learning which edges to
   keep---learned graph pruning \cite{McCarthy2026a}.
\item \textbf{Mixture-of-experts} routes tokens to specialized
   sub-networks rather than evaluating all parameters for all tokens:
   escalation at the parameter level.
\item \textbf{Retrieval-augmented generation} retrieves relevant context
   from an external store rather than holding everything in the context
   window: graph traversal from a query node.
\item \textbf{``Lost in the middle''} \cite{Liu2024}: empirical evidence
   that inference quality degrades as context fills with irrelevant
   content: precisely the $\rho$ degradation of
   Theorem~\ref{thm:topdown}.
\end{itemize}

The industry is converging toward hybrid architectures that incorporate
escalation. Dense attention is the starting point, not the destination.
The cost pressures the theorem identifies are driving the convergence.
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025

\textbf{Remark: the LLM theory of the case.}
Theorem~\ref{thm:topdown} does not claim that large language models
\emph{cannot work}. Dense attention over a large context window is a
<<<<<<< HEAD
viable strategy as long as the context can cover the problem space ---
=======
viable strategy as long as the context can cover the problem space,
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
that is, as long as the system operates below the crossover
(Section~\ref{sec:crossover}). Current LLMs succeed precisely because
their context windows are large enough, relative to the problems they
are asked to solve, that $\rho$ degradation is tolerable.

What the theorem establishes is that this strategy is
\emph{suboptimal and expensive}. Full self-attention recomputes the
pairwise adjacency structure from scratch at every query: $O(n^2)$ per
inference step. A system with stored adjacency pays $O(d)$ per
retrieval, where $d$ is the local degree. The cost ratio grows
quadratically with context size. For a 128k-token context window, the
<<<<<<< HEAD
system computes $\sim$$10^{10}$ pairwise attention weights per layer per
=======
system computes ${\sim}10^{10}$ pairwise attention weights per layer per
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
query, the vast majority of which are near-zero (irrelevant pairs). This
is the computational cost of not having a graph.

The theorem also predicts a ceiling. As the problems LLMs are asked to
<<<<<<< HEAD
solve grow in complexity --- requiring more knowledge, more reasoning
steps, more integration across domains --- the context required grows
=======
solve grow in complexity, requiring more knowledge, more reasoning
steps, more integration across domains, the context required grows
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
with the problem, and $\rho$ degrades: ``lost in the middle''
\cite{Liu2024} is the empirical manifestation. Growing the context
window defers the crossover but does not eliminate it, because the
overhead of attending to irrelevant content grows quadratically while
the benefit of additional context grows sub-linearly (most added context
is irrelevant to any specific problem).

<<<<<<< HEAD
The LLM architecture is not invalidated by this result --- it is
=======
The LLM architecture is not invalidated by this result; it is
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
\emph{located} by it. LLMs operate in the pre-crossover regime where
dense top-down attention is functional but costly. The industry
trajectory toward sparse attention, mixture-of-experts, and
retrieval-augmented generation is movement toward the post-crossover
architecture the theorem identifies as necessary: stored adjacency,
selective retrieval, escalation on failure.

<<<<<<< HEAD
\textbf{Remark on scope.} Theorem~\ref{thm:topdown} is not a claim
that top-down control is wrong in all cases. Occasional top-down
intervention --- where $L_n$ deliberately overrides a lower-level
response --- is a valid and sometimes necessary operation. The theorem
establishes that top-down allocation as the \emph{primary} control
architecture degrades under bounded $C_n$. The distinction is between
top-down as occasional intervention (compatible with escalation) and
top-down as the default routing mechanism (incompatible with bounded
context).
=======
\textbf{Remark on scope.} Theorem~\ref{thm:topdown} does not claim
that top-down control is unnecessary. Top-down allocation handles
genuinely novel problems that no lower level can resolve, and
occasional top-down intervention, where $L_n$ deliberately overrides a
lower-level response, is a valid and sometimes necessary operation. The
theorem establishes that top-down allocation as the \emph{sole} control
architecture is unsustainable under bounded $C_n$. Escalation is what
preserves the context capacity that top-down allocation needs. The
distinction is between top-down as the exclusive routing mechanism
(unsustainable under bounded context) and top-down as the
highest-level component in a complementary architecture where
escalation handles the routine.
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025

\textbf{Biological measurement.} The empirical signature of escalation
is the transition from active to automatic processing across skill
acquisition. When an informed action $a$ is first learned, it consumes
$L_n$ capacity: the learner must actively reason through each step. As
<<<<<<< HEAD
encoding proceeds --- as $K^{[a]}$ migrates from $L_n$ to lower levels
--- $a$ ceases to consume $L_n$ capacity. This is precisely the
=======
encoding proceeds, as $K^{[a]}$ migrates from $L_n$ to lower levels,
$a$ ceases to consume $L_n$ capacity. This is precisely the
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
escalation architecture: $a$ has been encoded at a level where it no
longer requires active context. The ubiquity of this transition across
motor learning, language acquisition, and expert skill development
\cite{Kahneman2011} is evidence that biological systems implement
escalation rather than persistent top-down allocation.

% ------------------------------------------------------------------
\section{Comparison: Escalation vs.\ Top-Down}
\label{sec:comparison}
% ------------------------------------------------------------------

The two architectures can be compared directly on their steady-state
context demands.

\subsection{Escalation architecture}

Under escalation, $L_n$ engages only when lower levels fail to bound
uncertainty for a given problem.

\begin{itemize}
\item \textbf{Steady-state context per problem:} $O(|K^{[a]}|)$.
   Only the knowledge relevant to the currently escalated problem
   occupies $L_n$. No delegation overhead is incurred because lower
<<<<<<< HEAD
   levels are not ``managed'' --- they operate autonomously.
=======
   levels are not ``managed''; they operate autonomously.
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025

\item \textbf{Effect of knowledge growth:} As $K$ grows and more
   actions are encoded at lower levels, \emph{fewer} problems reach
   $L_n$. The load on active context decreases or remains bounded as
   capability increases.

\item \textbf{Scaling property:} Capability growth \emph{relieves}
   $L_n$ rather than burdening it. Each successfully encoded action
   is one fewer demand on active context.
\end{itemize}

\subsection{Top-down allocation architecture}

Under top-down allocation, $L_n$ holds delegation context for all
managed subprocesses.

\begin{itemize}
\item \textbf{Steady-state context per problem:}
   $O(|K^{[a]}| + k \cdot \alpha)$, where $k$ is the number of
   concurrent delegations. The delegation overhead $k \cdot \alpha$
   is additive to the problem's own knowledge requirements.

\item \textbf{Effect of capability growth:} As $A$ grows, $k$ grows
   proportionally (the supervisor must track more subprocesses).
   $L_n$ load increases with capability.

\item \textbf{Scaling property:} Capability growth \emph{burdens}
   $L_n$. Each new capability adds to the supervisory overhead,
   eventually saturating active context.
\end{itemize}

\subsection{The inversion}

The escalation architecture inverts the scaling relationship between
capability and cognitive load:

\begin{center}
\begin{tabular}{@{}lcc@{}}
\toprule
& \textbf{Escalation} & \textbf{Top-Down} \\
\midrule
$L_n$ engagement & On failure only & Continuous supervision \\
Steady-state context & $O(|K^{[a]}|)$ & $O(|K^{[a]}| + k \cdot
\alpha)$ \\
As $|A|$ grows & $L_n$ load $\downarrow$ or bounded &
$L_n$ load $\uparrow$ proportionally \\
At capacity & Novel problems get full $C_n$ &
Novel problems get $C_n - k\alpha$ \\
\bottomrule
\end{tabular}
\end{center}

Under escalation, a system that has encoded 10{,}000 actions at lower
levels has \emph{more} active context available for novel problems than
<<<<<<< HEAD
when it had encoded 100 --- because 9{,}900 additional problems are
now handled without $L_n$ engagement. Under top-down allocation, a
system managing 10{,}000 subprocesses has \emph{less} active context
available than when it managed 100 --- because 9{,}900 additional
=======
when it had encoded 100, because 9{,}900 additional problems are
now handled without $L_n$ engagement. Under top-down allocation, a
system managing 10{,}000 subprocesses has \emph{less} active context
available than when it managed 100, because 9{,}900 additional
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
delegation slots consume context.

This is not a quantitative difference but a qualitative inversion of
the scaling direction. It explains why expertise feels like
\emph{effortlessness}: the expert has encoded actions that the novice
must actively reason through, freeing $L_n$ for genuinely novel
problems.

% ------------------------------------------------------------------
\section{Engagement with Prior Work}
\label{sec:prior}
% ------------------------------------------------------------------

\subsection{Free Energy Principle and Active Inference}

The Free Energy Principle \cite{Friston2010,Friston2017} posits that
biological systems minimize variational free energy, with attention
implemented as precision weighting: higher cortical levels set the gain
<<<<<<< HEAD
on prediction errors at lower levels. This is a top-down architecture
in the sense of Theorem~\ref{thm:topdown}: the supervisory level must
maintain a representation of expected precision for each lower-level
channel, and must evaluate incoming prediction errors against these
expectations.

Theorem~\ref{thm:topdown} shows that this precision allocation
degrades under bounded $C_n$ because the delegation overhead ---
maintaining expected precision for $k$ channels --- fills the active
context window proportionally to the number of monitored channels.
This is a critique of the architectural \emph{prescription}, not of the
minimization objective. Free energy minimization may well be the
correct objective function; the question is whether the control
architecture that implements it should be top-down (precision
allocation from above) or bottom-up (escalation on prediction error
that exceeds local capacity). Under bounded context, escalation
dominates.

More precisely: FEP's precision weighting may be optimal when $C_n$ is
not binding --- when active context is large relative to the number of
monitored processes. Under bounded context, the overhead of maintaining
precision weights for all channels eventually exceeds the benefit of
optimal allocation, and escalation (where channels signal only on
failure) achieves better inference quality at $L_n$.
=======
on prediction errors at lower levels. Precision weighting is not
top-down delegation in the sense of Definition~\ref{def:delegation}---it
does not track individual subproblems with return-integration context.
It is a gain modulation mechanism: a higher level sets a parameter on a
lower-level channel. The cost per channel is lighter than full
delegation overhead.

The convergence result still applies, because precision weighting has
a nonzero cost per channel. Each precision parameter the higher level
maintains is a representation it must hold, update, and evaluate against
incoming prediction errors. At small scale, when the number of
channels $k$ is modest, this cost fits comfortably within $C_n$ and
precision weighting is viable---potentially optimal. As the system
scales and $k$ grows, maintaining precision parameters for all channels
consumes an increasing fraction of $C_n$. The system must shed active
precision tracking for routine channels: channels whose prediction
errors are consistently small no longer need higher-level gain
modulation and can operate at locally set precision. This shedding is
the same convergence Theorem~\ref{thm:topdown} identifies: at scale,
the system retains active precision weighting only for channels with
unresolved uncertainty and lets routine channels self-regulate.

FEP's precision weighting is therefore an instance of the convergence,
not an exception to it. At small scale, centralized precision
allocation is viable. At scale, it converges toward a hybrid where
most channels operate at locally determined precision (escalation) and
higher-level precision allocation is reserved for channels where
prediction errors signal genuine novelty. The two mechanisms are
complementary: escalation handles routine precision; active allocation
directs the bounded capacity that remains.
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025

\subsection{System M: Dupoux, LeCun, and Malik}

Dupoux, LeCun, and Malik \cite{DupouxLeCunMalik2026} propose a
three-component architecture: System~A (observation/world models),
System~B (action/planning), and System~M (meta-control), where M is a
fixed evolutionary routing policy that directs when A and B engage.
System~M monitors epistemic signals and routes data between
subsystems.

This is explicitly top-down: M sits above A and B, monitoring their
states and directing their activity. Theorem~\ref{thm:topdown}
establishes that M's monitoring itself consumes $C_n$ proportionally
to the number of subsystems and channels it oversees. As the
<<<<<<< HEAD
architecture scales --- more world models, more action policies, more
channels --- M's overhead grows, eventually crowding out the capacity
for the reasoning that M is supposed to coordinate.

Their architecture would benefit from replacing top-down routing with
escalation: subsystems A and B signal M only on failure (when local
uncertainty exceeds a threshold), rather than M continuously polling
all subsystems for state. Under escalation, M engages only when a
subsystem cannot resolve its own uncertainty --- and the overhead at
M's level is proportional to the \emph{failure rate}, not the
\emph{system size}.
=======
architecture scales, more world models, more action policies, more
channels, M's overhead grows, eventually crowding out the capacity
for the reasoning that M is supposed to coordinate.

Their architecture would benefit from complementing top-down routing
with escalation: subsystems A and B signal M only on failure (when
local uncertainty exceeds a threshold), rather than M continuously
polling all subsystems for state. Under this hybrid, M retains
authority for genuinely novel coordination but engages only when a
subsystem cannot resolve its own uncertainty, so the overhead at
M's level is proportional to the \emph{failure rate}, not the
\emph{system size}. Escalation is what keeps M viable as the
architecture scales.
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025

\subsection{Global Workspace Theory}

Global Workspace Theory \cite{Baars1988,Dehaene2011} already
describes what is essentially an escalation architecture: specialized
local processors handle routine computation autonomously, and the
global workspace is engaged only when local processing fails or when
integration across modules is required. Broadcasting to the global
workspace is the cognitive analogue of escalation to $L_n$.

Theorem~\ref{thm:topdown} provides the formal grounding that GWT
assumes architecturally. GWT \emph{asserts} that local processors
handle routine work and the global workspace handles exceptions;
<<<<<<< HEAD
Theorem~\ref{thm:escalation} \emph{derives} the escalation criterion
from uncertainty-reduction cost, and Theorem~\ref{thm:topdown}
\emph{proves} why the alternative (top-down allocation from the global
workspace) fails. The $O(k \cdot \alpha)$ overhead is the formal
reason why global workspace engagement must be the exception rather
than the rule.
=======
the escalation policy (Definition~\ref{def:escalation}) identifies the
cost-minimizing criterion from the monotone cost structure, and
Theorem~\ref{thm:topdown}
\emph{proves} why the global workspace requires escalation to remain
viable. The $O(k \cdot \alpha)$ overhead is the formal
reason why global workspace engagement must be the exception rather
than the rule: escalation keeps $k$ bounded so the workspace retains
capacity for genuine integration.
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025

\subsection{System 1 and System 2}

Kahneman's System~1/System~2 framework \cite{Kahneman2011} is the
popular binary version of the encoding hierarchy. System~1 (fast,
automatic, low-cost) corresponds to lower levels of $E(c,r)$;
System~2 (slow, deliberate, high-cost) corresponds to $L_n$.

The continuous parameterization $E(c,r)$ corrects the binary
discretization. There are not two systems but a continuous space of
encodings with substrate-determined interfaces. A motor reflex
($c \approx 0$, $r \approx 0$) and a well-practiced heuristic
($c$ low, $r$ moderate) are both ``System~1'' in Kahneman's
taxonomy, but they occupy different regions of $E(c,r)$ with different
costs and different revisability. The binary collapses this structure
into a distinction that, while useful for popular exposition, obscures
the continuous nature of encoding and the multiplicity of interfaces.

<<<<<<< HEAD
Theorem~\ref{thm:escalation} also corrects the implicit control
=======
The escalation policy also corrects the implicit control
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
assumption: System~2 does not ``decide'' when to override System~1.
Rather, System~1 (lower levels) handles problems autonomously, and
System~2 ($L_n$) engages when System~1 fails. The direction of
control is bottom-up, not top-down.

<<<<<<< HEAD
=======
\subsection{Automatic and Controlled Processing}

Shiffrin and Schneider \cite{ShiffrinSchneider1977} established the
foundational distinction between automatic processing (fast, parallel,
low-effort, not limited by capacity) and controlled processing (slow,
serial, effortful, capacity-limited). Automatic processing corresponds
to actions encoded at lower levels of $E(c,r)$; controlled processing
corresponds to $L_n$. Their key empirical finding---that practice
converts controlled processing into automatic processing---is exactly
the promotion mechanism of the encoding hierarchy: repeated successful
execution at $L_n$ promotes an action to lower encoding levels, freeing
$C_n$ for novel problems.

Anderson's ACT-R \cite{Anderson1993} formalizes this as
proceduralization: declarative knowledge (propositions in $K$ loaded
into active context for deliberate use) compiles into procedural
knowledge (condition-action rules that fire without loading $K$). In
the present framework, proceduralization is movement along $E(c,r)$
from high-cost, high-reversibility regions toward low-cost,
low-reversibility regions. ACT-R's activation-based retrieval---where
propositions compete for limited working memory based on
recency, frequency, and contextual relevance---is a specific
implementation of contextual retrieval under bounded $C_n$.

Both frameworks describe the empirical trajectory that escalation
predicts: as entities scale their action repertoire, an increasing
fraction of actions must operate automatically (at lower encoding
levels) because controlled processing at $L_n$ cannot supervise
them all. The convergence result (Theorem~\ref{thm:topdown}) provides
the formal reason: the $O(k \cdot \alpha)$ overhead forces this
transition regardless of substrate.

\subsection{Control Theory: Requisite Variety and the Good Regulator}

Three results from control theory provide formal antecedents for the
convergence result.

\textbf{Ashby's Law of Requisite Variety} \cite{Ashby1956} establishes
that a controller must have at least as many distinct responses as the
disturbances it faces. Under bounded $C_n$, the regulator at $L_n$ has
variety bounded by $C_n$. When the system's variety---the number of
distinct situations requiring distinct responses---exceeds $C_n$, no
single regulator at $L_n$ can achieve requisite variety. The only
solution is to distribute the regulatory function across encoding
levels, each regulating its local domain, with $L_n$ regulating only
the residual that lower levels cannot handle. This distribution is
escalation. Escalation is how a bounded regulator achieves requisite
variety over an unbounded system: not by expanding the regulator's
capacity (bounded by substrate) but by distributing regulation to
levels where local variety is manageable.

\textbf{The Conant-Ashby Good Regulator Theorem} \cite{ConantAshby1970}
proves that every good regulator of a system must contain a model of
that system. Under top-down allocation, the delegation state
(Definition~\ref{def:delegation}---specification, expected return,
integration context) \emph{is} this model: it is $L_n$'s
representation of each subprocess it regulates. The good regulator
theorem establishes that this overhead is not a design inefficiency but
a \emph{theorem-mandated cost}: effective regulation requires the model,
and the model requires context. The $O(k \cdot \alpha)$ overhead of
Theorem~\ref{thm:topdown} is therefore irreducible for the supervised
fraction of $A$. This strengthens the convergence argument: the
overhead cannot be optimized away because it is the minimum model
that the good regulator theorem demands.

\textbf{Witsenhausen's counterexample} \cite{Witsenhausen1968}
demonstrates that optimal decentralized control, where multiple
controllers operate with different information, can be fundamentally
harder than centralized control. Escalation is a decentralized
architecture: each level controls with local information. In
Witsenhausen's framework, this sacrifices the optimality that a
centralized controller with full information could achieve. But
bounded $C_n$ rules out the centralized option at scale: the good
regulator theorem requires a model that exceeds $C_n$. The relevant
comparison is not between escalation and an ideal centralized
controller but between escalation and a centralized controller that
has overflowed its context and can no longer regulate effectively.
Survival does not require Witsenhausen-optimal control. It requires
a viable architecture, and escalation is the viable architecture when
the centralized alternative has been eliminated by bounded context.

\subsection{Subsumption Architecture}

Brooks's subsumption architecture \cite{Brooks1986} is the engineering
instantiation of the convergence result. In subsumption, lower-level
behavioral layers (obstacle avoidance, wandering) run autonomously and
continuously. Higher layers (exploration, goal-seeking) can subsume
lower-layer outputs when they have something to contribute, but do not
monitor or schedule lower layers during routine operation. The control
direction is bottom-up: lower layers act by default, higher layers
intervene on exception.

This is escalation. Each behavioral layer handles its local domain
without supervisory overhead from above; higher layers engage only when
lower layers produce outcomes that require coordination or override.
Brooks arrived at this architecture empirically, by observing that
centralized planners could not scale to real-time robotic control.
Theorem~\ref{thm:topdown} derives the same conclusion formally: the
$O(k \cdot \alpha)$ overhead of centralized supervision exceeds $C_n$
at scale, forcing the system toward distributed local control with
higher-level exception handling.

The subsumption architecture also illustrates the one-directionality
of the convergence. Brooks did not begin with escalation and converge
toward centralized planning. He began with centralized planning
(the STRIPS tradition), found it unscalable, and replaced it with
subsumption. The reverse trajectory---starting from subsumption and
converging toward centralized planning---has not occurred in the
robotics literature, consistent with the irreversibility of escalation
(Theorem~\ref{thm:topdown}).

>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
\subsection{Resource Rationality}

Griffiths, Lieder, and Goodman \cite{GriffithsEtAl2015} establish
that the optimal algorithm for a cognitive task depends on the
<<<<<<< HEAD
computational resources available --- resource rationality. The
=======
computational resources available: resource rationality. The
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
optimal unconstrained algorithm may be suboptimal under resource
limitations.

Theorem~\ref{thm:topdown} is precisely an instance of this principle.
Top-down allocation may be the optimal control architecture when $C_n$
is unbounded: with infinite active context, the overhead of maintaining
delegation state for all subprocesses is negligible, and the benefit
of optimal routing is maximal. Under bounded $C_n$, the overhead is
<<<<<<< HEAD
no longer negligible, and escalation --- a ``simpler'' architecture
that does not attempt global optimization of routing --- dominates
because it conserves the scarce resource (active context) for genuine
inference rather than spending it on coordination.
=======
no longer negligible, and escalation becomes necessary to conserve the
scarce resource (active context) for genuine inference rather than
spending it on coordination.
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025

Resource rationality provides the general principle;
Theorem~\ref{thm:topdown} provides the specific result for the
control architecture domain: the resource-rational control
<<<<<<< HEAD
architecture under bounded context is escalation.
=======
architecture under bounded context is a hybrid where escalation
handles routine problems and top-down allocation addresses the
residual that only $L_n$ can resolve.
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025

% ------------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}
% ------------------------------------------------------------------

<<<<<<< HEAD
We have shown that escalation dominance is a \emph{scaling result}.
Top-down allocation is viable --- and may be optimal --- when the action
repertoire $|A|$ fits within bounded context $C_n$. But intelligence
requires $|A| \gg C_n$: an entity must know far more than it can hold
in mind at once. Once $|A|$ exceeds the crossover point
$|A|^* = C_n / 2\alpha$ (Section~\ref{sec:crossover}), top-down
delegation overhead dominates active context and inference quality
degrades monotonically. The crossover is permanent under bounded $C_n$.

The argument has three components:

\begin{enumerate}
\item Theorem~\ref{thm:escalation} establishes that informed actions
   are handled at the lowest sufficient level, with escalation
   occurring only on failure.

\item Theorem~\ref{thm:topdown} establishes that top-down allocation
   degrades inference quality as $|A|$ grows, whether the architecture
   is flat ($k = |A|$) or hierarchical ($k = b \cdot \log_b |A|$).
   Recovery requires unbounded growth of $C_n$, which contradicts the
   boundedness assumption.

\item The crossover analysis (Section~\ref{sec:crossover}) identifies
   the regime boundary: top-down works below $|A|^*$, fails above it.
   Every system we recognize as intelligent operates well above it.
\end{enumerate}

The combination yields a precise result: escalation is not universally
superior to top-down --- it is superior at every scale that matters.
$L_n$ is the level of last resort, not the seat of control.
Capability growth, under escalation, \emph{relieves} the highest level
rather than burdening it. Under top-down allocation, capability growth
\emph{burdens} the highest level until it saturates.
=======
We have shown that any control architecture converges toward
escalation as the action repertoire $|A|$ grows under bounded context
$C_n$. Top-down allocation is viable at small scale, when concurrent
demands fit within $C_n$. But intelligence requires $|A| \gg C_n$,
and the convergence is forced by three steps:

\begin{enumerate}
\item The escalation policy (Definition~\ref{def:escalation}) is
   cost-minimizing under monotone encoding cost: handle each action at
   the lowest sufficient level, escalate on failure. This is immediate
   from the cost structure.

\item Theorem~\ref{thm:topdown} establishes that top-down tracking
   incurs $O(k \cdot \alpha)$ context overhead, and that reducing $k$
   by shedding supervisory tracking of routine actions is the only
   recovery under bounded $C_n$. The shed fraction operates under
   escalation.

\item The crossover analysis (Section~\ref{sec:crossover}) locates the
   threshold at $k^* = C_n / 2\alpha$---2--3 concurrent delegations
   for biological working memory. Beyond it, any additional concurrent
   demand must be handled by escalation or inference quality degrades.
\end{enumerate}

The convergence is one-directional. As top-down becomes more
autonomous at lower levels, it converges toward escalation.
Escalation never converges toward top-down, because it never
introduces anticipatory delegation context. The endpoint is a hybrid:
escalation handles the routine; top-down handles the novel. Capability
growth, under this hybrid, \emph{relieves} the highest level rather
than burdening it.

The falsifiable content of this result is the \emph{cost prediction},
not the label. The claim is not that every efficient architecture is
``really escalation''---which would be unfalsifiable by definitional
absorption. The claim is that top-down tracking costs $O(k \cdot
\alpha)$ per delegation cycle, and that $k$ grows with $|A|$ under
positive novelty rate (Section~\ref{sec:crossover}). A system that
maintains top-down tracking of all actions at scale without paying
$O(k \cdot \alpha)$ overhead would falsify the theorem.
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025

This result has implications for both cognitive science and AI
architecture. In cognitive science, it provides the formal grounding
for Global Workspace Theory's architectural assumptions and corrects
the implicit top-down control assumption in System~1/System~2 accounts.
<<<<<<< HEAD
In AI architecture, it identifies a structural ceiling on centralized
routing: systems scaling beyond $|A|^*$ should implement escalation ---
subsystems that operate autonomously and signal on failure --- rather
than supervisory routing that attempts to manage all subprocesses from
above. The trajectory of transformer architectures, from dense
attention toward sparse attention, mixture-of-experts, and
retrieval-augmented generation, is convergence toward this result.

The series to this point has established two results: knowledge under
bounded context necessarily forms a graph \cite{McCarthy2026a}; and the
control architecture over that graph is necessarily escalation under
bounded context (this paper). Together, these results constrain the
architecture of any system that accumulates knowledge, acts under
uncertainty, and operates within bounded active context --- which is
every physical system that persists.
=======
In AI architecture, it explains the trajectory from dense attention
toward sparse attention, mixture-of-experts, and retrieval-augmented
generation: this trajectory is not a design trend but the scaling
convergence the theorem predicts. Systems scaling beyond $k^*$
\emph{must} adopt escalation-like structure or degrade.

The series to this point has established two results: knowledge under
bounded context necessarily forms a graph \cite{McCarthy2026a}; and
the control architecture over that graph converges toward escalation
as capability grows under bounded context (this paper). Together,
these results constrain the architecture of any system that
accumulates knowledge, acts under uncertainty, and operates within
bounded active context---which is every physical system that persists.

\section*{Acknowledgments}

The author used Claude (Anthropic) for drafting assistance, literature review, and \LaTeX{} preparation. All theoretical content, proofs, and arguments are the author's own.
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025

\bibliographystyle{plain}
\begin{thebibliography}{99}

<<<<<<< HEAD
=======
\bibitem{Anderson1993}
J.~R. Anderson.
\newblock \emph{Rules of the Mind}.
\newblock Lawrence Erlbaum Associates, 1993.

\bibitem{Ashby1956}
W.~R. Ashby.
\newblock \emph{An Introduction to Cybernetics}.
\newblock Chapman \& Hall, 1956.

>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
\bibitem{Baars1988}
B.~J. Baars.
\newblock \emph{A Cognitive Theory of Consciousness}.
\newblock Cambridge University Press, 1988.

<<<<<<< HEAD
\bibitem{Clark2015}
A.~Clark.
\newblock \emph{Surfing Uncertainty: Prediction, Action, and the Embodied
  Mind}.
\newblock Oxford University Press, 2015.
=======
\bibitem{Brooks1986}
R.~A. Brooks.
\newblock A robust layered control system for a mobile robot.
\newblock \emph{IEEE Journal on Robotics and Automation}, 2(1):14--23, 1986.
>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025

\bibitem{Cowan2001}
N.~Cowan.
\newblock The magical number 4 in short-term memory: A reconsideration of
  mental storage capacity.
\newblock \emph{Behavioral and Brain Sciences}, 24(1):87--114, 2001.

<<<<<<< HEAD
=======
\bibitem{ConantAshby1970}
R.~C. Conant and W.~R. Ashby.
\newblock Every good regulator of a system must be a model of that system.
\newblock \emph{International Journal of Systems Science}, 1(2):89--97, 1970.

>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
\bibitem{Cover2006}
T.~M. Cover and J.~A. Thomas.
\newblock \emph{Elements of Information Theory}.
\newblock Wiley-Interscience, 2nd edition, 2006.

\bibitem{Dehaene2011}
S.~Dehaene.
\newblock \emph{Consciousness and the Brain: Deciphering How the Brain Codes
  Our Thoughts}.
\newblock Viking Press, 2011.

\bibitem{DupouxLeCunMalik2026}
E.~Dupoux, Y.~LeCun, and J.~Malik.
\newblock Why {AI} systems don't learn and what to do about it: Lessons on
  autonomous learning from cognitive science.
\newblock \emph{arXiv preprint arXiv:2603.15381}, 2026.

\bibitem{Friston2010}
K.~Friston.
\newblock The free-energy principle: A unified brain theory?
\newblock \emph{Nature Reviews Neuroscience}, 11(2):127--138, 2010.

\bibitem{Friston2017}
K.~Friston.
\newblock Active inference and learning.
\newblock \emph{Neuroscience \& Biobehavioral Reviews}, 68:862--879, 2017.

\bibitem{GriffithsEtAl2015}
T.~L. Griffiths, F.~Lieder, and N.~D. Goodman.
\newblock Rational use of cognitive resources: Levels of analysis between the
  computational and the algorithmic.
\newblock \emph{Topics in Cognitive Science}, 7(2):217--229, 2015.

\bibitem{Kahneman2011}
D.~Kahneman.
\newblock \emph{Thinking, Fast and Slow}.
\newblock Farrar, Straus and Giroux, 2011.

\bibitem{Liu2024}
N.~F. Liu, K.~Lin, J.~Hewitt, A.~Paranjape, M.~Bevilacqua, F.~Petroni, and
  P.~Liang.
\newblock Lost in the middle: How language models use long contexts.
\newblock \emph{Transactions of the Association for Computational Linguistics},
  12:157--173, 2024.

\bibitem{McCarthy2026a}
P.~D. McCarthy.
\newblock Graph structure is necessary for information preservation under
  bounded context.
\newblock Working paper, 2026.

\bibitem{Simon1955}
H.~A. Simon.
\newblock A behavioral model of rational choice.
\newblock \emph{The Quarterly Journal of Economics}, 69(1):99--118, 1955.

\bibitem{Simon1962}
H.~A. Simon.
\newblock The architecture of complexity.
\newblock \emph{Proceedings of the American Philosophical Society},
  106(6):467--482, 1962.

<<<<<<< HEAD
=======
\bibitem{ShiffrinSchneider1977}
R.~M. Shiffrin and W.~Schneider.
\newblock Controlled and automatic human information processing: {II}.
  Perceptual learning, automatic attending and a general theory.
\newblock \emph{Psychological Review}, 84(2):127--190, 1977.

>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
\bibitem{Williamson1967}
O.~E. Williamson.
\newblock Hierarchical control and optimum firm size.
\newblock \emph{Journal of Political Economy}, 75(2):123--138, 1967.

<<<<<<< HEAD
=======
\bibitem{Witsenhausen1968}
H.~S. Witsenhausen.
\newblock A counterexample in stochastic optimum control.
\newblock \emph{SIAM Journal on Control}, 6(1):131--147, 1968.

>>>>>>> 319f447daa46f8c1e949fb7929535d1aa6396025
\end{thebibliography}

\end{document}
