\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}
\hypersetup{hidelinks}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\theoremstyle{remark}
\newtheorem{remark}{Remark}

\setlength{\parskip}{4pt plus 1pt minus 1pt}
\setlength{\parindent}{0pt}

\title{Cancer as Escalation Chain Severance\\Under Bounded Context}
\author{Patrick D.\ McCarthy}
\date{}

\begin{document}
\maketitle

\begin{abstract}
We apply the bounded-context framework of Papers~1--4 to
multicellular organization, proposing that cancer is a predictable
failure mode of any multi-level escalation hierarchy. In a
multicellular organism, individual cells participate in an escalation
chain -- cell $\to$ tissue $\to$ organ $\to$ organism -- where
each level extends the effective bounded context
$C_n^{\mathrm{eff}}$ of the level below via intercellular signaling.
We argue that severing the chain at level $k$ forces the cell back
to $C_n^{(k)}$-level optimization, where the action space available under
cell-level bounded context converges on the recognized hallmarks of
cancer: aerobic glycolysis, unconstrained proliferation, immune
evasion, and motility. This convergence is not atavistic reversion
to ancient programs but \emph{structural necessity}. The same
constraint ceiling that governed unicellular life now governs the
severed cell. The framework offers structural explanations for three
phenomena that existing theories address incompletely: (1)~the
velocity of malignant transformation (structural failure, not
stochastic accumulation), (2)~the convergent phenotype across tissue
types (same constraint, same necessary structure), and (3)~the
therapeutic resistance of established tumors (locally rational
optimization under cell-level $C_n$). Clinical oncology is already
trending toward the therapeutic strategies this framework predicts
-- differentiation therapy, immunotherapy, and microenvironment
normalization -- but largely without a unified theoretical
justification. We propose that the bounded-context framework
provides one, and offer testable predictions that distinguish this
account from the somatic mutation theory, the atavistic theory, and
the tissue organization field theory.
\end{abstract}

% ------------------------------------------------------------------
\section{Introduction}
\label{sec:intro}
% ------------------------------------------------------------------

Cancer remains the central unsolved problem of multicellular biology.
Three major theoretical frameworks compete to explain it, each
capturing part of the phenomenon while missing the rest.

The \emph{somatic mutation theory} (SMT) holds that cancer arises from
accumulated genetic mutations that progressively disable growth
controls \cite{Vogelstein2013, Nowell1976}. SMT explains the
stochastic component of cancer incidence and the role of specific
oncogenes and tumor suppressors. It does not explain the \emph{velocity}
of malignant transformation: why cancers from different tissues, with
different mutational profiles, converge on the same phenotypic
hallmarks \cite{Hanahan2011} far faster than random mutation would
predict.

The \emph{atavistic theory} holds that cancer cells reactivate ancient
genetic programs from the unicellular era, effectively reverting to a
pre-multicellular phenotype \cite{Davies2011, Trigos2017}. The
phylostratigraphic evidence is striking: genes upregulated in cancer
are disproportionately ancient, dating to unicellular ancestors
\cite{Trigos2017}. But the theory lacks a mechanism: \emph{why} would
a modern cell access billion-year-old programs, and \emph{how} are
those programs still coherent after extensive evolutionary modification?

The \emph{tissue organization field theory} (TOFT) holds that cancer
is not a disease of individual cells but of tissue organization:
the breakdown of intercellular signaling and tissue architecture
enables default proliferative behavior \cite{Sonnenschein2000}. TOFT
correctly identifies the organizational level at which cancer operates
but lacks a formal account of \emph{why} organizational breakdown
produces the specific phenotypes it does.

This paper proposes that all three theories describe aspects of a
single structural phenomenon: \textbf{escalation chain severance}.
The bounded-context framework of Papers~1--4
\cite{McCarthy2026a, McCarthy2026b, McCarthy2026c, McCarthy2026d}
provides a formal lens through which to unify them. We model
multicellular organization as a multi-level escalation hierarchy
where each level extends the effective bounded context of the level
below. Severing the chain is a structural failure that forces
cell-level optimization. The convergent phenotype, on this account,
is not inherited from ancestors but \emph{necessary} under cell-level
bounded context. The same constraint that shaped unicellular life
now shapes the severed cell.

\textbf{Scope and epistemology.} This paper is a theoretical
application of the bounded-context framework to cancer biology.
Papers~1--4 derive their results from formal proofs; the present
paper applies that framework to a biological domain where the
mapping from formalism to biology is conjectural. We do not claim
to prove anything about cancer. We claim that the framework generates
a coherent structural account that (a)~unifies three competing
theories, (b)~aligns with the direction clinical oncology is already
moving, and (c)~produces testable predictions that distinguish it
from existing accounts. The value of the paper is the structural
reasoning and the predictions, not empirical validation, which
is the province of cancer biologists and clinicians.

% ------------------------------------------------------------------
\section{Formal Framework}
\label{sec:framework}
% ------------------------------------------------------------------

We extend the entity framework of Papers~1--4 to multi-level
biological organization.

\begin{definition}[Biological Entity at Level $k$]
\label{def:bio-entity}
A biological entity at organizational level $k$ is
$E_k = (K_k, A_k, \sigma_k, \gamma_k)$ with bounded active context
$C_n^{(k)}$, where:
\begin{itemize}
\item $K_k$ is the knowledge base at level $k$ (the regulatory state
   accessible for inference: gene expression profiles, signaling
   molecule concentrations, tissue-level morphogen gradients).
\item $A_k$ is the action space at level $k$ (the executable responses
   --- protein synthesis, cell division, apoptosis, migration,
   differentiation signals).
\item $\sigma_k$ is the sensing function at level $k$ (receptors,
   gap junctions, paracrine signaling).
\item $\gamma_k$ is the agency at level $k$ (the probability that
   $E_k$ engages its adaptive mechanism in response to signals).
\end{itemize}
The organizational levels in a metazoan are:
\begin{align*}
k = 0 &: \text{cell} \\
k = 1 &: \text{tissue} \\
k = 2 &: \text{organ} \\
k = 3 &: \text{organism}
\end{align*}
\end{definition}

\begin{definition}[Escalation Chain]
\label{def:chain}
An \emph{escalation chain} is a sequence of coupled entities
$E_0, E_1, \ldots, E_m$ where level $k+1$ receives escalation
signals from level $k$ when $E_k$ encounters states exceeding its
local capacity:
\[
U_k(w, K_k) > U_{\mathrm{threshold}}^{(k)}
\;\Rightarrow\;
\text{escalate to } E_{k+1}
\]
This is the biological instantiation of escalation dominance
\cite{McCarthy2026b}: higher levels engage only when lower levels
fail, achieving $O(1)$ steady-state context cost.
\end{definition}

\begin{definition}[Effective Context Extension]
\label{def:extension}
When $E_k$ is coupled to $E_{k+1}$ via an intact escalation chain,
the \emph{effective bounded context} of $E_k$ is:
\[
C_n^{\mathrm{eff}}(E_k) = C_n^{(k)} + \sum_{j=k+1}^{m}
\lambda_j \cdot C_n^{(j)}
\]
where $\lambda_j \in [0, 1]$ is the \emph{aggregate} coupling strength
between levels, summarizing the fidelity and bandwidth of intercellular
signaling across all channels at level $j$
(Section~\ref{sec:coupling-biology} decomposes $\lambda_j$ into
per-channel components). $\lambda_j = 1$ represents perfect coupling
(all of $E_j$'s context is available to $E_k$ via signaling);
$\lambda_j = 0$ represents complete decoupling.

The key property: $C_n^{\mathrm{eff}} > C_n^{(k)}$. The cell
``borrows'' context from higher organizational levels. A liver cell
does not need to independently represent the organism's metabolic
state; it receives that information via hormonal and neural
signaling from the organ and organism levels.
\end{definition}

\begin{definition}[Chain Severance]
\label{def:severance}
\emph{Chain severance at level $k$} occurs when the coupling
$\lambda_{k+1} \to 0$: the entity $E_k$ loses its connection to all
higher organizational levels. Post-severance:
\[
C_n^{\mathrm{eff}}(E_k) = C_n^{(k)}
\]
The cell's effective context collapses to its intrinsic cell-level
bound.
\end{definition}

\subsection{Biological Instantiation of the Coupling Coefficients}
\label{sec:coupling-biology}

The abstract coupling coefficients $\lambda_j$ correspond to
specific, measurable intercellular communication mechanisms at each
level of the hierarchy.

\textbf{Level $0 \to 1$ (cell $\to$ tissue coupling).}
This is the primary coupling layer and the level at which most
cancers sever. The coupling mechanisms are:
\begin{itemize}
\item \textbf{Gap junctions} (connexins, especially Cx43): direct
   cytoplasmic channels between adjacent cells that share ions,
   metabolites, and second messengers. Gap junctional intercellular
   communication (GJIC) is the highest-bandwidth cell-tissue coupling
   channel. Connexin-43 downregulation is among the earliest events
   in many epithelial cancers \cite{Aasen2016}.
\item \textbf{E-cadherin} (CDH1): homophilic adhesion molecule that
   mediates contact inhibition of proliferation. Loss of E-cadherin
   is the defining event of epithelial-mesenchymal transition (EMT)
   and directly severs the contact-dependent component of $\lambda_1$
   \cite{Hanahan2011}.
\item \textbf{Juxtacrine signaling} (Notch--Delta): direct
   cell-cell contact signaling that maintains tissue identity and
   differentiation state. Requires physical adjacency, destroyed
   by loss of tissue architecture.
\item \textbf{Integrin--ECM attachment}: anchorage to the
   extracellular matrix and basement membrane. Loss of anchorage
   dependence (anoikis resistance) severs the structural component
   of tissue coupling.
\end{itemize}

\textbf{Level $1 \to 2$ (tissue $\to$ organ coupling).}
\begin{itemize}
\item \textbf{Paracrine signaling}: local growth factors and
   morphogen gradients (Wnt, Hedgehog, BMP, TGF-$\beta$) that
   coordinate tissue-level organization within the organ context.
\item \textbf{Extracellular matrix architecture}: the structural
   scaffold that encodes tissue-level spatial organization and
   transmits mechanical signals.
\item \textbf{Local vasculature}: nutrient and oxygen delivery
   that ties tissue metabolism to organ-level regulation. Vascular
   normalization restores this coupling channel.
\end{itemize}

\textbf{Level $2 \to 3$ (organ $\to$ organism coupling).}
\begin{itemize}
\item \textbf{Endocrine signaling}: systemic hormones (insulin,
   estrogen, cortisol, thyroid hormone) that couple organ function
   to organismal state. Hormone-dependent cancers (breast, prostate)
   retain partial $\lambda_3$ coupling, explaining their relative
   treatability.
\item \textbf{Immune surveillance}: T-cells, NK cells, and
   macrophages that enforce organism-level quality control on
   aberrant cells. Immune checkpoint molecules (PD-L1, CTLA-4) are
   the molecular implementation of this coupling channel.
\item \textbf{Neural innervation}: autonomic nervous system
   regulation of organ function and tissue homeostasis.
\item \textbf{Circadian regulation}: organism-level temporal
   coordination via the SCN-driven hormonal cascade.
\end{itemize}

More precisely, each aggregate $\lambda_j$ decomposes into per-channel
components. Chain severance can be partial: a cell may lose
gap junctional coupling ($\lambda_1^{\mathrm{GJIC}} \to 0$) while
retaining cadherin-mediated contact inhibition
($\lambda_1^{\mathrm{CDH1}} > 0$). This partial severance produces
the intermediate phenotypes observed in pre-malignant lesions.
Complete severance across all channels at a given level produces the
full malignant phenotype.

% ------------------------------------------------------------------
\section{Main Result}
\label{sec:main}
% ------------------------------------------------------------------

\begin{proposition}[Escalation Chain Severance Produces Convergent
Malignancy]
\label{thm:cancer}
Let $E_0$ be a cell in a multicellular organism with intact
escalation chain $E_0 \to E_1 \to \cdots \to E_m$, modeled as an
entity under the bounded-context framework. If chain severance
occurs at level $k = 0$ (cell loses coupling to tissue), then:

\textbf{Part I.} $E_0$ optimizes under $C_n^{(0)}$ alone.
By graph necessity \cite{McCarthy2026a}, it maintains only the
propositions serviceable within cell-level bounded context.

\textbf{Part II.} By $K/A$ inseparability \cite{McCarthy2026c}, the
knowledge base $K_0$ contracts to index only cell-level actions
$A_0$: those executable without tissue-level coordination.

\textbf{Part III.} By the escalation convergence result \cite{McCarthy2026b}, with
no higher level to escalate to, all optimization pressure concentrates
at $E_0$. The cell becomes the terminal level of its own hierarchy.

\textbf{Part IV.} The action space under cell-level $C_n^{(0)}$
converges across cell types. $C_n^{(0)}$ is small relative to the
full repertoire of cell-level actions, so the asymmetric retention
criterion of \cite{McCarthy2026c} forces displacement toward the
same survival-essential set: with limited slots and the same
survival-critical actions competing for them, tissue-specific content
is progressively displaced by actions with nonzero expected benefit
under cell-level optimization.
\end{proposition}

\textbf{Argument.}

\textbf{Part I: Context Collapse.}

Before severance, $E_0$ operates with effective context
$C_n^{\mathrm{eff}} = C_n^{(0)} + \sum_{j=1}^{m} \lambda_j
C_n^{(j)}$. The cell maintains propositions that reference
tissue-level and organism-level states: differentiation signals,
growth factor dependencies, contact inhibition cues, apoptotic
triggers from the tissue microenvironment. These propositions are
serviceable \emph{because} the escalation chain provides the context
in which they have grounding.

After severance ($\lambda_j = 0$ for $j \geq 1$), these
propositions lose their grounding: they reference states that $E_0$
can no longer observe (tissue integrity, organ function, organismal
need). By graph necessity \cite{McCarthy2026a}, propositions without
grounding in observable states cannot be maintained under bounded
context without displacing grounded propositions. The retention
criterion of \cite{McCarthy2026c} -- retain when benefit is
uncertain, prune when benefit is certainly zero -- applies: the
benefit of tissue-referencing propositions is certainly zero to a
cell that cannot observe tissue state. They are pruned.

$K_0$ contracts to propositions grounded in cell-level observables:
local nutrient concentration, immediate mechanical environment,
intracellular state.

\textbf{Part II: Action Space Contraction.}

By $K/A$ inseparability \cite{McCarthy2026c}, $K_0$ indexes $A_0$
and vice versa. As $K_0$ contracts to cell-level propositions, $A_0$
contracts to cell-level actions. Actions that require tissue-level
coordination -- responding to contact inhibition, executing
apoptosis on organismal command, maintaining differentiated quiescence
-- lose their knowledge support. They are not ``disabled'' by
mutation; they become \emph{inaccessible} because the knowledge
required to execute them (tissue state, organismal need) is no longer
in $K_0$.

What remains in $A_0$ are actions executable with cell-level knowledge
alone:
\begin{itemize}
\item \textbf{Proliferation}: cell division requires only local
   nutrient availability and intracellular state; no tissue-level
   coordination needed.
\item \textbf{Metabolic autonomy}: aerobic glycolysis (the Warburg
   effect \cite{Warburg1956}) is the metabolic mode with the fewest
   tissue-level \emph{dependencies}: it does not require reliable
   tissue-level oxygen delivery infrastructure. Critically, cancer
   cells use glycolysis even when oxygen is present---not because
   oxygen is unavailable, but because after severance the cell lacks
   tissue-level information about oxygen supply reliability and
   defaults to the metabolic strategy that requires no such
   information. This does not imply mitochondrial shutdown (many
   cancer cells retain functional oxidative phosphorylation and use
   both pathways), but glycolytic capacity provides metabolic
   independence from tissue-level coordination.
\item \textbf{Motility}: movement in response to local nutrient
   gradients requires only cell-level sensing.
\item \textbf{Immune evasion}: both passive and active mechanisms.
   MHC-I surface markers, maintained by tissue-level signaling, decay
   when that signaling is lost. Additionally, cancer cells actively
   upregulate PD-L1 in response to immune attack (IFN-$\gamma$
   signaling)---an adaptive response that requires only cell-level
   sensing of immune pressure, not tissue-level coordination.
\end{itemize}

\textbf{Part III: Terminal Optimization.}

In the intact organism, the cell's optimization is bounded by
escalation: problems exceeding $C_n^{(0)}$ escalate to tissue, organ,
or organism level. The cell does not need to solve every problem;
it only needs to handle its local scope and signal failure upward.

After severance, there is no higher level. By the escalation
convergence result \cite{McCarthy2026b}, this means all optimization
pressure concentrates at $E_0$. The cell is no longer a component
in a hierarchy; it is a complete autonomous entity. Its optimization
target is its own persistence, gradient descent on $U$ at cell
level \cite{McCarthy2026d}, with no constraint from above.

This is precisely the optimization regime of a unicellular organism:
an autonomous entity with cell-level $C_n$, optimizing for its own
survival under bounded context. The cell has not ``reverted'' to a
unicellular program. It has been \emph{forced into the same
optimization regime} by the loss of its escalation chain.

\textbf{Part IV: Convergence.}

The convergence argument is not that all severed cells must have
identical knowledge graphs. Graph necessity \cite{McCarthy2026a}
requires graph \emph{structure}, not identical graph \emph{content}.
The convergence is at the level of the action space, and it follows
from the interaction of bounded context with the asymmetric retention
criterion.

$C_n^{(0)}$ is small relative to the full repertoire of possible
cell-level actions. Under the displacement criterion of
\cite{McCarthy2026c}, when $K_0$ is at capacity, a proposition $p$
is displaced by $q$ only when $\mathbb{E}[G(a_q, K)] >
\mathbb{E}[G(a_p, K)]$, strictly higher marginal survival value.
After severance, tissue-specific propositions have zero marginal
value (they reference unobservable states), while cell-level
survival actions (proliferation, nutrient acquisition, damage repair)
have nonzero value regardless of tissue of origin. The retention
criterion therefore drives displacement in the same direction for
all severed cells: tissue-specific content is replaced by cell-level
survival content.

The convergence is not perfect; residual tissue-specific $K_0$
content produces tissue-specific variation, especially early in
progression. But the attractor is the same: the set of actions with
highest marginal survival value under $C_n^{(0)}$. This set is
constrained by cell biology, not by tissue of origin, so it
converges across cancer types.

This offers a structural account for why Hanahan and Weinberg's
hallmarks \cite{Hanahan2011} are observed across cancer types: they
are not shared mutations or shared ancient programs, but the
\emph{convergent action space} that the retention criterion selects
under cell-level bounded context.
\medskip

\textbf{Status of this result.} Proposition~\ref{thm:cancer} is a
conceptual argument, not a formal derivation. Each Part invokes a
result from the formal chain (Papers~1--4) and applies it to the
biological setting by analogy. The analogies are specific and
testable---Section~\ref{sec:predictions} states falsifiable
predictions---but the Proposition does not prove that the hallmarks
are the \emph{unique} convergent action set under cell-level $C_n$.
It argues that convergence to a small, survival-essential set is
structurally predicted, and that the observed hallmarks are consistent
with that prediction.

% ------------------------------------------------------------------
\section{Corollaries}
\label{sec:corollaries}
% ------------------------------------------------------------------

\begin{corollary}[Velocity of Malignant Transformation]
\label{cor:velocity}
Malignant transformation does not require the accumulation of a
specific set of mutations. It requires chain severance: the
loss of coupling to the tissue-level escalation chain. Chain
severance can occur through a single event (loss of a key signaling
receptor, disruption of gap junctions, escape from the tissue
microenvironment) rather than through the sequential accumulation
of independent mutations.
\end{corollary}

\textbf{Argument.}
The somatic mutation theory requires multiple independent hits
\cite{Vogelstein2013}: each hallmark capability must be separately
acquired through mutation. Under our framework, the hallmarks are
not independently acquired; they are jointly \emph{released}
when the escalation chain severs. A single event that decouples the
cell from tissue-level signaling simultaneously enables all
hallmark behaviors, because all of them are cell-level actions that
were \emph{suppressed} by tissue-level coordination, not
\emph{absent} from the cell's intrinsic repertoire.

This explains why cancer can progress rapidly from a single
initiating event: the ``multi-hit'' appearance is a consequence of
progressive chain severance (each hit weakens coupling further)
rather than independent capability acquisition. The velocity is
structural, not stochastic.

\textbf{Compatibility with multi-hit observations.} The framework
does not contradict the observation that multiple mutations are
typically found in cancers. Rather, it reinterprets their role:
mutations are mechanisms of chain severance (disrupting signaling
pathways, disabling apoptotic responses to tissue-level commands),
not independent capability acquisitions. The ``multi-hit'' pattern
reflects progressive decoupling: each mutation weakens a different
coupling channel $\lambda_j$, rather than independent construction
of each hallmark.
\medskip

\begin{corollary}[Therapeutic Resistance]
\label{cor:resistance}
Established tumors exhibit persistent therapeutic resistance because
the severed cell is \emph{locally rational}: it is optimizing
correctly under its own bounded context.
\end{corollary}

\textbf{Argument.}
Therapies that target specific mutations (targeted therapy) or
specific proliferative mechanisms (chemotherapy) impose selection
pressure on an entity that is already performing gradient descent on
$U$ at cell level \cite{McCarthy2026d}. The cell's $\gamma > 0$
(Paper~4, \cite{McCarthy2026d}) guarantees continued engagement with
adaptive mechanisms. Its learning efficiency $\eta_M > 0$
(the rate of improvement per unit of learning signal;
\cite{McCarthy2026d}) guarantees learning from
therapeutic pressure. The therapy becomes a new environmental signal
that the cell incorporates into its optimization.

This explains why resistance is the norm rather than the exception:
the cell is not a broken machine to be fixed but an autonomous
entity optimizing under its own constraints. Therapy that does not
restore the escalation chain merely changes the optimization
landscape without changing the optimization regime.

Effective therapy, on this account, must either:
\begin{enumerate}
\item \textbf{Restore the chain}: re-establish tissue-level coupling
   so that $C_n^{\mathrm{eff}} > C_n^{(0)}$, returning tissue-level
   actions to the cell's accessible action space. Differentiation
   therapy (e.g., all-trans retinoic acid in acute promyelocytic
   leukemia) operates this way.
\item \textbf{Exploit the constraint}: target cell-level actions
   that are \emph{necessary} under $C_n^{(0)}$, not just contingently
   adopted. If an action is necessary under cell-level bounded
   context, the cell cannot evolve away from it without leaving the
   $C_n^{(0)}$ optimization regime.
\end{enumerate}
\medskip

\begin{corollary}[The Atavistic Pattern Without Atavistic Causation]
\label{cor:atavistic}
The phylostratigraphic signal -- cancer genes being
disproportionately ancient \cite{Trigos2017} -- is real but does
not imply atavistic reversion. It reflects convergent necessity.
\end{corollary}

\textbf{Argument.}
Genes that are ancient are precisely the genes that were optimized
for cell-level bounded context during the unicellular era. When a
modern cell undergoes chain severance and is forced back to
cell-level optimization, the \emph{same genes} become relevant
again, not because an ancient program is ``reactivated'' but
because the same constraint ($C_n^{(0)}$) selects for the same
solutions.

The atavistic theory \cite{Davies2011} correctly identifies the
phenotypic pattern (cancer cells resemble unicellular organisms) and
the genetic evidence (ancient genes are upregulated). Its error is
causal: it infers historical reversion where the mechanism is
structural convergence. The genes are not activated \emph{because}
they are ancient; they are activated because they encode the actions
that are necessary under cell-level bounded context, and they happen
to be ancient because unicellular organisms faced the same constraint.

This distinction is testable: the atavistic theory predicts that the
ancient gene set forms a coherent ``unicellular program'' that is
activated as a unit. Our framework predicts that the set of genes
activated in any given cancer will be determined by the specific
$C_n^{(0)}$-level optimization landscape, and will include ancient
genes \emph{to the extent that} they encode cell-level actions,
but will also include recent genes that happen to serve cell-level
optimization (e.g., genes co-opted for immune evasion that evolved
in the multicellular era).
\medskip

\begin{corollary}[Metastasis as Optimal Foraging Under Cell-Level
$C_n$]
\label{cor:metastasis}
Metastasis is not an acquired ``capability'' but the default
exploratory behavior of an autonomous entity optimizing under
bounded context in a spatially heterogeneous environment.
\end{corollary}

\textbf{Argument.}
In an intact organism, cell motility is constrained by tissue-level
signaling: basement membrane attachment, contact inhibition, tissue
boundary maintenance. These constraints are maintained by the
escalation chain. The tissue level enforces spatial organization
that individual cells could not compute independently.

After chain severance, the cell has no spatial constraint beyond its
local environment. By the non-termination result of
\cite{McCarthy2026d} (Theorem~1, Part~II), the reachable trajectory
space $T_{\mathrm{reachable}}(K_0)$ grows monotonically with $K_0$.
A cell in a nutrient-poor local environment will, through gradient
descent on $U$, discover that movement to nutrient-rich regions
reduces uncertainty about survival.

\textbf{Inefficiency is predicted, not anomalous.} Metastasis is
exploration under $C_n^{(0)}$---cell-level bounded context with no
tissue-level guidance. More than 99.9\% of circulating tumor cells
die without colonizing a distant site \cite{Massague2016}. This
massive inefficiency is exactly what the framework predicts:
exploration without higher-level coordination is high-variance.
The cell has $\gamma > 0$ (it engages adaptive mechanisms) but
its context window is too small to model the circulatory environment,
immune surveillance, or distant-site compatibility. Each step is
locally rational---movement reduces local uncertainty---but the
cell lacks the $C_n^{\mathrm{eff}}$ to plan a viable multi-step
trajectory. The rare successful metastases are not evidence of
acquired capability but survivorship bias over massive parallel
exploration.

The ``invasion-metastasis cascade'' of epithelial-mesenchymal
transition, intravasation, survival in circulation, and colonization
at distant sites is not a sequence of independently acquired
capabilities. It is the progressive unfolding of cell-level
exploration in a spatially structured environment, where each step
reveals new states that reduce local uncertainty.
\medskip

\begin{corollary}[Transplant Rejection as Chain Mismatch]
\label{cor:transplant}
Organ transplant rejection is not chain severance but \emph{chain
mismatch}: the transplanted tissue has intact internal coupling
(levels $0 \to 1$) but mismatched organism-level coupling
(level $2 \to 3$). The immune system's rejection response is the
escalation chain functioning correctly, detecting cells whose
coupling markers do not match the recipient organism.
\end{corollary}

\textbf{Argument.}
A transplanted organ retains its internal tissue organization: the
cells are coupled to each other and to tissue-level architecture
via gap junctions, cadherins, and paracrine signaling. The
$\lambda_1$ coupling channels are intact. What is mismatched is
$\lambda_3^{\mathrm{immune}}$: the organism-level immune
surveillance system recognizes the transplanted cells' MHC markers
as foreign, not coupled to \emph{this} organism's escalation
chain.

At the level of the coupling variable $\lambda_3^{\mathrm{immune}}$,
chain severance (cancer) and chain mismatch (transplant) produce the
same signal: absent or aberrant organism-level coupling. The immune
mechanisms differ in detail---transplant rejection is driven by MHC
mismatch recognition, while cancer immune evasion involves PD-L1
upregulation and MHC-I downregulation---but structurally, both
represent failures of the level $2 \to 3$ coupling, and the
organism-level enforcement mechanism responds to both.

This reframes the immunosuppression tradeoff structurally.
Immunosuppressive therapy (cyclosporine, tacrolimus) weakens the
organism-level coupling enforcement
$\lambda_3^{\mathrm{immune}} \to 0$ globally. This permits the
mismatched organ but simultaneously disables the surveillance
mechanism that detects genuine chain severance. The well-documented
2--5$\times$ elevation in cancer incidence among transplant
recipients \cite{Engels2011} is the predicted consequence:
weakened chain enforcement permits both the intended mismatch
\emph{and} unintended severance events.

\textbf{Tolerance induction} -- protocols that establish mixed
hematopoietic chimerism or regulatory T-cell-mediated acceptance
-- operates differently. Rather than weakening enforcement, it
re-establishes matching coupling: the recipient's immune system
learns to recognize the transplanted tissue as part of its own
escalation chain. The coupling
$\lambda_3^{\mathrm{immune}}$ is restored for the transplant
specifically, while global enforcement remains intact. The
framework predicts that tolerance induction should produce better
long-term outcomes than chronic immunosuppression, not only
for graft survival but for overall survival, because it
preserves the organism-level chain enforcement that prevents
cancer. Early clinical evidence from mixed chimerism protocols
is consistent with this prediction \cite{Kawai2008}.

HLA matching reduces the mismatch at the coupling interface itself,
minimizing the signal that triggers enforcement without weakening
enforcement. This is why HLA-matched transplants require less
immunosuppression and have better outcomes: less mismatch means
less enforcement to suppress.
\medskip

% ------------------------------------------------------------------
\section{Mechanisms of Chain Severance}
\label{sec:mechanisms}
% ------------------------------------------------------------------

Chain severance is not a single event but a spectrum of decoupling
mechanisms, each reducing $\lambda_j$ toward zero:

\textbf{Genetic mechanisms.} Mutations in signaling receptors (e.g.,
E-cadherin loss), tumor suppressor pathways (e.g., p53, Rb), or
growth factor dependencies reduce the cell's capacity to receive
tissue-level signals. These are the mechanisms identified by SMT
\cite{Vogelstein2013}. In our framework, they are not causes of
cancer but \emph{mechanisms of chain severance}: each mutation
weakens a specific coupling channel.

\textbf{Microenvironmental mechanisms.} Chronic inflammation,
fibrosis, or tissue damage disrupts the signaling infrastructure
through which higher levels communicate with cells. This is the
mechanism identified by TOFT \cite{Sonnenschein2000}: the tissue
architecture that maintains coupling is destroyed, and cells
decouple even without intrinsic mutations. The coupling
$\lambda_j \to 0$ occurs not because the cell's receivers are
broken but because the signal is no longer transmitted.

\textbf{Epigenetic mechanisms.} Methylation changes, chromatin
remodeling, or transcriptional reprogramming can silence the
genes encoding tissue-level signaling components without genomic
mutation. This produces chain severance through altered gene
expression rather than altered sequence.

\textbf{Spatial mechanisms.} Physical separation from the tissue
microenvironment -- through basement membrane breach, for
instance -- removes the cell from the signaling field. The cell's
coupling hardware is intact; the coupling medium is absent.

All four mechanisms produce the same outcome: $\lambda_j \to 0$,
effective context collapse to $C_n^{(0)}$, and optimization under
cell-level bounded context. The convergent phenotype (the hallmarks)
follows regardless of \emph{which} mechanism severed the chain.
This explains why cancers arising from different mechanisms
(hereditary, environmental, sporadic) converge on the same
phenotypic endpoints.

% ------------------------------------------------------------------
\section{The Encoding Hierarchy in Cancer}
\label{sec:encoding}
% ------------------------------------------------------------------

The encoding hierarchy of \cite{McCarthy2026d} has a specific
consequence for cancer that connects to the empirical phenomenon of
tumor heterogeneity.

In the intact organism, actions at each encoding level have been
validated to the threshold $\theta_i = 1 - \varepsilon / C_i$. The
genomic level ($L_0$) encodes actions validated across evolutionary
timescales. The epigenetic level ($L_1$) encodes tissue-specific
programs validated across developmental timescales. The active
context level ($L_n$) encodes moment-to-moment responses.

After chain severance, the encoding hierarchy is disrupted at all
levels above the cell:

\begin{itemize}
\item \textbf{Organism-level encodings} (hormonal responses,
   circadian regulation, immune cooperation) lose their validation
   context. They were validated under $C_n^{\mathrm{eff}}$; they are
   now evaluated under $C_n^{(0)}$. Many fail the retention criterion.

\item \textbf{Tissue-level encodings} (differentiation state,
   contact inhibition, growth factor dependence) similarly lose
   validation context.

\item \textbf{Cell-level encodings} (basic metabolism, division
   machinery, DNA repair) retain their validation. They were
   validated under $C_n^{(0)}$ originally and remain valid.
\end{itemize}

The result is \emph{selective devalidation}: the encoding hierarchy
does not collapse uniformly but selectively loses its higher-level
content. What remains is a cell operating on cell-level encodings,
the same encodings that a unicellular organism would maintain. This
is the encoding-level explanation for the atavistic gene expression
pattern observed by Trigos et al.\ \cite{Trigos2017}: ancient genes
are preferentially retained not because they are ``reactivated'' but
because they are the ones whose validation survives context collapse.

\textbf{Tumor heterogeneity.} The progressive nature of context
collapse explains intratumor heterogeneity. Cells at different
stages of chain severance retain different amounts of higher-level
encoding. Early-stage tumor cells may retain partial tissue-level
coupling (some $\lambda_j > 0$), producing a mixed phenotype. As
severance progresses, the phenotype converges toward the cell-level
optimum. This gradient of severance within a single tumor produces
the phenotypic heterogeneity observed empirically, without requiring
different clonal origins for different phenotypes.

Clonal evolution \cite{Nowell1976} describes the \emph{dynamics}
of this process: cells with more complete severance have higher
fitness in the decoupled regime and are selected. But the direction
of evolution is not random; it converges on the cell-level
optimum, which is structurally determined by $C_n^{(0)}$.

% ------------------------------------------------------------------
\section{Engagement with Prior Work}
\label{sec:prior}
% ------------------------------------------------------------------

\textbf{Hanahan and Weinberg \cite{Hanahan2000, Hanahan2011}:
Hallmarks of Cancer.}
The hallmarks -- sustaining proliferative signaling, evading growth
suppressors, resisting cell death, enabling replicative immortality,
inducing angiogenesis, activating invasion and metastasis, plus the
enabling characteristics and emerging hallmarks of the 2011 update
-- are described as ``capabilities'' that must be ``acquired.'' Our
framework reinterprets them: they are not acquired but
\emph{revealed}. They are the necessary action space of any entity
optimizing under cell-level bounded context. The hallmarks are
universal across cancer types because $C_n^{(0)}$ is universal
across cell types. What differs across cancers is the \emph{route}
of chain severance (which coupling channels fail first), not the
\emph{destination} (cell-level optimization).

\textbf{Davies and Lineweaver \cite{Davies2011}: Atavistic Theory.}
See Corollary~\ref{cor:atavistic}. The atavistic theory correctly
identifies the phenotypic and phylostratigraphic pattern. Our
framework provides the mechanism it lacks (chain severance) and
corrects the causal inference (convergent necessity, not historical
reversion). The two theories make different predictions about which
genes will be activated in any given cancer: the atavistic theory
predicts a coherent ancient program; our framework predicts
gene activation determined by $C_n^{(0)}$-level optimization,
which overlaps heavily but not completely with the ancient gene set.

\textbf{Sonnenschein and Soto \cite{Sonnenschein2000}: TOFT.}
TOFT is the closest existing framework to our account. It correctly
identifies the organizational level (tissue, not cell) at which
cancer originates, and correctly argues that proliferation is the
default state of cells, with tissue organization as the restraint.
Our framework provides TOFT with the formalism it has lacked: the
escalation chain is the formal structure of tissue organization,
$C_n^{\mathrm{eff}}$ is the formal measure of organizational
integration, and chain severance is the formal description of
organizational breakdown. TOFT's ``default proliferative state'' is
our ``necessary action space under cell-level $C_n$.''

\textbf{Nowell \cite{Nowell1976}: Clonal Evolution.}
Clonal evolution describes the dynamics of cancer progression as
Darwinian selection among cell populations. Our framework does not
replace this account but provides its \emph{attractor}: clonal
evolution does not wander randomly through phenotype space but
converges on the cell-level optimum determined by $C_n^{(0)}$. The
direction of clonal evolution is structurally determined; the
stochastic component is in \emph{which cells} reach the attractor
first, not \emph{where} the attractor is.

\textbf{Huang et al.\ \cite{Huang2009}: Attractor Landscape.}
Huang's gene-regulatory-network framework models cell states as
attractors in a high-dimensional expression landscape, with cancer
as occupation of abnormal attractors that normal cells rarely visit.
Our framework subsumes this picture: the attractors that Huang
identifies are the \emph{stable configurations} of cell-level
optimization under $C_n^{(0)}$. Chain severance does not create new
attractors; it removes the tissue-level coupling that steered cells
away from the cell-level optimum. What Huang describes as an
``abnormal attractor'' is, on our account, the \emph{default}
attractor under cell-level bounded context---abnormal only relative
to tissue-integrated operation. The two frameworks make compatible
predictions, but ours provides a principled reason \emph{why} the
attractor landscape has the structure it does: the bounded-context
constraint $C_n^{(0)}$ determines which configurations are stable
for an autonomous cell.

\textbf{Aktipis et al.\ \cite{Aktipis2015}: Cancer as Cheating.}
The evolutionary cooperation framework describes cancer as a
breakdown of multicellular cooperation: cells that ``cheat'' by
proliferating at the expense of the collective. Our framework
provides the structural basis: cooperation is maintained by the
escalation chain (higher organizational levels enforce cooperative
behavior by extending $C_n^{\mathrm{eff}}$). ``Cheating'' is not a
strategy adopted by the cell but the necessary outcome of chain
severance: a cell that cannot observe tissue-level state cannot
cooperate with the tissue, because cooperation requires the
knowledge base ($K_0$) to contain tissue-referencing propositions
that are no longer grounded.

\textbf{Merlo et al.\ \cite{Merlo2006}: Cancer as Ecological Process.}
The ecological framework describes tumor progression as an
evolutionary process within the ecosystem of the body. Our framework
adds the structural constraint: the ``ecosystem'' is the escalation
chain, and cancer progression is the progressive collapse of that
chain. The ecological metaphor is correct but incomplete; it
describes dynamics without explaining why those dynamics converge.

% ------------------------------------------------------------------
\section{Testable Predictions}
\label{sec:predictions}
% ------------------------------------------------------------------

The framework generates predictions that distinguish it from existing
theories:

\begin{enumerate}
\item \textbf{Coupling measurement predicts malignancy.} Quantifying
   $\lambda_j$ (the coupling strength between a cell and its tissue
   context) should predict malignant potential more accurately than
   mutational burden alone. Cells with low $\lambda_j$ but few
   mutations should still progress to malignancy; cells with high
   mutational burden but intact $\lambda_j$ should not. This can be
   tested by measuring intercellular communication (gap junction
   activity, paracrine signaling responsiveness, contact inhibition)
   alongside mutational profiling.

\item \textbf{Convergent gene expression is constraint-determined,
   not phylogeny-determined.} Cancer gene expression should converge
   on a set determined by cell-level optimization, which \emph{mostly
   but not entirely} overlaps with phylogenetically ancient genes.
   In particular, recently evolved genes that serve cell-level
   autonomous function (e.g., certain immune evasion mechanisms that
   evolved in the multicellular era) should be upregulated in cancer
   despite being phylogenetically recent. The atavistic theory
   predicts they should not be.

\item \textbf{Differentiation therapy should outperform cytotoxic
   therapy.} Therapies that restore tissue-level coupling
   (differentiation therapy, microenvironment normalization) should
   produce more durable responses than therapies that kill cells
   without restoring coupling, because the former addresses the
   structural cause (chain severance) while the latter imposes
   selection pressure on a locally rational optimizer.

\item \textbf{Tumor heterogeneity should correlate with severance
   gradient.} Within a single tumor, phenotypic heterogeneity should
   correlate with measured coupling strength: cells with residual
   tissue coupling ($\lambda_j > 0$) should exhibit more
   differentiated phenotypes than fully decoupled cells. This is
   testable by spatial transcriptomics combined with intercellular
   communication assays.

\item \textbf{Necessary hallmarks vs.\ contingent hallmarks.}
   The framework predicts that some hallmarks are \emph{necessary}
   under cell-level $C_n$ (proliferation, metabolic autonomy) while
   others are \emph{contingent} on the specific optimization
   landscape (angiogenesis induction, specific immune evasion
   strategies). Necessary hallmarks should appear in every cancer;
   contingent hallmarks should vary by context. This is testable by
   systematic cross-cancer hallmark prevalence analysis.
\end{enumerate}

% ------------------------------------------------------------------
\section{Classification of Therapies by Chain Restoration}
\label{sec:therapy}
% ------------------------------------------------------------------

The framework yields a natural classification of cancer therapies:
those that attempt to \emph{restore} the escalation chain (addressing
the structural cause) versus those that impose \emph{selection
pressure} on a decoupled optimizer (treating the symptom). This
classification predicts durability of response.

\subsection{Chain-Restoring Therapies}

These therapies target the coupling coefficients $\lambda_j$,
attempting to re-establish the cell's connection to higher
organizational levels.

\textbf{Differentiation therapy (Level $0 \to 1$ restoration).}
All-trans retinoic acid (ATRA) in acute promyelocytic leukemia (APL)
is the paradigm case: it restores the cell's tissue-level identity,
re-engaging the differentiation program that chain severance
disabled. ATRA does not kill the cancer cells; it restores their
coupling to the tissue-level escalation chain, returning tissue-level
actions (differentiation, growth arrest) to the accessible action
space $A_0$. The result is a $>90\%$ cure rate, the highest of any
cancer therapy. On our account, this is not a coincidence but a
\emph{prediction}: therapies that restore the chain should
outperform therapies that do not.

\textbf{Immune checkpoint inhibition (Level $2 \to 3$ restoration).}
PD-1/PD-L1 and CTLA-4 inhibitors restore the organism-level
surveillance coupling $\lambda_3^{\mathrm{immune}}$. Cancer cells
that have downregulated immune recognition markers (severing the
organism-level coupling channel) are re-exposed to immune
surveillance. The therapy does not kill cancer cells directly; it
restores the organizational relationship between the immune system
and the tumor. The durable responses observed in a subset of patients
(complete responses lasting years) are consistent with chain
restoration: once organism-level surveillance is re-established, it
is self-maintaining.

\textbf{Microenvironment normalization (Level $0 \to 1$ and $1 \to 2$
restoration).}
Strategies that restore the tissue microenvironment -- ECM
remodeling, stromal normalization, basement membrane repair --
address the tissue-level coupling infrastructure. Low-dose
anti-angiogenic therapy (vascular normalization rather than vascular
destruction) restores the tissue-organ coupling channel
$\lambda_2^{\mathrm{vascular}}$ by normalizing tumor vasculature,
improving drug delivery and re-establishing metabolic coupling to
the organ level.

\textbf{Gap junction restoration (Level $0 \to 1$ restoration).}
Experimental approaches to restore connexin expression (particularly
Cx43) directly target the highest-bandwidth cell-tissue coupling
channel \cite{Aasen2016}. The framework predicts that restoring GJIC
should suppress malignant behavior even without addressing mutations,
because the coupling restoration returns tissue-level actions to the
accessible action space.

\subsection{Non-Chain-Restoring Therapies}

These therapies impose selection pressure on a cell that is already
optimizing under cell-level bounded context. The cell's $\gamma > 0$
and $\eta_M > 0$ \cite{McCarthy2026d} guarantee that it will adapt.

\textbf{Cytotoxic chemotherapy.} Kills dividing cells without
restoring any coupling channel. The surviving subpopulation has
undergone selection for resistance -- slow-cycling cells, drug
efflux pumps, DNA repair enhancement -- all of which are within the
cell-level action space $A_0$. Resistance is not a failure of the
drug; it is the predictable outcome of imposing selection pressure on
a locally rational optimizer.

\textbf{Radiation therapy.} Induces DNA damage. DNA repair is a
cell-level action already in $A_0$; it requires no tissue-level
coupling. The cell's gradient descent on $U$ will optimize repair
pathways in response to radiation pressure.

\textbf{Most targeted therapies.} Block a specific signaling pathway
(e.g., EGFR inhibitors, BRAF inhibitors). The cell, performing
gradient descent on $U$, discovers alternative pathways that restore
the blocked function. Pathway redundancy is a feature of cell-level
action spaces that have been optimized over billions of years of
unicellular evolution. Single-pathway targeting is structurally
insufficient because the cell-level action space contains multiple
routes to the same optimization target.

\textbf{Surgery.} Removes the decoupled entity entirely. This is
effective for localized disease but does not prevent new severance
events in remaining tissue, and cannot address disseminated disease
(metastasis).

\subsection{The Hybrid Strategy}

The framework predicts that the optimal therapeutic strategy combines
chain restoration with exploitation of structural constraints:
\begin{enumerate}
\item Restore coupling where possible (differentiation therapy,
   immunotherapy, microenvironment normalization).
\item For cells where coupling cannot be restored, target actions
   that are \emph{necessary} under cell-level $C_n^{(0)}$, actions
   the cell cannot optimize away from because they are structurally
   required by the bounded-context constraint. The Warburg effect
   (aerobic glycolysis) is a candidate: if glycolytic metabolism is
   \emph{necessary} under cell-level bounded context (because
   oxidative phosphorylation requires tissue-level oxygen delivery
   infrastructure), then targeting glycolysis imposes a constraint
   the cell cannot evolve around.
\item Use non-chain-restoring therapies (chemotherapy, radiation)
   only as cytoreductive measures to reduce tumor burden before
   chain-restoring interventions, not as standalone strategies.
\end{enumerate}

\textbf{Clinical evidence.} The relative success of combination
immunotherapy (chain restoration at level $2 \to 3$) over
single-agent chemotherapy (non-restoring) in melanoma, non-small cell
lung cancer, and renal cell carcinoma is consistent with this
prediction. The curative success of ATRA in APL (chain restoration at
level $0 \to 1$) versus the near-universal resistance to
chemotherapy-only regimens in other acute leukemias provides the
strongest existing evidence.

% ------------------------------------------------------------------
\section{Implications}
\label{sec:implications}
% ------------------------------------------------------------------

\textbf{Cancer is not a disease of cells.} It is a disease of
organizational coupling. The cell itself is functioning correctly;
it is optimizing under its bounded context, exactly as the theorems
predict any entity must. The pathology is in the severed escalation
chain, not in the cell.

\textbf{The ``war on cancer'' metaphor is structurally wrong.} Cancer
cells are not invaders to be destroyed; they are former cooperators
who have lost the organizational infrastructure that made cooperation
possible. Killing them (chemotherapy, radiation) treats the symptom.
Restoring the organizational infrastructure (differentiation therapy,
microenvironment normalization, coupling restoration) treats the
cause. The distinction is between fighting an entity and repairing a
relationship.

\textbf{Prevention over cure.} If cancer is chain severance, then
prevention is coupling maintenance. Factors that maintain tissue
organization (anti-inflammatory interventions, microenvironment
homeostasis, gap junction maintenance) should reduce cancer risk
more effectively than factors that reduce mutational burden,
though the two are related, since mutations are one mechanism of
severance.

\textbf{Peto's paradox.} Large, long-lived organisms do not develop
cancer at rates proportional to their cell count and lifespan
\cite{Peto1975}, a puzzle under the somatic mutation theory,
which predicts that more cell divisions should produce more
mutations and therefore more cancer. The escalation chain framework
offers a structural explanation: larger organisms have more
elaborate tissue organization, stronger intercellular coupling
mechanisms, and more escalation levels. Their
$C_n^{\mathrm{eff}}$ is higher, meaning more severance is required
before cell-level optimization dominates. The additional
organizational infrastructure acts as redundancy in the coupling
channels: multiple mechanisms must fail before the chain severs
completely. Peto's paradox, on this account, is not paradoxical:
more cells but also more coupling yields a roughly constant
severance rate.

\textbf{Benign tumors and non-progression.} Most dysplastic lesions,
adenomas, and benign tumors never become malignant \cite{Nowell1976}.
The framework predicts this: partial decoupling ($0 < \lambda_j < 1$)
does not produce full convergence to the cell-level action space.
A cell with residual tissue coupling retains tissue-referencing
propositions in $K_0$ and does not undergo complete displacement
toward cell-level survival content. The retention criterion
(Paper~3, \cite{McCarthy2026c}) only drives displacement when
tissue-specific propositions have zero marginal value---which
requires $\lambda_j \approx 0$, not merely reduced. Most initiated
cells remain in a partially coupled state indefinitely; their
$C_n^{\mathrm{eff}}$ is reduced but not collapsed to $C_n^{(0)}$.

\textbf{The threshold for progression.} The clinical observation
\cite{Folkman1971} that many cancers persist at low grade for years
before rapid progression is explained by this severance gradient.
Partial decoupling is stable while $\lambda_j$ remains above a
threshold. Progression occurs when $\lambda_j$ crosses that threshold
and cascading decoupling begins: each loss of coupling removes a
constraint that was maintaining the remaining coupling channels.

% ------------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}
% ------------------------------------------------------------------

Cancer is not random. Cancer is not ancient. Cancer is what happens
when a cell loses its escalation chain.

This paper has applied the bounded-context framework of Papers~1--4
to cancer biology as a speculative but structurally grounded
proposal. The framework offers what the three major cancer theories
have separately lacked: a formal account of \emph{why} organizational
breakdown would produce the specific, convergent phenotype that it
does. SMT identifies the mechanisms of chain severance (mutations).
The atavistic theory identifies the phenotypic pattern (convergence
on unicellular-like behavior). TOFT identifies the organizational
level (tissue, not cell). Our framework unifies them: mutations
sever the escalation chain (SMT), producing a cell that optimizes
under cell-level bounded context (TOFT), which converges on the same
action space that shaped unicellular life (atavistic pattern),
not by reversion but by structural necessity.

The therapeutic predictions align with where clinical oncology is
already heading: differentiation therapy, immunotherapy, and
microenvironment normalization are gaining ground over purely
cytotoxic approaches. The framework provides a structural reason
for this trend: these therapies restore the escalation chain,
while cytotoxic therapies impose selection pressure on a locally
rational optimizer. The framework generates testable predictions
(Section~\ref{sec:predictions}) that distinguish this account from
existing theories.

The framework's contribution is not empirical but structural: it
offers a target for therapy that is not the cell but the coupling,
not the action but the context that determines which actions are
selected. Whether this structural reasoning maps faithfully onto
cancer biology is an empirical question. The predictions are offered
in that spirit.

\section*{Acknowledgments}

The author used Claude (Anthropic) for drafting assistance, literature review, and \LaTeX{} preparation. All theoretical content, proofs, and arguments are the author's own.

\bibliographystyle{plain}
\begin{thebibliography}{99}

\bibitem{Aasen2016}
T.~Aasen, M.~Mesnil, C.~C. Naus, P.~D. Lampe, and D.~W. Laird.
\newblock Gap junctions and cancer: Communicating for 50 years.
\newblock \emph{Nature Reviews Cancer}, 16(12):775--788, 2016.

\bibitem{Aktipis2015}
C.~A. Aktipis, A.~M. Boddy, G.~Jansen, U.~Hibner, C.~C. Maley, and
  G.~S. Wilkinson.
\newblock Cancer across the tree of life: Cooperation and cheating in
  multicellularity.
\newblock \emph{Philosophical Transactions of the Royal Society B},
  370(1673):20140219, 2015.

\bibitem{Folkman1971}
J.~Folkman, E.~Merler, C.~Abernathy, and G.~Williams.
\newblock Isolation of a tumor factor responsible for angiogenesis.
\newblock \emph{The Journal of Experimental Medicine}, 133(2):275--288, 1971.

\bibitem{Engels2011}
E.~A. Engels, R.~M. Pfeiffer, J.~F. Fraumeni, B.~L. Kasiske,
  A.~K. Israni, J.~J. Snyder, R.~A. Wolfe, N.~P. Goodrich,
  A.~I. Bayakly, G.~L. Clarke, and others.
\newblock Spectrum of cancer risk among {US} solid organ transplant
  recipients.
\newblock \emph{JAMA}, 306(17):1891--1901, 2011.

\bibitem{Davies2011}
P.~C.~W. Davies and C.~H. Lineweaver.
\newblock Cancer tumors as {Metazoa} 1.0: Tapping genes of ancient ancestors.
\newblock \emph{Physical Biology}, 8(1):015001, 2011.

\bibitem{Kawai2008}
T.~Kawai, A.~B. Cosimi, T.~R. Spitzer, N.~Tolkoff-Rubin, M.~Suthanthiran,
  S.~L. Saidman, J.~Shaffer, F.~I. Preffer, R.~Ber, G.~Pascual, and
  D.~H. Sachs.
\newblock {HLA}-mismatched renal transplantation without maintenance
  immunosuppression.
\newblock \emph{New England Journal of Medicine}, 358(4):353--361, 2008.

\bibitem{Hanahan2000}
D.~Hanahan and R.~A. Weinberg.
\newblock The hallmarks of cancer.
\newblock \emph{Cell}, 100(1):57--70, 2000.

\bibitem{Hanahan2011}
D.~Hanahan and R.~A. Weinberg.
\newblock Hallmarks of cancer: The next generation.
\newblock \emph{Cell}, 144(5):646--674, 2011.

\bibitem{McCarthy2026a}
P.~D. McCarthy.
\newblock Graph structure is necessary for information preservation under
  bounded context.
\newblock Manuscript, 2026.

\bibitem{McCarthy2026b}
P.~D. McCarthy.
\newblock Convergence of control architectures to escalation under bounded context.
\newblock Manuscript, 2026.

\bibitem{McCarthy2026c}
P.~D. McCarthy.
\newblock Knowledge and action are inseparable under survival pressure.
\newblock Manuscript, 2026.

\bibitem{McCarthy2026d}
P.~D. McCarthy.
\newblock Evaluation-driven descent, encoding permanence, and the structural
  invariants of intelligence.
\newblock Manuscript, 2026.

\bibitem{Merlo2006}
L.~M.~F. Merlo, J.~W. Pepper, B.~J. Reid, and C.~C. Maley.
\newblock Cancer as an evolutionary and ecological process.
\newblock \emph{Nature Reviews Cancer}, 6(12):924--935, 2006.

\bibitem{Peto1975}
R.~Peto, F.~J.~C. Roe, P.~N. Lee, L.~Levy, and J.~Clack.
\newblock Cancer and ageing in mice and men.
\newblock \emph{British Journal of Cancer}, 32(4):411--426, 1975.

\bibitem{Nowell1976}
P.~C. Nowell.
\newblock The clonal evolution of tumor cell populations.
\newblock \emph{Science}, 194(4260):23--28, 1976.

\bibitem{Sonnenschein2000}
C.~Sonnenschein and A.~M. Soto.
\newblock Somatic mutation theory of carcinogenesis: Why it should be dropped
  and replaced.
\newblock \emph{Molecular Carcinogenesis}, 29(4):205--211, 2000.

\bibitem{Trigos2017}
A.~S. Trigos, R.~B. Pearson, A.~T. Papenfuss, and D.~L. Goode.
\newblock Altered interactions between unicellular and multicellular genes
  drive hallmarks of transformation in a diverse range of solid tumors.
\newblock \emph{Proceedings of the National Academy of Sciences},
  114(24):6406--6411, 2017.

\bibitem{Vogelstein2013}
B.~Vogelstein, N.~Papadopoulos, V.~E. Velculescu, S.~Zhou, L.~A. Diaz, and
  K.~W. Kinzler.
\newblock Cancer genome landscapes.
\newblock \emph{Science}, 339(6127):1546--1558, 2013.

\bibitem{Huang2009}
S.~Huang, I.~Ernberg, and S.~Kauffman.
\newblock Cancer attractors: a systems view of tumors from a gene network
  dynamics and developmental perspective.
\newblock \emph{Seminars in Cell \& Developmental Biology}, 20(7):869--876,
  2009.

\bibitem{Massague2016}
J.~Massagu{\'e} and A.~C. Obenauf.
\newblock Metastatic colonization by circulating tumour cells.
\newblock \emph{Nature}, 529(7586):298--306, 2016.

\bibitem{Warburg1956}
O.~Warburg.
\newblock On the origin of cancer cells.
\newblock \emph{Science}, 123(3191):309--314, 1956.

\end{thebibliography}

\end{document}
